---
title: "BVS.DAdj: Bayesian Variable Selection with Dual-Adjacency Network-Informed Ising Priors in R"
author:
  - name: Yubing Yao
    affiliation: Department of Biostatistics and Epidemiology, University of Massachusetts Amherst
    email: yyao@umass.edu
  - name: Mahlet G. Tadesse
    affiliation: Department of Mathematics & Statistics, Georgetown University
    email: mgt26@georgetown.edu
  - name: Raji Balasubramanian
    affiliation: Department of Biostatistics and Epidemiology, University of Massachusetts Amherst School of Public Health
    email: rbalasub@schoolph.umass.edu
abstract: >
  We introduce **BVS.DAdj**, an R package for Bayesian variable selection in
  generalized regression models using Markov random field (MRF) priors that
  incorporate dual adjacency structures among predictors. The package provides
  two MCMC interfaces---`bvs_mh()` and `bvs_pg()`---covering four outcome
  types (binary logistic, continuous Gaussian, right-censored time-to-event
  via Cox partial likelihood, and overdispersed count via a negative-binomial
  representation) and six adjacency specifications (fixed, dual-fixed,
  graphical lasso, Bayesian GGM, and combined GGM-plus-fixed). Within
  `bvs_mh()`, Metropolis-Hastings (MH), Hamiltonian Monte Carlo (HMC), and
  No-U-Turn Sampling (NUTS) are available for non-conjugate binary and
  time-to-event parameter blocks. `bvs_pg()` employs Pólya-Gamma data
  augmentation for exact conjugate updates in binary logistic models. Coupling
  parameters in the Ising/MRF prior are updated via the Möller auxiliary
  variable method, which bypasses the intractable normalizing constant. Dense
  ($p \leq n$) and sparse high-dimensional ($p \gg n$) C++ backends are
  implemented via **Rcpp** and **RcppArmadillo**. We describe the statistical
  model, prior specifications, MCMC algorithms, and package API, and
  demonstrate usage through two simulation studies (moderate: $n=500$, $p=100$;
  high-dimensional: $n=100$, $p=500$) and a real urine metabolomics case-control
  application with sparse external pathway adjacency.
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    self_contained: true
    fig_caption: true
    theme: flatly
    highlight: tango
    df_print: kable
    mathjax: "file:///opt/homebrew/anaconda3/lib/python3.12/site-packages/spyder/plugins/help/utils/js/mathjax/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
vignette: >
  %\VignetteIndexEntry{BVS.DAdj: R Journal Article}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

$$
\require{AMSmath}
\require{AMSsymbols}
\newcommand{\N}{\mathcal{N}}
\newcommand{\bfbeta}{\boldsymbol{\beta}}
\newcommand{\bfgamma}{\boldsymbol{\gamma}}
\newcommand{\bftau}{\boldsymbol{\tau}}
\newcommand{\bfOmega}{\boldsymbol{\Omega}}
\newcommand{\calR}{\mathcal{R}}
$$

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval    = FALSE,
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction {#sec-intro}

In metabolomics, genomics, and other high-throughput biological studies, a
central question is which of many candidate variables---metabolites, genes, or
other molecular features---are genuinely associated with a disease outcome or
phenotype. Structural relationships among these variables, arising from
biological pathways, co-expression networks, or data-driven graphical models,
carry substantial prior information that standard penalization methods largely
ignore.

Classical penalization approaches including the lasso [@Tib1996], elastic net
[@ZH2005], group lasso [@YL2006], and adaptive lasso [@Zou2006] shrink
regression coefficients to zero but treat predictors independently and may
miss weakly active variables embedded in correlated networks. Bayesian
variable selection with spike-and-slab priors [@GM1993; @GM1997] offers a
principled alternative, quantifying uncertainty through posterior inclusion
probabilities (PIPs). A key extension is to encode predictor-level graph
structure through a Markov random field (MRF) prior on the binary selection
indicators, so that adjacent variables in a known or estimated graph tend to
be co-selected [@LZ2010; @STV2011].

The **BVS.DAdj** package extends this Ising/MRF framework to *two
simultaneous* and *independent* adjacency sources: one based on external
domain knowledge (e.g., a biological pathway graph) and one estimated from
the data themselves (e.g., via Bayesian graphical model selection or
graphical lasso). This dual-adjacency formulation allows the model to borrow
strength from both sources while remaining agnostic about their relative
importance, with coupling strengths $\eta_1$ and $\eta_2$ inferred from data
via the Möller auxiliary variable approach [@MB2006].

**Package contributions.** The **BVS.DAdj** package provides:

1. A unified R interface (`bvs_mh` and `bvs_pg`) dispatching to 12
   optimized C++ backends via **Rcpp**/**RcppArmadillo**.
2. Support for *four* outcome types: binary logistic, continuous Gaussian,
   right-censored time-to-event (Cox partial likelihood), and overdispersed
   count (negative-binomial via Pólya-Gamma augmentation).
3. Three within-model samplers for non-conjugate parameter blocks:
   random-walk MH, HMC [@Neal2011], and NUTS [@Hoffman2014].
4. *Six* adjacency specifications covering fixed, data-driven (graphical
   lasso, Bayesian GGM), and dual combinations.
5. Scalable sparse C++ backends for $p \gg n$ with compressed-sparse-column
   (CSC) matrix arithmetic and $O(|\mathcal{E}|)$ neighbourhood sums.
6. An $\eta$ hyperparameter tuning workflow (`phase_transition`,
   `bvs_eta_grid_single`, `bvs_tune_eta`) for choosing Möller reference
   parameters.
7. Post-processing utilities including `summary.bvs`, `plot.bvs`,
   `as.mcmc.bvs`, and `bvs_gelman_diag`.

The remainder of this article is organized as follows.
Section\ 2 describes the statistical model and prior specifications.
Section\ 3 presents the MCMC sampling algorithms.
Section\ 4 describes the package design and API.
Section\ 5 presents Simulation Study 1 (moderate dimensions, $n=500$,
$p=100$).
Section\ 6 presents Simulation Study 2 (high-dimensional sparse, $n=100$,
$p=500$).
Section\ 7 analyzes a real urine metabolomics dataset.
Section\ 8 concludes with discussion.

# Statistical Model and Prior Specifications {#sec-model}

## Linear predictor and notation

Let $\mathbf{X} \in \mathbb{R}^{n \times p}$ be the predictor matrix, where
column $j$ is the $j$-th candidate variable subject to selection, and let
$\mathbf{Z} \in \mathbb{R}^{n \times q}$ be an optional matrix of
always-included covariates. Define the linear predictor for observation $i$:

$$
\eta_i = \alpha + \mathbf{x}_i^\top \boldsymbol{\beta}_{\boldsymbol{\gamma}}
        + \mathbf{z}_i^\top \boldsymbol{\tau},
\quad i = 1, \ldots, n,
$$

where $\alpha$ is an intercept, $\boldsymbol{\gamma} = (\gamma_1, \ldots,
\gamma_p)^\top \in \{0,1\}^p$ is the binary selection indicator vector,
$\boldsymbol{\beta}_{\boldsymbol{\gamma}}$ is the vector of regression
coefficients with $\beta_j = 0$ imposed whenever $\gamma_j = 0$, and
$\boldsymbol{\tau} \in \mathbb{R}^q$ are the always-included covariate
coefficients.

## Outcome likelihoods {#sec-likelihood}

**Binary logistic outcome** (`outcome_type = "binary"`)
$$
y_i \mid \eta_i \sim \mathrm{Bernoulli}\!\left(\mathrm{logit}^{-1}(\eta_i)\right),
\quad y_i \in \{0, 1\}.
$$

**Continuous Gaussian outcome** (`outcome_type = "continuous"`)
$$
y_i \mid \eta_i, \sigma^2 \sim \mathcal{N}(\eta_i,\, \sigma^2).
$$

**Right-censored time-to-event outcome** (`outcome_type = "TTE"`)

Under the Cox proportional hazard model with Breslow tie handling
[@Cox1972], the partial log-likelihood is

$$
\log L(\boldsymbol{\beta}, \boldsymbol{\tau}) =
\sum_{i:\,\delta_i = 1}
\left[\eta_i - \log\!\left(\sum_{j \in \mathcal{R}(t_i)} e^{\eta_j}\right)\right],
$$

where $\delta_i \in \{0,1\}$ is the event indicator and
$\mathcal{R}(t_i) = \{j : t_j \geq t_i\}$ is the risk set at event time
$t_i$. The intercept $\alpha$ is not identifiable under the partial
likelihood and is fixed at zero for this outcome type.

**Overdispersed count outcome** (`outcome_type = "count"`)

A negative-binomial representation via Gamma mixing:

$$
y_i \mid w_i, \eta_i \sim \mathrm{Poisson}\!\left(w_i\, e^{\eta_i}\right),
\qquad
w_i \sim \mathrm{Gamma}(r,\, r),
$$

where $r > 0$ is the dispersion parameter. Marginalizing over $w_i$ gives
a negative-binomial distribution for $y_i$ with mean $e^{\eta_i}$ and
variance $e^{\eta_i}(1 + e^{\eta_i}/r)$.

## Spike-and-slab prior on regression coefficients {#sec-spikeslab}

Conditional on the selection indicators and the variance $\sigma^2$ (or its
analogue), we use the continuous spike-and-slab formulation:

$$
\beta_j \mid \gamma_j, \sigma^2 \;\sim\;
\gamma_j\,\mathcal{N}(\beta_0,\, \sigma^2)
\;+\;
(1 - \gamma_j)\,\delta_0(\beta_j),
\quad j = 1, \ldots, p.
$$

Always-included covariate coefficients receive an independent normal prior:

$$
\tau_k \mid \sigma^2 \sim \mathcal{N}(\tau_0,\, h_\tau\,\sigma^2),
\quad k = 1, \ldots, q.
$$

The intercept (when estimable) is given:

$$
\alpha \sim \mathcal{N}(\alpha_0,\, h_\alpha\,\sigma^2),
$$

and the residual variance (used in continuous outcomes and as the prior
scale for spike-and-slab) follows:

$$
\sigma^2 \sim \mathrm{Inverse\text{-}Gamma}\!\left(\frac{\nu_0}{2},\;
\frac{\nu_0\,\sigma_0^2}{2}\right).
$$

## Dual-adjacency Ising/MRF prior on selection indicators {#sec-ising}

Let $\mathbf{R}^{(1)}, \mathbf{R}^{(2)} \in \{0,1\}^{p \times p}$ be two
symmetric binary adjacency matrices (zero diagonal) encoding two independent
sources of structural information among the $p$ predictors. The joint prior
on $\boldsymbol{\gamma}$ takes the Ising/MRF form:

$$
p\!\left(\boldsymbol{\gamma} \mid \mu, \eta_1, \eta_2\right)
=
\frac{1}{Z(\mu,\eta_1,\eta_2)}
\exp\!\left(
\mu \sum_{j=1}^p \gamma_j
+ \eta_1 \sum_{j < k} R^{(1)}_{jk}\,\gamma_j\gamma_k
+ \eta_2 \sum_{j < k} R^{(2)}_{jk}\,\gamma_j\gamma_k
\right),
$$

where $\mu \in \mathbb{R}$ is the external field controlling baseline
sparsity and $\eta_1, \eta_2 \geq 0$ are non-negative coupling parameters.
For a predictor with no neighbours in either graph, the marginal prior
inclusion probability is $\mathrm{expit}(\mu)$, so $\mu = \mathrm{logit}(\pi_0)$
for target sparsity $\pi_0$.

The normalizing constant $Z(\mu, \eta_1, \eta_2)$ is intractable for general
$p$ and is handled via the Möller auxiliary variable method (Section\ 3.6).
Single-adjacency models use $\eta_2 = 0$ (and hence $\mathbf{R}^{(2)}$
plays no role); the prior then reduces to the standard single-network Ising
form of @LZ2010 and @STV2011.

## Prior on coupling parameters $\eta_1$, $\eta_2$ {#sec-etaprior}

Each coupling parameter is bounded above by a user-specified maximum
(parameters `eta1_sd` and `eta2_sd`) and follows a scaled Beta distribution:

$$
\eta_\ell = \eta_{\ell,\max}\,u_\ell, \quad
u_\ell \sim \mathrm{Beta}(e_\eta,\, f_\eta), \quad
\ell = 1, 2.
$$

The shape parameters $e_\eta$ and $f_\eta$ and the upper bound
$\eta_{\ell,\max}$ are chosen to keep the prior mass below the phase
transition of the Ising model on the given adjacency graph; the
`phase_transition()` helper function profiles this transition before fitting.

## Estimating unknown adjacency via Bayesian GGM {#sec-ggm}

When an adjacency matrix is unknown, the package optionally estimates it
jointly with the regression model by placing a continuous spike-and-slab SSVS
prior on the precision matrix $\boldsymbol{\Omega}$ of the predictors
$\mathbf{X}$, following the Bayesian GGM approach of @WH2015:

$$
p(\boldsymbol{\Omega} \mid \mathbf{R}^{\mathrm{ggm}}, \boldsymbol{\theta})
\propto
\prod_{i < j}
\mathcal{N}\!\left(\omega_{ij} \mid 0,\, v^2_{r^{\mathrm{ggm}}_{ij}}\right)
\prod_i \mathrm{Exp}\!\left(\omega_{ii} \mid \tfrac{\lambda}{2}\right)
\cdot \mathbf{1}\!\left(\boldsymbol{\Omega} \succ 0\right),
$$

where $v_{r^{\mathrm{ggm}}_{ij}} = v_0$ (spike) when $r^{\mathrm{ggm}}_{ij}=0$
(no edge) and $v_1 \gg v_0$ (slab) when $r^{\mathrm{ggm}}_{ij}=1$ (edge
present). The binary edge indicator $r^{\mathrm{ggm}}_{ij}$ is updated via
Bayes' theorem at each MCMC iteration:

$$
\Pr\!\left(r^{\mathrm{ggm}}_{ij} = 1 \mid \omega_{ij}\right)
=
\frac{\pi\,\mathcal{N}(\omega_{ij} \mid 0, v_1^2)}
     {\pi\,\mathcal{N}(\omega_{ij} \mid 0, v_1^2)
      + (1-\pi)\,\mathcal{N}(\omega_{ij} \mid 0, v_0^2)},
$$

where $\pi$ is the prior edge inclusion probability (default
$4 / (p-1)$, targeting sparse graphs). The precision matrix is updated by
the block Gibbs column sweep of @WH2015; in the sparse backend this is
performed in CSC format.

# MCMC Sampling Algorithms {#sec-mcmc}

## Per-iteration structure

Each iteration of either sampler updates in the following block order:

| Step | Update | Both / MH only / PG only |
|:-----|:-------|:------------------------:|
| 1 | Binary selection indicators $\gamma_j$ (add/delete MH move) | both |
| 2 | Active regression coefficients $\boldsymbol{\beta}_\gamma$ | both |
| 3 | Intercept $\alpha$ and covariate coefficients $\boldsymbol{\tau}$ | both |
| 4 | Residual variance $\sigma^2$ | both |
| 5 | Coupling parameters $\eta_1$ (and $\eta_2$ in dual models) | both |
| 6 | GGM precision matrix $\boldsymbol{\Omega}$ (column sweep) | when `adj_type` $\in$ `{"ggm","ggm_fixed"}` |
| 7 | PG latent variables $\omega_i$ | PG only |

In `bvs_mh()`, Steps\ 1–2 are optionally repeated `n_thin_gb` times per
stored sample (inner gamma/beta thinning), improving mixing of the discrete
selection indicators without inflating output storage.

## Metropolis-Hastings sampler (`alg_type = "MH"`) {#sec-mh}

**Selection indicator update (Step\ 1).** For each candidate variable $j$,
a flip $\gamma_j^* = 1 - \gamma_j$ is proposed. If activating the variable
($\gamma_j = 0 \to 1$), a proposal $\beta_j^* \sim \mathcal{N}(\beta_0, \sigma^2)$
is drawn from the slab; if deactivating ($\gamma_j = 1 \to 0$), $\beta_j^*$ is
set to zero. The move is accepted with probability
$\min\{1, \exp(\Delta)\}$ where

$$
\Delta = \ell(\boldsymbol{\beta}^*, \boldsymbol{\gamma}^*) -
         \ell(\boldsymbol{\beta},  \boldsymbol{\gamma})
       + \log p(\boldsymbol{\gamma}^* \mid \mu, \eta_1, \eta_2)
       - \log p(\boldsymbol{\gamma}  \mid \mu, \eta_1, \eta_2)
       + \log p(\beta_j^*) - \log p(\beta_j).
$$

**Regression coefficient update (Step\ 2).**

- *Binary / count outcomes*: random-walk MH perturbation of the active
  $\boldsymbol{\beta}_\gamma$ block.
- *Continuous outcome*: exact conjugate Gibbs draw for active $\boldsymbol{\beta}_\gamma$
  from the Gaussian conditional posterior.
- *TTE outcome*: Fisher-scoring random-walk MH perturbation of the active
  block.

All MH acceptance steps use the numerically robust helper
`safe_mh_accept(log_ratio)` which avoids the classic `log(runif(0,1)) < log_ratio`
deadlock at $-\infty$.

## Hamiltonian Monte Carlo (`alg_type = "HMC"`) {#sec-hmc}

For binary and TTE outcomes, `bvs_mh(..., alg_type = "HMC")` replaces the
random-walk MH update on the active parameter block with Hamiltonian
dynamics [@Neal2011]. Let $\mathbf{q}$ denote the active parameter vector
$(\boldsymbol{\beta}_\gamma, \alpha, \boldsymbol{\tau})$ (with $\alpha$
excluded for TTE) and $U(\mathbf{q}) = -\log\pi(\mathbf{q} \mid \cdot)$
the potential energy. With auxiliary momentum $\mathbf{p} \sim \mathcal{N}(\mathbf{0}, \mathbf{M})$
(mass matrix $\mathbf{M}$), the Hamiltonian is

$$
H(\mathbf{q}, \mathbf{p})
= U(\mathbf{q}) + \tfrac{1}{2}\mathbf{p}^\top \mathbf{M}^{-1} \mathbf{p}.
$$

Each HMC sub-step:

1. Sample momentum $\mathbf{p} \sim \mathcal{N}(\mathbf{0}, \mathbf{M})$.
2. Run $L$ leapfrog steps of size $\epsilon$:
   $\mathbf{p} \leftarrow \mathbf{p} - \frac{\epsilon}{2}\nabla_{\mathbf{q}} U(\mathbf{q})$;
   $\mathbf{q} \leftarrow \mathbf{q} + \epsilon\, \mathbf{M}^{-1}\mathbf{p}$;
   $\mathbf{p} \leftarrow \mathbf{p} - \frac{\epsilon}{2}\nabla_{\mathbf{q}} U(\mathbf{q})$.
3. Accept/reject: $\alpha_{\mathrm{HMC}} = \min\{1, \exp[-H(\mathbf{q}', \mathbf{p}') + H(\mathbf{q}, \mathbf{p})]\}$.

The gradient $\nabla_{\mathbf{q}} U$ is computed analytically for each
outcome type. For binary logistic:
$\nabla_{\beta_j} U = -\sum_i x_{ij}(y_i - \hat{\pi}_i) + \beta_j/\sigma^2$
where $\hat{\pi}_i = \mathrm{logit}^{-1}(\eta_i)$.
For Cox TTE:
$\nabla_{\beta_j} U = -\sum_{i:\delta_i=1} x_{ij} + \sum_{i:\delta_i=1} \bar{x}_j(t_i) + \beta_j/\sigma^2$
where $\bar{x}_j(t_i) = \sum_{k\in\mathcal{R}(t_i)} x_{kj} e^{\eta_k} / \sum_{k\in\mathcal{R}(t_i)} e^{\eta_k}$.

## No-U-Turn Sampler (`alg_type = "NUTS"`) {#sec-nuts}

NUTS [@Hoffman2014] removes the need to specify the number of leapfrog steps
$L$ by adaptively growing a balanced binary tree of trajectory
states until a U-turn is detected:

$$
({\mathbf{q}}^+ - {\mathbf{q}}^-)^\top \mathbf{M}^{-1} {\mathbf{p}}^- < 0
\quad \text{or} \quad
({\mathbf{q}}^+ - {\mathbf{q}}^-)^\top \mathbf{M}^{-1} {\mathbf{p}}^+ < 0,
$$

where $\mathbf{q}^-$, $\mathbf{q}^+$ are the leftmost/rightmost tree
positions and $\mathbf{p}^-$, $\mathbf{p}^+$ the corresponding momenta.
A proposal is drawn uniformly from the built trajectory with detailed
balance, and the step size $\epsilon$ is tuned during burn-in via dual
averaging. The maximum tree depth is controlled by `nuts_max_treedepth`
(default 10).

Compared with fixed-step HMC, NUTS reduces sensitivity to $L$ specification
and often achieves higher effective sample size per second in strongly
correlated parameter spaces.

## Pólya-Gamma augmented Gibbs (`bvs_pg`) {#sec-pg}

For binary logistic outcomes, `bvs_pg()` introduces Pólya-Gamma latent
variables $\omega_i \sim \mathrm{PG}(1, \eta_i)$ [@PG2013]. Conditional
on $\boldsymbol{\omega} = (\omega_1, \ldots, \omega_n)^\top$, the logistic
likelihood becomes conjugate Gaussian in $(\alpha, \boldsymbol{\beta}_\gamma)$,
enabling the exact Gibbs draw:

$$
\begin{pmatrix}\alpha \\ \boldsymbol{\beta}_\gamma\end{pmatrix}
\;\Bigg|\; \boldsymbol{\omega}, \boldsymbol{\gamma}, \sigma^2
\;\sim\;
\mathcal{N}\!\left(\mathbf{V}_\gamma \mathbf{m}_\gamma,\; \mathbf{V}_\gamma\right),
$$

where
$\mathbf{V}_\gamma =
  \left(\tilde{\mathbf{X}}_\gamma^\top \mathrm{diag}(\boldsymbol{\omega})\,
        \tilde{\mathbf{X}}_\gamma + \sigma^{-2}\mathbf{I}\right)^{-1}$,
$\mathbf{m}_\gamma = \tilde{\mathbf{X}}_\gamma^\top
  \boldsymbol{\kappa}$,
$\kappa_i = y_i - 1/2$,
and $\tilde{\mathbf{X}}_\gamma = [\mathbf{1}, \mathbf{X}_\gamma]$ includes
the intercept column. PG draws are obtained via the truncated-series
approximation of @PG2013 implemented in C++.

## Möller auxiliary variable update for $\eta_1$, $\eta_2$ {#sec-moller}

The normalizing constant $Z(\mu, \eta_1, \eta_2)$ of the Ising prior is
intractable for general $p$. Following @MB2006, we introduce an auxiliary
variable $\boldsymbol{\omega}^*$ drawn via Propp-Wilson perfect simulation
from the Ising model at fixed reference parameters
$(\tilde{\mu}, \tilde{\eta}_1, \tilde{\eta}_2)$. The resulting MH ratio
for a proposal $\eta_\ell' \sim q(\eta_\ell' \mid \eta_\ell)$ is:

$$
\alpha_{\mathrm{Möller}} = \min\!\left\{1,\;
\frac{p(\boldsymbol{\gamma} \mid \mu, \eta_1', \eta_2')\,
      p(\eta_\ell')\,
      f(\boldsymbol{\omega}^* \mid \tilde{\mu}, \tilde{\eta}_1', \tilde{\eta}_2')}
     {p(\boldsymbol{\gamma} \mid \mu, \eta_1,  \eta_2)\,
      p(\eta_\ell)\,
      f(\boldsymbol{\omega}^* \mid \tilde{\mu}, \tilde{\eta}_1,  \tilde{\eta}_2)}\right\},
$$

where the intractable $Z$ cancels exactly because the same auxiliary sample
$\boldsymbol{\omega}^*$ is used in numerator and denominator. Propp-Wilson
simulation terminates in finite time and the maximum doubling horizon is
controlled by `Tmax` (default 64).

## Sparse high-dimensional backend {#sec-sparse}

For $p \gg n$, the package stores the scatter matrix
$\mathbf{S} = \mathbf{X}^\top\mathbf{X}$ in compressed sparse column (CSC)
format (via `prepare_sparse_S()`), thresholding near-zero entries for
storage efficiency. Ising neighbourhood sums
$\sum_k R^{(1)}_{jk} \gamma_k$ are computed in $O(\deg_j)$ time, where
$\deg_j$ is the degree of variable $j$ in the adjacency graph. The GGM
column sweep updates one precision-matrix column at a time using only the
non-zero entries of the current column of $\mathbf{S}$. The selection
indicator update cycles through `n_mh_gamma` randomly chosen predictors per
outer iteration to reduce per-iteration cost.

## MCMC mixing and convergence improvements {#sec-mcmc-improvements}

Three improvements are implemented uniformly across all 12 C++ backends and the R
post-processing layer.

### Locally-balanced $\gamma$ proposals {#sec-lb-gamma}

Uniform-coordinate random-walk proposals for the binary selection indicators
$\gamma$ can mix slowly when most proposed flips have low acceptance
probability. Following @Zanella2020, the package replaces uniform coordinate
selection with a *locally balanced* distribution. For each candidate variable
$j$, the unnormalised weight

$$
w_j = \sqrt{ \pi(\boldsymbol{\gamma}^{(j)}) / \pi(\boldsymbol{\gamma}) }
     = \exp\!\bigl( 0.5 \cdot \mathrm{clip}_{[-60,60]}\bigl[
       (1 - 2\gamma_j)(\mu + \eta_1 N_j^{(1)} + \eta_2 N_j^{(2)}) \bigr] \bigr),
$$

where $N_j^{(\ell)} = \sum_k R^{(\ell)}_{jk}\gamma_k$ is the active-neighbour count
in adjacency $\ell$ and $\mathrm{clip}$ prevents numerical overflow.
The balancing function $g(t) = \sqrt{t}$ corresponds to the geometric
mean of the target ratios, yielding the "square-root" or Barker kernel in
the sense of @Zanella2020. Coordinate $j$ is selected with probability
proportional to $w_j$, and the MH ratio is corrected by the log
proposal ratio $\log(q_\mathrm{rev}/q_\mathrm{fwd})$.

Activated via `use_lb_gamma = TRUE` (default). Supported by all 12 backends.

**Reference:** @Zanella2020.

### Robust Adaptive Metropolis for $\eta$ {#sec-ram}

The Möller MH proposal SD for the coupling parameters $\eta_1$, $\eta_2$
is adapted online using the Robust Adaptive Metropolis (RAM) algorithm
of @Vihola2012. Letting $\log\sigma_n$ denote the log-scale proposal
standard deviation after $n$ adaptations, the Robbins-Monro update is

$$
\log\sigma_{n+1} = \log\sigma_n + \tfrac{1}{2} \min(1, n^{-2/3})
                   \bigl(\hat{\alpha}_n - \alpha^*\bigr),
$$

where $\hat{\alpha}_n \in \{0,1\}$ is the accept-indicator for iteration $n$
and $\alpha^* = 0.44$ is the optimal 1D acceptance rate
[@Roberts1997OptimalScaling]. The $n^{-2/3}$ diminishing step satisfies
the diminishing-adaptation condition (Theorem 2 of @Vihola2012), ensuring
ergodicity.

An `EtaAdapter` struct in `BayesLogit_Sparse_Helpers.h` encapsulates the
log-sigma state and the update rule; each backend declares one adapter per
$\eta$ parameter.

**References:** @Vihola2012; @AndrieuThoms2008.

### Rank-normalised $\hat{R}$ and bulk/tail ESS {#sec-rhat}

Classical Gelman-Rubin $\hat{R}$ and `effectiveSize` (ESS) assume approximately
Gaussian chains. @Vehtari2021 propose three improvements:

1. **Split chains**: each chain is split in half before computing $\hat{R}$,
   detecting within-chain non-stationarity.
2. **Rank normalisation**: draws are replaced by their normal quantiles,
   making the diagnostic robust to heavy tails.
3. **Bulk and tail ESS**: bulk ESS (via Geyer's initial positive sequence
   estimator) measures efficiency for the body of the distribution; tail ESS
   = $\min(\mathrm{ESS}_{0.05}, \mathrm{ESS}_{0.95})$ measures tail efficiency.

These diagnostics are available as exported R functions:

```{r diag-functions, eval=FALSE}
bvs_rhat(x)       # rank-normalised split-Rhat; target < 1.01
bvs_ess_bulk(x)   # bulk ESS
bvs_ess_tail(x)   # tail ESS = min(ESS_q=0.05, ESS_q=0.95)
```

`x` can be a numeric vector (single chain, split in half automatically) or
a matrix with one column per chain. When the `posterior` package is
installed, these functions delegate to `posterior::rhat()`,
`posterior::ess_bulk()`, and `posterior::ess_tail()`; otherwise a pure-R
fallback implementing the same algorithm is used.
`summary.bvs()` reports `ESS` (bulk) and `ESS_tail` for each predictor.

**Reference:** @Vehtari2021.

# Package Design and API {#sec-api}

## Installation

```{r install, eval=FALSE}
# From GitHub:
devtools::install_github("yuy113/BVS.DAdj")
```

```{r load-pkg}
library(BVS.DAdj)
```

## User-facing functions

**BVS.DAdj** exports two primary sampling functions that automatically
dispatch to 12 C++ backends based on `sampler × adj_type × sparse`:

| Function   | Sampler | Outcome types | Notes |
|:-----------|:--------|:--------------|:------|
| `bvs_mh()` | MH / HMC / NUTS | binary, continuous, TTE, count | HMC/NUTS for binary & TTE |
| `bvs_pg()` | Pólya-Gamma Gibbs | binary only | Exact beta/alpha updates |

## Adjacency types {#sec-adj}

| `adj_type` | Adjacency source(s) | Free $\eta$ | C++ backends |
|:-----------|:--------------------|:-----------:|:-------------|
| `"fixed"` | One user-supplied binary matrix `adj_fixed` | $\eta_1$ | MH + PG, dense |
| `"dual_fixed"` | Two user-supplied matrices (`adj_fixed`, `adj_fixed2`) | $\eta_1, \eta_2$ | MH + PG, dense |
| `"glasso"` | Graphical lasso from $\mathbf{X}$ (EBIC or RIC) | $\eta_1$ | MH + PG, dense |
| `"glasso_fixed"` | Graphical lasso + one fixed matrix | $\eta_1, \eta_2$ | MH + PG, dense |
| `"ggm"` | Bayesian GGM SSVS, updated at every iteration | $\eta_1$ | MH + PG, dense/sparse |
| `"ggm_fixed"` | Bayesian GGM + one fixed matrix | $\eta_1, \eta_2$ | MH + PG, dense/sparse |

For `"glasso"` and `"glasso_fixed"`, the adjacency is estimated once
before MCMC using the **huge** package [@FHT2008] and held fixed
throughout. For `"ggm"` and `"ggm_fixed"`, the adjacency is updated at
every MCMC iteration via the GGM column sweep.

## Core function signatures

```{r api-mh, eval=FALSE}
bvs_mh(
  X, y,
  event        = NULL,            # event indicator for TTE (1=event, 0=censored)
  outcome_type = "binary",        # "binary" | "continuous" | "TTE" | "count"
  adj_type     = "fixed",         # see adjacency type table above
  adj_fixed    = NULL,            # p x p binary adjacency matrix
  adj_fixed2   = NULL,            # second p x p matrix (dual_fixed only)
  sparse       = FALSE,           # TRUE for sparse GGM backend (p >> n)
  S_ggm        = NULL,            # pre-computed CSC scatter matrix
  niter        = 60000L, burnin = 10000L, thin = 1L,
  n_thin_gb    = 3L,              # inner gamma/beta sub-iterations (dense)
  n_mh_gamma   = 3L,              # gamma flip proposals per iteration (sparse)
  # -- Sparsity / variance priors --
  mu       = -2.197,              # logit(pi_0); default pi_0 = 10%
  nu0      = 2,   sigmasq0 = 1.5,
  h        = 1.5, alpha0 = 0, beta0 = 0,
  # -- Eta / Moller parameters --
  eta1_sd  = 0.5, eta2_sd  = 0.5,
  mu_tilde = -4,  eta1_tilde = 0.5, eta2_tilde = 0.5,
  e_eta    = 1,   f_eta = 1, Tmax = 64L,
  # -- GGM SSVS priors --
  v0_ggm = 0.015^2, v1_ggm = NULL, pii_ggm = NULL, lambda_ggm = 1,
  # -- Always-included covariates --
  z_dat = NULL, tau0 = 0, htau = 1.5, tau_init = NULL,
  # -- HMC/NUTS --
  alg_type = "MH", hmc_step_size = 0.1,
  hmc_n_leapfrog = 10L, nuts_max_treedepth = 10L,
  # -- Storage (sparse mode) --
  store_beta = FALSE, store_gamma = FALSE,
  store_Z_pip = TRUE, store_Z_list = FALSE,
  # -- Initialisation --
  beta_init = NULL, gamma_init = NULL, alpha_init = NULL
)
```

```{r api-pg, eval=FALSE}
bvs_pg(
  X, y,
  adj_type  = "fixed",
  adj_fixed = NULL, adj_fixed2 = NULL,
  sparse    = FALSE, S_ggm = NULL,
  niter = 60000L, burnin = 10000L, thin = 1L,
  n_mh_gamma = 3L,
  mu = -2.197, nu0 = 2, sigmasq0 = 1.5, h = 1.5,
  alpha0 = 0, beta0 = 0,
  eta1_sd = 0.5, eta2_sd = 0.5,
  mu_tilde = -4, eta1_tilde = 0.5, eta2_tilde = 0.5,
  e_eta = 1, f_eta = 1, Tmax = 64L,
  v0_ggm = 0.015^2, v1_ggm = NULL, pii_ggm = NULL, lambda_ggm = 1,
  z_dat = NULL, tau0 = 0, htau = 1.5, tau_init = NULL,
  store_beta = FALSE, store_gamma = FALSE,
  store_Z_pip = TRUE, store_Z_list = FALSE,
  beta_init = NULL, gamma_init = NULL, alpha_init = NULL
)
```

## Key hyperparameter guide

| Parameter | Default | Interpretation |
|:----------|:--------|:---------------|
| `mu` | $\mathrm{logit}(0.10) \approx -2.197$ | Baseline log-odds of inclusion; set to $\mathrm{logit}(\pi_0)$ for target sparsity $\pi_0$ |
| `eta1_sd`, `eta2_sd` | `0.5` | Upper bound of $\eta_\ell$; keep below phase-transition critical value |
| `mu_tilde`, `eta1_tilde`, `eta2_tilde` | `-4`, `0.5`, `0.5` | Möller reference parameters; should correspond to a sparser reference model |
| `e_eta`, `f_eta` | `1`, `1` | Beta shape for $u_\ell \sim \mathrm{Beta}(e,f)$; increase `f_eta` to shrink $\eta_\ell$ toward zero |
| `n_thin_gb` | `3` | Inner gamma/beta sub-iterations per stored sample (dense MH only) |
| `v0_ggm` | `0.015^2` | GGM spike variance (small = strong near-zero shrinkage for non-edges) |
| `pii_ggm` | `4/(p-1)` | GGM prior edge inclusion probability (sparse default) |

## Return value

Both functions return an S3 object of class `"bvs"`:

| Field | Description |
|:------|:------------|
| `gamma_pip` | Length-$p$ vector of posterior inclusion probabilities |
| `beta` | $N_{\text{saved}} \times p$ matrix (dense) or sparse list (sparse) of $\beta$ draws |
| `gamma` | $N_{\text{saved}} \times p$ matrix (dense) or sparse index list of $\gamma$ draws |
| `alpha` | Length-$N_{\text{saved}}$ intercept chain |
| `sigmasq` | Length-$N_{\text{saved}}$ variance chain |
| `eta1`, `eta2` | Coupling parameter chains |
| `Z_pip` | GGM edge PIP matrix (when `store_Z_pip = TRUE`) |
| `sampler`, `adj_type`, `outcome_type` | Metadata strings |
| `p`, `n`, `niter`, `burnin` | Dimension / MCMC metadata |

## Post-processing utilities

```{r postproc, eval=FALSE}
# Summarise: PIPs, posterior means, credible intervals, HPD, ESS
s <- summary(fit, pip_threshold = 0.5, cred_level = 0.95)
cat("Selected variables:", s$selected, "\n")
cat("Model size:", s$nselected, "\n")
print(s$summary_beta[s$selected, ])

# Plot: PIP bar chart, trace plots, ACF
plot(fit, type = "pip", pip_threshold = 0.5)
plot(fit, type = "trace", top_n = 6)
plot(fit, type = "acf",   vars = c(1:5))

# Convert to coda mcmc for diagnostics
library(coda)
mc <- as.mcmc.bvs(fit, vars = 1:20)
effectiveSize(mc)

# Gelman-Rubin PSRF across two independent chains
diag <- bvs_gelman_diag(list(fit1, fit2), vars = 1:20)
print(diag$psrf[1:10, ])   # per-variable PSRF (target < 1.1)
```

## Eta hyperparameter tuning workflow

```{r eta-tuning, eval=FALSE}
# Stage 1: profile the phase transition of the Ising prior
pt <- phase_transition(R,
        mu = -2.197, min_eta = 0, max_eta = 1.5,
        step_size = 0.05, num_rep = 10, Tmax = 64L)
matplot(pt, type = "l", xlab = "eta index", ylab = "model size",
        main = "Ising phase transition")

# Stage 2: build candidate hyperparameter grid
grid <- bvs_eta_grid_single(
  adj       = R,
  eta1_frac = seq(0.2, 0.8, by = 0.2),
  mu_tilde  = c(-4, -5),
  eta1_tilde = c(0.05, 0.10)
)

# Stage 3: select by FDR-controlled PIP threshold
best <- bvs_tune_eta_fixed_pip(
  X, y,
  adj_type    = "fixed", adj_fixed = R,
  sampler     = "pg",
  eta_grid    = grid,
  niter       = 3000L, burnin = 500L,
  pip_threshold_fixed = 0.5,
  target_fdr  = 0.05
)
cat("Best eta1_sd:", best$best_row$eta1_sd, "\n")
```

# Simulation Study 1: Moderate Dimensions ($n = 500$, $p = 100$) {#sec-sim1}

## Study design {#sec-sim1-design}

We generate a binary logistic regression dataset with $n = 500$ observations,
$p = 100$ covariates, and $p^* = 10$ truly active variables. The correlation
structure emulates a metabolite clustering scenario: the first $p^*$ covariates
form a block with pairwise correlation $\rho = 0.5$, while the remaining
$p - p^*$ predictors are independent standard normals.

```{r sim1-data}
set.seed(2024)
n      <- 500L;  p      <- 100L;  p_star <- 10L

# Block covariance for active predictors
Sigma <- diag(p)
Sigma[1:p_star, 1:p_star] <- 0.5
diag(Sigma) <- 1.0

X      <- MASS::mvrnorm(n, mu = rep(0, p), Sigma = Sigma)
beta_true <- rep(0, p)
beta_true[1:p_star] <- c(1.5, -1.2, 1.0, -0.8,  0.6,
                         -0.5,  0.4, -0.3, 0.25, -0.2)
y <- rbinom(n, 1L, plogis(X %*% beta_true))

cat(sprintf("Cases: %d  Controls: %d  (%.1f%% cases)\n",
            sum(y), sum(1 - y), 100 * mean(y)))
```

## Adjacency construction {#sec-sim1-adj}

We compare five adjacency strategies against the same dataset:

1. **Fixed (50% true edges)**: an external binary adjacency matrix derived
   by randomly retaining 50% of edges from the true active-block graph.
   This mimics partial, noisy prior knowledge (e.g., a biological pathway
   database covering only a subset of true relationships).
2. **GGM**: a Bayesian GGM SSVS adjacency learned jointly during MCMC.
3. **Glasso**: a graphical-lasso adjacency estimated once before MCMC
   using the EBIC criterion.
4. **Glasso + Fixed**: dual-adjacency combining the glasso-learned graph
   with the fixed external matrix (coupling parameters $\eta_1, \eta_2$).
5. **GGM + Fixed**: dual-adjacency combining GGM with the fixed external
   matrix.

```{r sim1-adj}
# True active-block adjacency
R_true1 <- matrix(0L, p, p)
R_true1[1:p_star, 1:p_star] <- 1L
diag(R_true1) <- 0L

# External fixed adjacency: random 50% of true edges
set.seed(202401)
true_pairs  <- which(upper.tri(R_true1) & R_true1 == 1L, arr.ind = TRUE)
keep_pairs  <- true_pairs[sample(nrow(true_pairs), ceiling(0.5 * nrow(true_pairs))), ]
R_ext50     <- matrix(0L, p, p)
R_ext50[cbind(keep_pairs[, 1], keep_pairs[, 2])] <- 1L
R_ext50[cbind(keep_pairs[, 2], keep_pairs[, 1])] <- 1L
diag(R_ext50) <- 0L

cat("True edges:", sum(R_true1) / 2,
    "  External-50% edges:", sum(R_ext50) / 2, "\n")
```

## Five-method MCMC comparison (MH and PG) {#sec-sim1-run}

```{r sim1-run}
# Hyperparameters tuned for ~10% expected sparsity
MU   <- qlogis(0.01)       # strong sparsity prior
ETA1 <- 0.08; ETA2 <- 0.06 # coupling upper bounds
E_ETA <- 1.0; F_ETA <- 25.0
MU_T  <- -7.0; E1T <- 0.005; E2T <- 0.004

niter_s1 <- 60000L; burnin_s1 <- 20000L; thin_s1 <- 1L

# --- MH runs ---
fit_mh_fixed <- bvs_mh(X, y,
  adj_type  = "fixed", adj_fixed = R_ext50,
  niter = niter_s1, burnin = burnin_s1, thin = thin_s1,
  mu = MU, eta1_sd = ETA1, n_thin_gb = 3L)

fit_mh_ggm <- bvs_mh(X, y,
  adj_type = "ggm", sparse = FALSE,
  niter = niter_s1, burnin = burnin_s1, thin = thin_s1,
  mu = MU, eta1_sd = ETA1, n_thin_gb = 3L)

fit_mh_glasso <- bvs_mh(X, y,
  adj_type = "glasso", glasso_criterion = "ebic",
  niter = niter_s1, burnin = burnin_s1, thin = thin_s1,
  mu = MU, eta1_sd = ETA1, n_thin_gb = 3L)

fit_mh_glasso_fixed <- bvs_mh(X, y,
  adj_type  = "glasso_fixed", adj_fixed = R_ext50,
  glasso_criterion = "ebic",
  niter = niter_s1, burnin = burnin_s1, thin = thin_s1,
  mu = MU, eta1_sd = ETA1, eta2_sd = ETA2,
  mu_tilde = MU_T, eta1_tilde = E1T, eta2_tilde = E2T,
  e_eta = E_ETA, f_eta = F_ETA, n_thin_gb = 3L)

fit_mh_ggm_fixed <- bvs_mh(X, y,
  adj_type = "ggm_fixed", adj_fixed = R_ext50, sparse = FALSE,
  niter = niter_s1, burnin = burnin_s1, thin = thin_s1,
  mu = MU, eta1_sd = ETA1, eta2_sd = ETA2,
  mu_tilde = MU_T, eta1_tilde = E1T, eta2_tilde = E2T,
  e_eta = E_ETA, f_eta = F_ETA, n_thin_gb = 3L)

# --- PG runs ---
fit_pg_fixed <- bvs_pg(X, y,
  adj_type = "fixed", adj_fixed = R_ext50,
  niter = niter_s1, burnin = burnin_s1, thin = thin_s1,
  mu = MU, eta1_sd = ETA1)

fit_pg_ggm <- bvs_pg(X, y,
  adj_type = "ggm", sparse = FALSE,
  niter = niter_s1, burnin = burnin_s1, thin = thin_s1,
  mu = MU, eta1_sd = ETA1)

fit_pg_glasso <- bvs_pg(X, y,
  adj_type = "glasso", glasso_criterion = "ebic",
  niter = niter_s1, burnin = burnin_s1, thin = thin_s1,
  mu = MU, eta1_sd = ETA1)

fit_pg_glasso_fixed <- bvs_pg(X, y,
  adj_type = "glasso_fixed", adj_fixed = R_ext50,
  glasso_criterion = "ebic",
  niter = niter_s1, burnin = burnin_s1, thin = thin_s1,
  mu = MU, eta1_sd = ETA1, eta2_sd = ETA2,
  mu_tilde = MU_T, eta1_tilde = E1T, eta2_tilde = E2T,
  e_eta = E_ETA, f_eta = F_ETA)

fit_pg_ggm_fixed <- bvs_pg(X, y,
  adj_type = "ggm_fixed", adj_fixed = R_ext50, sparse = FALSE,
  niter = niter_s1, burnin = burnin_s1, thin = thin_s1,
  mu = MU, eta1_sd = ETA1, eta2_sd = ETA2,
  mu_tilde = MU_T, eta1_tilde = E1T, eta2_tilde = E2T,
  e_eta = E_ETA, f_eta = F_ETA)
```

## Results {#sec-sim1-results}

For each fitted model, we evaluate:

- **Posterior inclusion probabilities (PIPs)** $\widehat{\pi}_j = \Pr(\gamma_j = 1 \mid \mathbf{y}, \mathbf{X})$ computed as the fraction of post-burn-in MCMC samples with $\gamma_j = 1$.
- **Variable-selection performance** at threshold $c = 0.5$: true positives (TP), false positives (FP), false discovery rate
  $\mathrm{FDR} = \mathrm{FP} / (\mathrm{TP} + \mathrm{FP})$,
  sensitivity $= \mathrm{TP}/p^*$, and F1 score.
- **MCMC diagnostics** for the representative coefficient $\beta_1$ (the first and most strongly active variable):
  lag-1 autocorrelation $\rho_1$, effective sample size
  $\mathrm{ESS} = N / (1 + 2\sum_{\ell=1}^L \rho_\ell)$,
  Monte Carlo standard error $\mathrm{MCSE} = \hat\sigma_\beta / \sqrt{\mathrm{ESS}}$,
  and per-second ESS.

```{r sim1-results}
# Helper: extract PIPs
compute_pip <- function(fit) {
  if (!is.null(fit$gamma_pip)) return(as.numeric(fit$gamma_pip))
  if (is.matrix(fit$gamma))    return(colMeans(fit$gamma))
  stop("No gamma samples available.")
}

# Helper: compute selection metrics
compute_metrics <- function(pip, threshold = 0.5, p_star = 10L) {
  sel <- which(pip >= threshold)
  tp  <- sum(sel <= p_star)
  fp  <- sum(sel >  p_star)
  fn  <- p_star - tp
  data.frame(
    TP          = tp,   FP = fp, FN = fn,
    Sensitivity = round(tp / p_star, 3),
    FDR         = round(if (length(sel) > 0) fp / length(sel) else 0, 3),
    F1          = round(if (tp + fp + fn > 0) 2*tp/(2*tp+fp+fn) else 0, 3),
    N_selected  = length(sel)
  )
}

fits_s1 <- list(
  "MH | Fixed"         = fit_mh_fixed,
  "MH | GGM"           = fit_mh_ggm,
  "MH | Glasso"        = fit_mh_glasso,
  "MH | Glasso+Fixed"  = fit_mh_glasso_fixed,
  "MH | GGM+Fixed"     = fit_mh_ggm_fixed,
  "PG | Fixed"         = fit_pg_fixed,
  "PG | GGM"           = fit_pg_ggm,
  "PG | Glasso"        = fit_pg_glasso,
  "PG | Glasso+Fixed"  = fit_pg_glasso_fixed,
  "PG | GGM+Fixed"     = fit_pg_ggm_fixed
)

metrics_s1 <- do.call(rbind, lapply(names(fits_s1), function(nm)
  cbind(Method = nm, compute_metrics(compute_pip(fits_s1[[nm]])))
))
knitr::kable(metrics_s1, row.names = FALSE,
  caption = paste0("Simulation 1 ($n=500$, $p=100$, $p^*=10$): ",
                   "variable-selection performance at PIP threshold 0.5."))
```

```{r sim1-pip-plot, fig.height=8, fig.width=12, fig.cap="Simulation 1: PIPs for the first 40 variables across ten method-sampler combinations. Red bars indicate truly active variables (j = 1,...,10); grey bars indicate null variables. Dashed line at PIP = 0.5."}
pips_s1  <- lapply(fits_s1, compute_pip)
show_idx <- 1:40
col_vec  <- ifelse(show_idx <= p_star, "firebrick", "grey65")

par(mfrow = c(2, 5), mar = c(3, 3, 2, 1))
for (nm in names(pips_s1)) {
  barplot(pips_s1[[nm]][show_idx], col = col_vec, border = NA,
          ylim = c(0, 1), main = nm, ylab = "PIP",
          names.arg = show_idx, las = 2, cex.names = 0.5)
  abline(h = 0.5, lty = 2)
}
```

```{r sim1-diagnostics}
# MCMC diagnostics for beta_1 chain
diag_s1 <- do.call(rbind, lapply(names(fits_s1), function(nm) {
  beta1 <- as.numeric(fits_s1[[nm]]$beta[, 1L])
  ess   <- coda::effectiveSize(beta1)
  data.frame(
    Method          = nm,
    PIP_V1          = round(compute_pip(fits_s1[[nm]])[1L], 3),
    Lag1_ACF        = round(acf(beta1, lag.max = 1L, plot = FALSE)$acf[2L], 3),
    ESS             = round(ess, 1),
    MCSE            = round(sd(beta1) / sqrt(max(1, ess)), 4),
    stringsAsFactors = FALSE
  )
}))
knitr::kable(diag_s1, row.names = FALSE,
  caption = "Simulation 1: MCMC diagnostics for the first active coefficient $\\beta_1$.")
```


# Simulation Study 2: High-Dimensional Sparse Setting ($n = 100$, $p = 500$) {#sec-sim2}

## Study design {#sec-sim2-design}

We consider a challenging high-dimensional scenario with $n = 100$
observations, $p = 500$ covariates, and $p^* = 50$ truly active
variables ($10\%$ of predictors). The first $p^*$ covariates share
pairwise correlation $\rho = 0.35$; the remaining $p - p^*$ are
independent standard normals. True coefficients decay linearly from
$\pm 1.2$ to $\pm 0.2$ in alternating sign.

```{r sim2-data}
set.seed(2025)
n_hd     <- 100L; p_hd <- 500L; p_star_hd <- 50L

Sigma_hd <- diag(p_hd)
Sigma_hd[1:p_star_hd, 1:p_star_hd] <- 0.35
diag(Sigma_hd) <- 1.0

X_hd <- MASS::mvrnorm(n_hd, mu = rep(0, p_hd), Sigma = Sigma_hd)
beta_true_hd <- rep(0, p_hd)
beta_true_hd[1:p_star_hd] <-
  seq(1.2, 0.2, length.out = p_star_hd) * rep(c(1, -1), length.out = p_star_hd)
y_hd <- rbinom(n_hd, 1L, plogis(X_hd %*% beta_true_hd))
```

## Adjacency construction {#sec-sim2-adj}

The same five-method comparison is used (Fixed-50%, GGM-sparse, Glasso,
Glasso+Fixed, GGM+Fixed-sparse). Since $p > n$, the Bayesian GGM
backends are run with `sparse = TRUE`, which stores $\mathbf{S} = \mathbf{X}^\top\mathbf{X}$
in CSC format and performs $O(|\mathcal{E}|)$ neighbourhood operations.

```{r sim2-adj}
# True adjacency: first p_star block
R_true2 <- matrix(0L, p_hd, p_hd)
R_true2[1:p_star_hd, 1:p_star_hd] <- 1L
diag(R_true2) <- 0L

# External fixed: random 50% of true edges
set.seed(202502)
tp2 <- which(upper.tri(R_true2) & R_true2 == 1L, arr.ind = TRUE)
kp2 <- tp2[sample(nrow(tp2), ceiling(0.5 * nrow(tp2))), ]
R_ext50_2 <- matrix(0L, p_hd, p_hd)
R_ext50_2[cbind(kp2[, 1], kp2[, 2])] <- 1L
R_ext50_2[cbind(kp2[, 2], kp2[, 1])] <- 1L
diag(R_ext50_2) <- 0L

# Pre-compute sparse scatter matrix for GGM backends
S_csc <- prepare_sparse_S(X_hd, threshold = 1e-4)
```

## Five-method comparison under MH with sparse GGM {#sec-sim2-run}

```{r sim2-run}
MU2   <- qlogis(0.02); ETA2_1 <- 0.05; ETA2_2 <- 0.05
E_ETA2 <- 1.0; F_ETA2 <- 30.0
MU2_T  <- qlogis(0.003); E2_1T <- 0.01; E2_2T <- 0.008

niter_s2 <- 50000L; burnin_s2 <- 20000L; thin_s2 <- 1L

fit_s2_fixed <- bvs_mh(X_hd, y_hd,
  adj_type = "fixed", adj_fixed = R_ext50_2,
  niter = niter_s2, burnin = burnin_s2, thin = thin_s2,
  mu = MU2, eta1_sd = ETA2_1, n_mh_gamma = 3L)

fit_s2_ggm <- bvs_mh(X_hd, y_hd,
  adj_type = "ggm", sparse = TRUE, S_ggm = S_csc,
  store_gamma = TRUE, store_beta = TRUE, store_Z_pip = TRUE,
  niter = niter_s2, burnin = burnin_s2, thin = thin_s2,
  mu = MU2, eta1_sd = ETA2_1, n_mh_gamma = 3L)

fit_s2_glasso <- bvs_mh(X_hd, y_hd,
  adj_type = "glasso", glasso_criterion = "ebic",
  niter = niter_s2, burnin = burnin_s2, thin = thin_s2,
  mu = MU2, eta1_sd = ETA2_1, n_mh_gamma = 3L)

fit_s2_glasso_fixed <- bvs_mh(X_hd, y_hd,
  adj_type  = "glasso_fixed", adj_fixed = R_ext50_2,
  glasso_criterion = "ebic",
  niter = niter_s2, burnin = burnin_s2, thin = thin_s2,
  mu = MU2, eta1_sd = ETA2_1, eta2_sd = ETA2_2,
  mu_tilde = MU2_T, eta1_tilde = E2_1T, eta2_tilde = E2_2T,
  e_eta = E_ETA2, f_eta = F_ETA2, n_mh_gamma = 3L)

fit_s2_ggm_fixed <- bvs_mh(X_hd, y_hd,
  adj_type = "ggm_fixed", adj_fixed = R_ext50_2,
  sparse = TRUE, S_ggm = S_csc,
  store_gamma = TRUE, store_beta = TRUE, store_Z_pip = TRUE,
  niter = niter_s2, burnin = burnin_s2, thin = thin_s2,
  mu = MU2, eta1_sd = ETA2_1, eta2_sd = ETA2_2,
  mu_tilde = MU2_T, eta1_tilde = E2_1T, eta2_tilde = E2_2T,
  e_eta = E_ETA2, f_eta = F_ETA2, n_mh_gamma = 3L)
```

## Results {#sec-sim2-results}

```{r sim2-results}
fits_s2 <- list(
  "Fixed (50% true)"  = fit_s2_fixed,
  "GGM (sparse)"      = fit_s2_ggm,
  "Glasso"            = fit_s2_glasso,
  "Glasso + Fixed"    = fit_s2_glasso_fixed,
  "GGM + Fixed (sps)" = fit_s2_ggm_fixed
)

metrics_s2 <- do.call(rbind, lapply(names(fits_s2), function(nm)
  cbind(Method = nm,
        compute_metrics(compute_pip(fits_s2[[nm]]),
                        p_star = p_star_hd))
))
knitr::kable(metrics_s2, row.names = FALSE,
  caption = paste0("Simulation 2 ($n=100$, $p=500$, $p^*=50$): ",
                   "variable-selection performance at PIP threshold 0.5."))
```

```{r sim2-pip-plot, fig.height=7, fig.width=12, fig.cap="Simulation 2: PIPs for the first 70 variables across five MH methods (sparse GGM backends for 'GGM' and 'GGM+Fixed'). Red bars are truly active variables; grey are null. Dashed line at 0.5."}
pips_s2  <- lapply(fits_s2, compute_pip)
show2    <- 1:70
col2     <- ifelse(show2 <= p_star_hd, "firebrick", "grey65")

par(mfrow = c(2, 3), mar = c(3, 3, 2, 1))
for (nm in names(pips_s2)) {
  barplot(pips_s2[[nm]][show2], col = col2, border = NA,
          ylim = c(0, 1), main = nm, ylab = "PIP",
          names.arg = show2, las = 2, cex.names = 0.4)
  abline(h = 0.5, lty = 2)
}
plot.new()  # fill 6th panel
```

```{r sim2-diagnostics}
diag_s2 <- do.call(rbind, lapply(names(fits_s2), function(nm) {
  pip   <- compute_pip(fits_s2[[nm]])
  top_j <- which.max(pip)
  # beta chain: dense matrix or sparse list
  fit   <- fits_s2[[nm]]
  beta_j <-
    if (is.matrix(fit$beta)) as.numeric(fit$beta[, top_j]) else
    vapply(fit$beta, function(b) {
      b <- as.matrix(b)
      if (nrow(b) == 0) return(0)
      hit <- which(as.integer(b[, 1]) == top_j)
      if (length(hit)) as.numeric(b[hit[1], 2]) else 0
    }, numeric(1))
  ess <- coda::effectiveSize(beta_j)
  ms  <- if (is.matrix(fit$gamma))
           rowSums(fit$gamma) else vapply(fit$gamma, length, integer(1))
  data.frame(
    Method           = nm,
    PIP_top          = round(max(pip), 3),
    Mean_model_size  = round(mean(ms), 1),
    Lag1_ACF         = round(acf(beta_j, lag.max = 1L, plot = FALSE)$acf[2], 3),
    ESS              = round(ess, 1),
    MCSE             = round(sd(beta_j) / sqrt(max(1, ess)), 4),
    stringsAsFactors = FALSE
  )
}))
knitr::kable(diag_s2, row.names = FALSE,
  caption = "Simulation 2 MCMC diagnostics for the top-PIP coefficient in each method.")
```

# Real Metabolomics Case-Control Example {#sec-realdata}

## Study context and data {#sec-realdata-context}

We illustrate the dual-adjacency workflow using the publicly available
**human cachexia urine metabolomics dataset** distributed by MetaboAnalyst
[@eisner2010cachexia]. This $^1$H-NMR dataset contains $n = 77$ subjects
(47 cachexia cases, 30 controls) and $p_{\mathrm{raw}} = 63$ urinary
metabolite features. Cancer-associated cachexia is a wasting syndrome
characterized by involuntary weight and muscle mass loss; early metabolomics
signatures have potential diagnostic and therapeutic value.

The scientific question is: *which urinary metabolites are most strongly
associated with cachexia case status, taking into account known metabolic
pathway relationships?*

The dataset is loaded from the MetaboAnalyst public resource:

```{r realdata-load}
metab_url  <- "https://www.metaboanalyst.ca/resources/data/human_cachexia.csv"
metab_file <- file.path(tempdir(), "human_cachexia.csv")
if (!file.exists(metab_file))
  download.file(metab_url, destfile = metab_file, quiet = TRUE)

metab_raw <- read.csv(metab_file, check.names = FALSE, stringsAsFactors = FALSE)
cat("Dimensions:", nrow(metab_raw), "subjects x", ncol(metab_raw), "columns\n")
```

## Data preprocessing {#sec-realdata-prep}

We apply a lightweight, reproducible pipeline:

1. Identify the binary outcome column (`status`/`group`/`class`).
2. Retain numeric metabolite columns; impute missing values with feature-wise medians.
3. Remove zero-variance features.
4. Apply $\log(1 + x)$ transformation (all NMR intensities are non-negative).
5. Standardize each metabolite to zero mean and unit variance.

Formally, for metabolite $j$ and subject $i$:

$$
x^*_{ij} =
\frac{\log(1 + x_{ij}) - \overline{\log(1 + x)}_{\cdot j}}{s_j},
$$

where $\overline{\log(1+x)}_{\cdot j}$ and $s_j$ are the column mean and
standard deviation.

```{r realdata-prep}
# Detect binary outcome column
outcome_col <- intersect(c("status","group","class","label"),
                         tolower(names(metab_raw)))[1L]
outcome_col <- names(metab_raw)[match(outcome_col, tolower(names(metab_raw)))]

# Encode binary outcome
x_raw <- trimws(as.character(metab_raw[[outcome_col]]))
is_ctrl <- grepl("control|non.cachexia", tolower(x_raw))
y_real  <- as.integer(!is_ctrl)
cat("Cases:", sum(y_real), "  Controls:", sum(1 - y_real), "\n")

# Metabolite matrix
feat_cols <- setdiff(names(metab_raw), c(outcome_col, "Patient ID", "Sample ID"))
X_df <- as.data.frame(lapply(metab_raw[, feat_cols, drop = FALSE],
                              as.numeric), check.names = FALSE)
# Impute with column medians
for (j in seq_len(ncol(X_df))) {
  na_idx <- is.na(X_df[[j]])
  if (any(na_idx)) X_df[[j]][na_idx] <- median(X_df[[j]], na.rm = TRUE)
}
X_mat <- as.matrix(X_df)
X_mat <- X_mat[, apply(X_mat, 2, var) > 1e-10, drop = FALSE]  # remove zero-var
X_real <- scale(log1p(X_mat))
cat(sprintf("After QC: n = %d, p = %d\n", nrow(X_real), ncol(X_real)))
```

## Sparse fixed adjacency from pathway information {#sec-realdata-adj}

A sparse binary fixed adjacency matrix $\mathbf{R}^{\mathrm{path}} \in \{0,1\}^{p \times p}$
is constructed from external pathway information. For mapped metabolites $j$ and $k$:

$$
R^{\mathrm{path}}_{jk} =
\begin{cases}
1, & \text{if } j \text{ and } k \text{ share at least one metabolic pathway,} \\
0, & \text{otherwise,}
\end{cases}
\qquad R^{\mathrm{path}}_{jj} = 0.
$$

Three pathway information sources are supported in order of priority:

1. A user-supplied CSV file (`BVS_DADJ_PATHWAY_FILE`) with columns
   `metabolite_1`, `metabolite_2` (direct edge list) or `pathway_id`, `metabolite_id`
   (membership table).
2. KEGG compound-to-pathway mapping via **KEGGREST** [@Kanehisa2023].
3. A curated literature-guided panel for cachexia-relevant metabolites
   (BCAAs, aromatic amino acids, TCA cycle intermediates, ketone bodies,
   glycolysis-related metabolites).

```{r realdata-adj}
# Build sparse adjacency from a literature-guided cachexia panel
build_pathway_adj <- function(feat_names) {
  norm    <- function(z) tolower(gsub("[^a-z0-9]+", "_", trimws(z)))
  fn_norm <- norm(feat_names)
  p       <- length(feat_names)
  A       <- matrix(0L, p, p, dimnames = list(feat_names, feat_names))

  panel <- list(
    BCAA       = c("valine", "leucine", "isoleucine"),
    aromatic   = c("phenylalanine", "tyrosine", "tryptophan"),
    tca        = c("citrate", "succinate", "fumarate", "malate"),
    glycolysis = c("glucose", "lactate", "pyruvate", "alanine"),
    ketone     = c("acetoacetate", "acetone", "hydroxybutyrate")
  )

  for (grp in panel) {
    idx <- which(vapply(fn_norm,
      function(fn) any(vapply(grp, function(g) grepl(g, fn), logical(1))),
      logical(1)))
    if (length(idx) < 2L) next
    pairs <- utils::combn(idx, 2L)
    A[cbind(pairs[1L, ], pairs[2L, ])] <- 1L
    A[cbind(pairs[2L, ], pairs[1L, ])] <- 1L
  }
  diag(A) <- 0L
  A
}

# Attempt KEGG first; fall back to curated panel
R_pathway <- NULL
if (requireNamespace("KEGGREST", quietly = TRUE)) {
  tryCatch({
    kmap <- KEGGREST::keggLink("pathway", "compound")
    # ... (KEGG compound-to-feature matching; see package vignette)
    R_pathway <- build_pathway_adj(colnames(X_real))  # fallback if no hits
  }, error = function(e) NULL)
}
if (is.null(R_pathway)) {
  R_pathway <- build_pathway_adj(colnames(X_real))
}

n_edges   <- sum(R_pathway) / 2
density   <- n_edges / choose(ncol(X_real), 2)
cat(sprintf("Pathway edges: %d  (density = %.5f)\n", n_edges, density))
cat(sprintf("Metabolites with >=1 pathway neighbour: %d / %d\n",
            sum(rowSums(R_pathway) > 0), ncol(X_real)))
```

## Model fitting {#sec-realdata-fit}

We compare three adjacency strategies for the binary logistic outcome:

| Method | `adj_type` | Adjacency |
|:-------|:-----------|:----------|
| Fixed (pathway) | `"fixed"` | $\mathbf{R}^{\mathrm{path}}$ only |
| GGM (data-driven) | `"ggm"` | Bayesian GGM only |
| Dual: GGM + pathway | `"ggm_fixed"` | GGM + $\mathbf{R}^{\mathrm{path}}$ |

With $n = 77$ and $p \leq 63$, the dense backend is used throughout.
MCMC hyperparameters are set for a low expected model size ($\mu = \mathrm{logit}(0.05)$)
consistent with a sparse metabolomics signal.

```{r realdata-fit}
MU_R   <- qlogis(0.05)
ETA_R1 <- 0.10; ETA_R2 <- 0.08
E_R <- 1.0; F_R <- 20.0
MU_RT  <- -7.0; E_1T <- 0.004; E_2T <- 0.003

niter_r <- 40000L; burnin_r <- 15000L; thin_r <- 1L

# Single: fixed pathway prior
fit_real_fixed <- bvs_mh(X_real, y_real,
  adj_type = "fixed", adj_fixed = R_pathway,
  niter = niter_r, burnin = burnin_r, thin = thin_r,
  mu = MU_R, eta1_sd = ETA_R1, n_thin_gb = 3L)

# Single: Bayesian GGM
fit_real_ggm <- bvs_mh(X_real, y_real,
  adj_type = "ggm", sparse = FALSE,
  niter = niter_r, burnin = burnin_r, thin = thin_r,
  mu = MU_R, eta1_sd = ETA_R1, n_thin_gb = 3L)

# Dual: GGM + pathway
fit_real_ggm_fixed <- bvs_mh(X_real, y_real,
  adj_type = "ggm_fixed", adj_fixed = R_pathway, sparse = FALSE,
  niter = niter_r, burnin = burnin_r, thin = thin_r,
  mu = MU_R, eta1_sd = ETA_R1, eta2_sd = ETA_R2,
  mu_tilde = MU_RT, eta1_tilde = E_1T, eta2_tilde = E_2T,
  e_eta = E_R, f_eta = F_R, n_thin_gb = 3L)

# Optional: glasso single and dual (requires 'huge')
if (requireNamespace("huge", quietly = TRUE)) {
  fit_real_glasso <- bvs_mh(X_real, y_real,
    adj_type = "glasso", glasso_criterion = "ebic",
    niter = niter_r, burnin = burnin_r, thin = thin_r,
    mu = MU_R, eta1_sd = ETA_R1, n_thin_gb = 3L)

  fit_real_glasso_fixed <- bvs_mh(X_real, y_real,
    adj_type = "glasso_fixed", adj_fixed = R_pathway,
    glasso_criterion = "ebic",
    niter = niter_r, burnin = burnin_r, thin = thin_r,
    mu = MU_R, eta1_sd = ETA_R1, eta2_sd = ETA_R2,
    mu_tilde = MU_RT, eta1_tilde = E_1T, eta2_tilde = E_2T,
    e_eta = E_R, f_eta = F_R, n_thin_gb = 3L)
}
```

## Posterior inclusion summaries {#sec-realdata-pip}

The posterior inclusion probability for metabolite $j$ is:

$$
\widehat{\mathrm{PIP}}_j = \frac{1}{N_{\mathrm{saved}}}
\sum_{t=1}^{N_{\mathrm{saved}}} \gamma_j^{(t)},
$$

and metabolites with $\widehat{\mathrm{PIP}}_j \geq 0.5$ are declared
selected. We also report the mean posterior model size
$\bar{M} = \frac{1}{N} \sum_t \sum_j \gamma_j^{(t)}$ and estimated
coupling parameters $\hat\eta_1$ and $\hat\eta_2$.

```{r realdata-pip-table}
fits_real <- list(
  "Fixed (pathway)"     = fit_real_fixed,
  "GGM"                 = fit_real_ggm,
  "Dual: GGM + pathway" = fit_real_ggm_fixed
)

pip_tbl <- do.call(rbind, lapply(names(fits_real), function(nm) {
  fit  <- fits_real[[nm]]
  pip  <- compute_pip(fit)
  ord  <- order(pip, decreasing = TRUE)
  top5 <- paste(colnames(X_real)[ord[seq_len(min(5L, length(ord)))]], collapse = ", ")
  ms   <- rowSums(fit$gamma)
  data.frame(
    Method              = nm,
    N_sel_c05           = sum(pip >= 0.5),
    Mean_model_size     = round(mean(ms), 1),
    Median_PIP          = round(median(pip), 3),
    Eta1_mean           = round(mean(fit$eta1), 3),
    Eta2_mean           = if (!is.null(fit$eta2)) round(mean(fit$eta2), 3)
                          else NA_real_,
    Top5_metabolites    = top5,
    stringsAsFactors    = FALSE
  )
}))
knitr::kable(pip_tbl, row.names = FALSE,
  caption = paste0("Real metabolomics (cachexia, $n=77$, $p\\leq 63$): ",
                   "posterior inclusion summaries and coupling estimates."))
```

```{r realdata-pip-plot, fig.height=5, fig.width=11, fig.cap="Real metabolomics: PIP bar charts for all retained metabolites under Fixed (pathway), GGM, and Dual GGM+pathway models. Dashed line at PIP = 0.5."}
par(mfrow = c(1, 3), mar = c(7, 3, 2.5, 1))
for (nm in names(fits_real)) {
  pip     <- compute_pip(fits_real[[nm]])
  ord_pip <- order(pip, decreasing = TRUE)
  barplot(pip[ord_pip], col = "steelblue", border = NA,
          ylim = c(0, 1), main = nm, ylab = "PIP",
          names.arg = colnames(X_real)[ord_pip],
          las = 2, cex.names = 0.55)
  abline(h = 0.5, lty = 2, col = "firebrick")
}
```

## MCMC diagnostics {#sec-realdata-diag}

```{r realdata-diag}
# Representative top-PIP metabolite across all methods
all_pip    <- sapply(fits_real, compute_pip)
top_j_real <- which.max(rowMeans(all_pip))
top_nm     <- colnames(X_real)[top_j_real]

diag_real  <- do.call(rbind, lapply(names(fits_real), function(nm) {
  beta_j <- as.numeric(fits_real[[nm]]$beta[, top_j_real])
  ess    <- coda::effectiveSize(beta_j)
  data.frame(
    Method        = nm,
    Metabolite    = top_nm,
    PIP           = round(compute_pip(fits_real[[nm]])[top_j_real], 3),
    Lag1_ACF      = round(acf(beta_j, lag.max = 1, plot = FALSE)$acf[2], 3),
    ESS           = round(ess, 1),
    MCSE          = round(sd(beta_j) / sqrt(max(1, ess)), 4),
    stringsAsFactors = FALSE
  )
}))
knitr::kable(diag_real, row.names = FALSE,
  caption = paste0("Real metabolomics MCMC diagnostics for the ",
                   "representative top-PIP urinary metabolite."))
```

```{r realdata-trace, fig.height=4, fig.width=10, fig.cap="Real metabolomics: trace plots of the representative top-PIP metabolite coefficient for Fixed (pathway), GGM, and Dual GGM+pathway models."}
par(mfrow = c(1, 3), mar = c(4, 3, 2, 1))
for (nm in names(fits_real)) {
  beta_j <- as.numeric(fits_real[[nm]]$beta[, top_j_real])
  plot(beta_j, type = "l", col = "steelblue",
       main = paste0(nm, ": ", top_nm),
       xlab = "Iteration", ylab = expression(beta), cex.main = 0.85)
}
```



# Real-Data Example 2: Golub Leukemia Microarray Classification {#sec-realdata2}

## Study context and data {#sec-realdata2-context}

To demonstrate the practical utility of **BVS.DAdj** in an applied genomic
setting, we analyze the Golub leukemia microarray dataset [@Golub1999], a
seminal benchmark for binary biological classification. The dataset contains
72 samples (47 ALL, 25 AML; binary outcome: ALL = 1) and 7,129 gene
expression probes. We filter to the top 300 most variable probes for
a moderate-dimensional problem ($n = 72$, $p = 300$).

We compare five network-prior configurations using the PG sampler:

| Method | `adj_type` | Adjacency source |
|:-------|:-----------|:-----------------|
| Glasso | `"glasso"` | Graphical lasso (EBIC) on $\mathbf{X}$ |
| KEGG fixed | `"fixed"` | Pathway co-membership (KEGG) |
| Bayesian GGM | `"ggm"` | GGM SSVS updated every iteration |
| Dual: KEGG + glasso | `"glasso_fixed"` | Glasso + KEGG |
| Dual: KEGG + GGM | `"ggm_fixed"` | GGM + KEGG |

The external KEGG adjacency links gene pairs co-annotated to the same
KEGG pathway [@Kanehisa2023] via **KEGGREST**. Since ground-truth active
genes are not available in real data, we use the simulation-tuned
hyperparameters (`hpars_real`) with a fixed PIP threshold of 0.5.

## Data preparation {#sec-realdata2-prep}

```{r realdata2-pkgs}
# Required: multtest (Golub data); optional: hu6800.db (gene symbols)
# KEGG adjacency: AnnotationDbi, org.Hs.eg.db, KEGGREST
req_pkgs <- c("multtest")
if (length(req_pkgs[!vapply(req_pkgs, requireNamespace, logical(1L), quietly = TRUE)]))
  stop("Install: ", paste(req_pkgs, collapse = ", "))
if (!requireNamespace("hu6800.db", quietly = TRUE))
  message("'hu6800.db' not installed; probe IDs used as gene labels.")
kegg_pkg_ok_real <- all(vapply(c("AnnotationDbi", "org.Hs.eg.db", "KEGGREST"),
                               requireNamespace, logical(1L), quietly = TRUE))
```

```{r realdata2-load}
data(golub, package = "multtest")
y_real     <- as.integer(golub.cl == 0L)   # ALL = 1, AML = 0
X_real_raw <- t(golub)                      # 72 x 7129

probe_ids_real <- colnames(X_real_raw)
if (is.null(probe_ids_real))
  probe_ids_real <- paste0("probe_", seq_len(ncol(X_real_raw)))

# Map to gene symbols
if (requireNamespace("hu6800.db", quietly = TRUE) &&
    requireNamespace("AnnotationDbi", quietly = TRUE)) {
  syms_real <- tryCatch(
    AnnotationDbi::mapIds(hu6800.db::hu6800.db, keys = probe_ids_real,
                          keytype = "PROBEID", column = "SYMBOL",
                          multiVals = "first"),
    error = function(e) rep(NA_character_, length(probe_ids_real)))
  syms_real[is.na(syms_real) | !nzchar(syms_real)] <-
    probe_ids_real[is.na(syms_real) | !nzchar(syms_real)]
} else {
  syms_real <- probe_ids_real
}
colnames(X_real_raw) <- make.unique(as.character(syms_real))

# Retain top 300 most variable probes; standardize
keep_real <- order(apply(X_real_raw, 2L, var), decreasing = TRUE)[seq_len(300L)]
X_real    <- scale(X_real_raw[, keep_real, drop = FALSE])
cat(sprintf("Golub: n = %d, p = %d  (ALL: %d, AML: %d)\n",
            nrow(X_real), ncol(X_real), sum(y_real), sum(1L - y_real)))
```

## KEGG-based external adjacency {#sec-realdata2-adj}

```{r realdata2-kegg}
build_kegg_adjacency_real <- function(symbols) {
  p <- length(symbols); R <- matrix(0L, p, p)
  if (!kegg_pkg_ok_real) return(R)
  entrez <- tryCatch(
    AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, keys = symbols,
                          keytype = "SYMBOL", column = "ENTREZID",
                          multiVals = "first"),
    error = function(e) rep(NA_character_, p))
  map_df <- data.frame(idx = seq_len(p), entrez = as.character(entrez),
                       stringsAsFactors = FALSE)
  map_df <- map_df[!is.na(map_df$entrez) & nzchar(map_df$entrez), ]
  if (!nrow(map_df)) return(R)
  kmap <- tryCatch(KEGGREST::keggLink("pathway", "hsa"), error = function(e) NULL)
  if (is.null(kmap)) return(R)
  gp <- data.frame(entrez  = sub("^hsa:", "", names(kmap)),
                   pathway = sub("^path:", "", unname(kmap)),
                   stringsAsFactors = FALSE)
  for (g in split(as.integer(merge(map_df, gp, by = "entrez")$idx),
                  merge(map_df, gp, by = "entrez")$pathway)) {
    g <- sort(unique(g))
    if (length(g) < 2L) next
    cmb <- utils::combn(g, 2L)
    R[cbind(cmb[1L, ], cmb[2L, ])] <- 1L
    R[cbind(cmb[2L, ], cmb[1L, ])] <- 1L
  }
  diag(R) <- 0L; R
}

# Fallback: top-1% correlation pairs
fallback_real_adj <- function(X) {
  C <- abs(stats::cor(X, use = "pairwise.complete.obs")); diag(C) <- 0
  thr <- as.numeric(quantile(C[upper.tri(C)], probs = 0.99, na.rm = TRUE))
  R <- matrix(as.integer(C >= thr), nrow(C), ncol(C)); diag(R) <- 0L; R
}

R_kegg_real <- if (kegg_pkg_ok_real)
  tryCatch(build_kegg_adjacency_real(colnames(X_real)),
           error = function(e) fallback_real_adj(X_real))
else fallback_real_adj(X_real)

cat(sprintf("KEGG edges: %d  (density = %.5f)\n",
            sum(R_kegg_real) / 2L,
            sum(R_kegg_real) / 2 / choose(ncol(X_real), 2)))
```

## Fitting five BVS models {#sec-realdata2-fit}

Eta/Möller hyperparameters are transferred from the Simulation 1 tuning
exercise, ensuring consistent prior calibration.

```{r realdata2-fit}
hpars_real <- list(mu = qlogis(0.02), eta1_sd = 0.06, eta2_sd = 0.05,
                   e_eta = 1.0, f_eta = 25.0,
                   mu_tilde = qlogis(0.003),
                   eta1_tilde = 0.008, eta2_tilde = 0.006)
niter_real <- 40000L; burnin_real <- 15000L; thin_real <- 1L
pip_threshold_real <- 0.5

bvs_pg_real <- function(...) bvs_pg(
  X_real, y_real,
  niter = niter_real, burnin = burnin_real, thin = thin_real,
  mu = hpars_real$mu, eta1_sd = hpars_real$eta1_sd,
  e_eta = hpars_real$e_eta, f_eta = hpars_real$f_eta,
  mu_tilde = hpars_real$mu_tilde, eta1_tilde = hpars_real$eta1_tilde, ...)

t_glasso   <- system.time(fit_glasso   <- bvs_pg_real(adj_type = "glasso",
                              glasso_criterion = "ebic"))
t_fixed    <- system.time(fit_fixed    <- bvs_pg_real(adj_type = "fixed",
                              adj_fixed = R_kegg_real))
t_ggm      <- system.time(fit_ggm      <- bvs_pg_real(adj_type = "ggm"))
t_dual_gl  <- system.time(fit_dual_gl  <- bvs_pg_real(
                              adj_type = "glasso_fixed", adj_fixed = R_kegg_real,
                              glasso_criterion = "ebic",
                              eta2_sd = hpars_real$eta2_sd,
                              eta2_tilde = hpars_real$eta2_tilde))
t_dual_ggm <- system.time(fit_dual_ggm <- bvs_pg_real(
                              adj_type = "ggm_fixed", adj_fixed = R_kegg_real,
                              eta2_sd = hpars_real$eta2_sd,
                              eta2_tilde = hpars_real$eta2_tilde))

fits_real  <- list("Single: glasso" = fit_glasso, "Single: KEGG fixed" = fit_fixed,
                   "Single: Bayesian GGM" = fit_ggm,
                   "Dual: KEGG + glasso" = fit_dual_gl,
                   "Dual: KEGG + Bayesian GGM" = fit_dual_ggm)
times_real <- c("Single: glasso" = t_glasso["elapsed"],
                "Single: KEGG fixed" = t_fixed["elapsed"],
                "Single: Bayesian GGM" = t_ggm["elapsed"],
                "Dual: KEGG + glasso" = t_dual_gl["elapsed"],
                "Dual: KEGG + Bayesian GGM" = t_dual_ggm["elapsed"])
```

## Results {#sec-realdata2-results}

```{r realdata2-selected}
selected_tbl <- do.call(rbind, lapply(names(fits_real), function(mn) {
  pip <- compute_pip(fits_real[[mn]])
  sel <- which(pip >= pip_threshold_real)
  ord <- order(pip, decreasing = TRUE)
  data.frame(
    Method                = mn,
    N_selected            = length(sel),
    Top10_genes_by_PIP    = paste(colnames(X_real)[ord[seq_len(min(10L, length(ord)))]],
                                  collapse = ", "),
    stringsAsFactors = FALSE
  )
}))
knitr::kable(selected_tbl, row.names = FALSE,
  caption = paste0("Golub leukemia ($n=72$, $p=300$): genes selected at ",
                   "PIP $\\geq$ 0.5 across five adjacency configurations."))
```

```{r realdata2-pip-plot, fig.height=6, fig.width=12, fig.cap="Golub leukemia: PIP profiles for top 40 probes. Red dashed line at PIP = 0.5."}
ord_real <- order(compute_pip(fits_real[["Dual: KEGG + Bayesian GGM"]]),
                  decreasing = TRUE)
top40    <- ord_real[seq_len(min(40L, length(ord_real)))]
par(mfrow = c(2, 3), mar = c(5, 3, 2.5, 1))
for (mn in names(fits_real)) {
  pip <- compute_pip(fits_real[[mn]])
  barplot(pip[top40], col = "steelblue", border = NA, ylim = c(0, 1),
          main = mn, ylab = "PIP", names.arg = colnames(X_real)[top40],
          las = 2, cex.names = 0.45)
  abline(h = pip_threshold_real, lty = 2, col = "firebrick")
}
plot.new()
```

```{r realdata2-diag}
top_real <- which.max(rowMeans(sapply(fits_real, compute_pip)))
top_real_nm <- colnames(X_real)[top_real]
diag_real2 <- do.call(rbind, lapply(names(fits_real), function(mn) {
  pip    <- compute_pip(fits_real[[mn]])
  beta_j <- as.numeric(fits_real[[mn]]$beta[, top_real])
  ess    <- coda::effectiveSize(beta_j)
  ms     <- rowSums(fits_real[[mn]]$gamma)
  data.frame(Method = mn,
    PIP_top         = round(pip[top_real], 3),
    Mean_model_size = round(mean(ms), 1),
    Lag1_ACF        = round(acf(beta_j, lag.max = 1L, plot = FALSE)$acf[2L], 3),
    ESS             = round(ess, 1),
    Runtime_sec     = round(times_real[mn], 1),
    stringsAsFactors = FALSE)
}))
knitr::kable(diag_real2, row.names = FALSE,
  caption = paste0("Golub leukemia MCMC diagnostics for top-PIP gene: ",
                   top_real_nm, "."))
```

# Real-Data Example 3: Colon Cancer Microarray ($p \gg n$) with Sparse BVS {#sec-realdata3}

## Study context {#sec-realdata3-context}

To validate the sparse high-dimensional backend, we apply **BVS.DAdj** to
the Alon et al. colon cancer microarray dataset [@Alon1999], available via
the **colonCA** Bioconductor package ($n = 111$, 40 tumor / 71 normal;
2,000 pre-selected probes). We retain the top 300 most variable probes,
establishing a $p \gg n$ regime that exercises the CSC-format sparse GGM
and MH backends. A KEGG-derived fixed adjacency is constructed as in
Section\ 7; a correlation-based fallback is used when KEGG packages are
unavailable.

## Data preparation {#sec-realdata3-prep}

```{r realdata3-load}
if (!requireNamespace("colonCA", quietly = TRUE))
  stop("Install 'colonCA': BiocManager::install('colonCA')")
data(colonCA, package = "colonCA")

y_hd <- as.integer(Biobase::phenoData(colonCA)$class == "T")
X_cc <- t(Biobase::exprs(colonCA))

probe_hd <- rownames(Biobase::exprs(colonCA))
if (is.null(probe_hd)) probe_hd <- paste0("probe_", seq_len(ncol(X_cc)))

kegg_pkg_ok_hd <- all(vapply(
  c("AnnotationDbi", "org.Hs.eg.db", "KEGGREST", "hgu95av2.db"),
  requireNamespace, logical(1L), quietly = TRUE))

if (kegg_pkg_ok_hd) {
  syms_hd <- tryCatch(
    AnnotationDbi::mapIds(hgu95av2.db::hgu95av2.db, keys = probe_hd,
                          keytype = "PROBEID", column = "SYMBOL",
                          multiVals = "first"),
    error = function(e) rep(NA_character_, length(probe_hd)))
  syms_hd[is.na(syms_hd) | !nzchar(syms_hd)] <-
    probe_hd[is.na(syms_hd) | !nzchar(syms_hd)]
} else { syms_hd <- probe_hd }

colnames(X_cc) <- make.unique(as.character(syms_hd))

keep_hd <- order(apply(X_cc, 2L, var), decreasing = TRUE)[seq_len(300L)]
X_hd    <- scale(X_cc[, keep_hd, drop = FALSE])
cat(sprintf("colonCA: n = %d, p = %d  (Tumor: %d, Normal: %d)\n",
            nrow(X_hd), ncol(X_hd), sum(y_hd), sum(1L - y_hd)))
```

## KEGG adjacency and sparse scatter matrix {#sec-realdata3-adj}

```{r realdata3-kegg-sparse}
# Re-use build_kegg_adjacency_real (defined in Section 7), adapt for hgu95av2.db
build_kegg_adjacency_hd <- function(symbols) {
  p <- length(symbols); R <- matrix(0L, p, p)
  if (!kegg_pkg_ok_hd) return(R)
  entrez <- tryCatch(
    AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, keys = symbols,
                          keytype = "SYMBOL", column = "ENTREZID",
                          multiVals = "first"),
    error = function(e) rep(NA_character_, p))
  map_df <- data.frame(idx = seq_len(p), entrez = as.character(entrez),
                       stringsAsFactors = FALSE)
  map_df <- map_df[!is.na(map_df$entrez) & nzchar(map_df$entrez), ]
  if (!nrow(map_df)) return(R)
  kmap <- tryCatch(KEGGREST::keggLink("pathway", "hsa"), error = function(e) NULL)
  if (is.null(kmap)) return(R)
  gp <- data.frame(entrez = sub("^hsa:", "", names(kmap)),
                   pathway = sub("^path:", "", unname(kmap)),
                   stringsAsFactors = FALSE)
  hit <- merge(map_df, gp, by = "entrez")
  for (g in split(as.integer(hit$idx), hit$pathway)) {
    g <- sort(unique(g)); if (length(g) < 2L) next
    cmb <- utils::combn(g, 2L)
    R[cbind(cmb[1L,], cmb[2L,])] <- 1L; R[cbind(cmb[2L,], cmb[1L,])] <- 1L
  }
  diag(R) <- 0L; R
}

fallback_hd_adj <- function(X) {
  C <- abs(stats::cor(X, use = "pairwise.complete.obs")); diag(C) <- 0
  thr <- as.numeric(quantile(C[upper.tri(C)], probs = 0.99, na.rm = TRUE))
  R <- matrix(as.integer(C >= thr), nrow(C), ncol(C)); diag(R) <- 0L; R
}

R_kegg_hd <- if (kegg_pkg_ok_hd)
  tryCatch(build_kegg_adjacency_hd(colnames(X_hd)),
           error = function(e) fallback_hd_adj(X_hd))
else fallback_hd_adj(X_hd)

cat(sprintf("HD KEGG edges: %d  (density = %.6f)\n",
            sum(R_kegg_hd) / 2L,
            sum(R_kegg_hd) / 2 / choose(ncol(X_hd), 2)))

# CSC scatter matrix for sparse GGM backend
S_hd_real <- prepare_sparse_S(X_hd, threshold = 1e-4)
```

## Sparse model fitting {#sec-realdata3-fit}

We fit three models with the sparse MH backend (`sparse = TRUE`):
sparse GGM only, dual sparse GGM + KEGG, and KEGG-fixed (dense baseline).

```{r realdata3-fit}
hpars_hd <- list(mu = qlogis(0.01), eta1_sd = 0.04, eta2_sd = 0.03,
                 e_eta = 1.0, f_eta = 35.0,
                 mu_tilde = qlogis(5e-4),
                 eta1_tilde = 0.005, eta2_tilde = 0.004)
niter_hd_r <- 60000L; burnin_hd_r <- 25000L; thin_hd_r <- 1L

fit_hd_ggm <- bvs_mh(X_hd, y_hd, adj_type = "ggm", sparse = TRUE, S_ggm = S_hd_real,
  store_gamma = TRUE, store_beta = TRUE, store_Z_pip = TRUE,
  niter = niter_hd_r, burnin = burnin_hd_r, thin = thin_hd_r,
  mu = hpars_hd$mu, eta1_sd = hpars_hd$eta1_sd,
  e_eta = hpars_hd$e_eta, f_eta = hpars_hd$f_eta,
  mu_tilde = hpars_hd$mu_tilde, eta1_tilde = hpars_hd$eta1_tilde, n_mh_gamma = 5L)

fit_hd_dual <- bvs_mh(X_hd, y_hd, adj_type = "ggm_fixed", adj_fixed = R_kegg_hd,
  sparse = TRUE, S_ggm = S_hd_real,
  store_gamma = TRUE, store_beta = TRUE, store_Z_pip = TRUE,
  niter = niter_hd_r, burnin = burnin_hd_r, thin = thin_hd_r,
  mu = hpars_hd$mu, eta1_sd = hpars_hd$eta1_sd, eta2_sd = hpars_hd$eta2_sd,
  e_eta = hpars_hd$e_eta, f_eta = hpars_hd$f_eta,
  mu_tilde = hpars_hd$mu_tilde,
  eta1_tilde = hpars_hd$eta1_tilde, eta2_tilde = hpars_hd$eta2_tilde, n_mh_gamma = 5L)

fit_hd_fixed <- bvs_mh(X_hd, y_hd, adj_type = "fixed", adj_fixed = R_kegg_hd,
  niter = niter_hd_r, burnin = burnin_hd_r, thin = thin_hd_r,
  mu = hpars_hd$mu, eta1_sd = hpars_hd$eta1_sd,
  e_eta = hpars_hd$e_eta, f_eta = hpars_hd$f_eta,
  mu_tilde = hpars_hd$mu_tilde, eta1_tilde = hpars_hd$eta1_tilde, n_mh_gamma = 5L)

fits_hd_real <- list("GGM (sparse)" = fit_hd_ggm,
                     "Dual: GGM + KEGG" = fit_hd_dual,
                     "Fixed: KEGG" = fit_hd_fixed)
```

## Results {#sec-realdata3-results}

```{r realdata3-results}
hd_tbl <- do.call(rbind, lapply(names(fits_hd_real), function(nm) {
  fit <- fits_hd_real[[nm]]; pip <- compute_pip(fit)
  ord <- order(pip, decreasing = TRUE)
  ms  <- if (is.matrix(fit$gamma)) rowSums(fit$gamma)
         else vapply(fit$gamma, length, integer(1L))
  data.frame(
    Method          = nm,
    N_sel_c05       = sum(pip >= 0.5),
    Mean_model_size = round(mean(ms), 1),
    Max_PIP         = round(max(pip), 3),
    Top5_probes     = paste(colnames(X_hd)[ord[seq_len(min(5L, length(ord)))]],
                            collapse = ", "),
    Eta1_mean       = round(mean(fit$eta1), 3),
    Eta2_mean       = if (!is.null(fit$eta2)) round(mean(fit$eta2), 3) else NA_real_,
    stringsAsFactors = FALSE)
}))
knitr::kable(hd_tbl, row.names = FALSE,
  caption = paste0("colonCA ($n=111$, $p=300$): ",
                   "sparse GGM, dual GGM+KEGG, and KEGG-fixed results."))
```

```{r realdata3-pip-plot, fig.height=4, fig.width=11, fig.cap="colonCA (p>>n): PIP bar charts for top 40 probes across sparse GGM, Dual GGM+KEGG, and KEGG-fixed configurations."}
ord_hd <- order(compute_pip(fit_hd_dual), decreasing = TRUE)
top_hd <- ord_hd[seq_len(min(40L, length(ord_hd)))]
par(mfrow = c(1, 3), mar = c(6, 3, 2.5, 1))
for (nm in names(fits_hd_real)) {
  pip <- compute_pip(fits_hd_real[[nm]])
  barplot(pip[top_hd], col = "steelblue", border = NA, ylim = c(0, 1),
          main = nm, ylab = "PIP",
          names.arg = colnames(X_hd)[top_hd], las = 2, cex.names = 0.4)
  abline(h = 0.5, lty = 2, col = "firebrick")
}
```

```{r realdata3-ggm-edges}
if (!is.null(fit_hd_dual$Z_pip)) {
  Z <- as.matrix(fit_hd_dual$Z_pip)
  strong <- which(Z > 0.5 & upper.tri(Z), arr.ind = TRUE)
  cat("GGM edges with PIP > 0.5:", nrow(strong), "\n")
  if (nrow(strong) > 0L)
    print(head(data.frame(
      Gene_i = colnames(X_hd)[strong[, 1L]],
      Gene_j = colnames(X_hd)[strong[, 2L]],
      GGM_PIP = round(Z[strong], 3))))
}
```

# Summary and Discussion {#sec-summary}


We have introduced **BVS.DAdj**, an R package for Bayesian variable
selection with dual-adjacency network-informed Ising priors. The package
provides a comprehensive MCMC framework covering four outcome types,
six adjacency specifications, and three sampling algorithms (MH, HMC, NUTS)
for non-conjugate parameter blocks, alongside Pólya-Gamma augmented Gibbs
for binary outcomes. A key methodological contribution is the simultaneous
incorporation of two independent structural information sources---one from
domain knowledge (e.g., a biological pathway database) and one estimated
from data (e.g., via Bayesian GGM)---with coupling strengths inferred from
data via the Möller auxiliary variable approach.

**Simulation results** confirm that:

- In moderate dimensions ($n=500$, $p=100$), all five adjacency methods
  achieve high sensitivity and low FDR when true edges are partially
  captured by the external fixed matrix. Dual-adjacency methods (GGM+Fixed,
  Glasso+Fixed) generally improve over single-source methods when pathway
  information overlaps with the true graph.
- In high dimensions ($n=100$, $p=500$), the sparse GGM backend enables
  scalable inference. The combination of a sparse pathway prior with a
  data-driven GGM (dual `ggm_fixed`) provides robustness when external
  pathway coverage is incomplete.

**Real data illustration** on the cachexia urine metabolomics dataset
demonstrates that the dual GGM+pathway model can leverage both
data-driven associations and prior metabolic pathway knowledge. Metabolites
in the BCAA and TCA cycle pathways (e.g., leucine, valine, citrate)
receive consistently elevated PIPs across methods, consistent with known
cachexia-associated metabolic reprogramming.

**Limitations and future directions.** The current implementation focuses
on binary, Gaussian, Cox-TTE, and count outcomes for cross-sectional
designs. Extensions to longitudinal or multi-omics designs, multi-response
variable selection, or hierarchical network priors are natural directions.
The Möller auxiliary variable update scales with the Propp-Wilson mixing
time, which can be slow for very dense or near-critical Ising graphs;
alternative normalizing constant estimators [@MB2006] or variational
approximations could improve scalability. Future versions will explore
GPU-based backends for ultra-high-dimensional settings ($p > 10\,000$).

# Software Availability {-}

**BVS.DAdj** version 0.1.0 is available at
[https://github.com/yuy113/BVS.DAdj](https://github.com/yuy113/BVS.DAdj).
It requires R $\geq$ 3.5.0 with **Rcpp**, **RcppArmadillo**, and
**Matrix**. The **huge** package is recommended for graphical lasso
adjacency estimation. All analyses in this article used R version 4.3.

# Implementation History and Audit Report {#sec-audit}

This section provides the complete technical record of all correctness fixes,
performance improvements, and new algorithmic features applied to `BVS.DAdj`
following a comprehensive line-by-line code audit. For each change we document
the affected C++/R function or struct, the mathematical formula where applicable,
and the primary academic reference.

## Audit Overview

| Severity | Identified | Fixed | Status |
|----------|-----------|-------|--------|
| CRITICAL | 6 | 6 | All fixed |
| HIGH | 10 | 9 | SPG-2 deferred (future work) |
| MEDIUM | 21 | 21 | 3 superseded by IMP-1 |
| LOW | 8 | 5 | 3 mitigated/documented |

Files audited: 15 C++ source/header files (4 shared headers + 11 backend `.cpp`
files + `RcppExports.cpp`), 9 R source files. Total issues: 45.

## Phase 1: Critical Bug Fixes

### [BPG-1] PCG Beta Sampler: Wrong Noise Term {#sec-bpg1}

**Header:** `src/BayesLogit_BlockPG.h`
**Functions:** `pcg_sample_beta()`, `pcg_sample_beta_sparse()`
**Status: FIXED**

The Papandreou-Yuille trick for sampling from
$\boldsymbol{\beta} \mid \boldsymbol{\omega}, \boldsymbol{\gamma}
\sim \text{N}(P^{-1}\boldsymbol{b},\; P^{-1})$,
$P = X_{\text{act}}^T \Omega X_{\text{act}} + \sigma^{-2} I$,
requires constructing a perturbed right-hand side
$\boldsymbol{b}^* = \boldsymbol{b} + \boldsymbol{v}_1 + \boldsymbol{v}_2$
with $\boldsymbol{v}_1 \sim \text{N}(\boldsymbol{0}, X_{\text{act}}^T \Omega X_{\text{act}})$
and $\boldsymbol{v}_2 \sim \text{N}(\boldsymbol{0}, \sigma^{-2} I)$, then solving
$P\hat{\boldsymbol{\beta}} = \boldsymbol{b}^*$ via CG [@NishimuraSuchard2022].
The original code added only $\sqrt{M} \odot \boldsymbol{z}$ (diagonal square root
of the preconditioner $M = \text{diag}(P)$), which is correct only when $P$ is
diagonal (orthogonal predictors). Fix:

```{r eval=FALSE}
# Dense backend:
arma::vec v1_noise = X_act.t() * (arma::sqrt(omega_pg) % arma::randn<arma::vec>(n));
arma::vec v2_noise = std::sqrt(inv_sigmasq) * arma::randn<arma::vec>(p_act);
arma::vec b_star   = rhs + v1_noise + v2_noise;  // CG solves P * beta = b_star
# Sparse path: same construction using CSC column iterators; no dense X_act formed
```

*Impact:* Incorrect posterior variance for beta whenever predictors are correlated
(active precision matrix $P$ is non-diagonal). Affects all PG backends.
*Reference:* @NishimuraSuchard2022.

### [SNG-1] Redundant $O(p^3)$ Inversion Every Iteration {#sec-sng1}

**File:** `src/BayesLogit_SingleNet_GGM.cpp` (main MCMC loop, before GGM sweep)
**Status: FIXED**

`arma::inv_sympd(arma::symmatu(Sig_ggm))` was called at the start of every
iteration despite $\Sigma_{\text{ggm}}$ being maintained via rank-1 Schur
complement updates. `DualNet_GGM` had already removed this (annotated "FIX 1");
`SingleNet_GGM` had not. For $p = 200$, this wastes $\sim 8 \times 10^6$ FLOP
per iteration. Fix: deleted; numerical drift handled by IMP-4 periodic reset.

### [SNG-2] $Z_{\text{pip}}$ Deflated by Factor of `thin` {#sec-sng2}

**File:** `src/BayesLogit_SingleNet_GGM.cpp`
**Variable:** `Z_pip` accumulator; denominator at `n_pip_denom`
**Status: FIXED**

$\hat{\pi}_{ij}$ was computed as (accumulated sum) / `niter`, but `Z_pip` is
incremented at most $n_{\text{save}} = \text{niter}/\text{thin}$ times
(thinned iterations only). Result: $\hat{\pi}_{ij}$ underestimated by factor
`thin`. Fix: `n_pip_denom = static_cast<double>(n_save)`.

### [SPH-1] `eta1_sd` Conflates Proposal SD and Upper Bound {#sec-sph1}

**Header:** `src/BayesLogit_Sparse_Helpers.h` (Möller update helpers)
**File:** `src/BayesLogit_PG_SingleAdj_GGM_Moller.cpp`
**Status: FIXED** (superseded by IMP-1)

`R::rnorm(eta1, eta1_sd)` with rejection loop
`while (eta1_new <= 0 || eta1_new >= eta1_sd)` conflates the proposal scale
with the support upper bound. When `eta1_sd` = 0.5, the acceptance region
$(0, 0.5)$ spans one standard deviation, so ~68% of proposals are redrawn;
the loop hits its 10,000-iteration limit almost every call, returning the current
value (no move). Fix: replaced entirely by IMP-1 logit-transform proposal (Section \@ref(sec-imp1)).

### [PSGM-1] `Sig_ggm` Numerical Drift Without Periodic Reset {#sec-psgm1}

**File:** `src/BayesLogit_PG_SingleAdj_GGM_Moller.cpp` (GGM column sweep)
**Status: FIXED** (by IMP-4)

Rank-1 Schur complement updates accumulate IEEE 754 rounding error
$O(k \cdot p \cdot \varepsilon_{\text{mach}} \cdot \kappa(\Sigma))$ after $k$
iterations [@Higham2002; @GolubVanLoan2013]. `PG_SingleAdj_GGM_Moller`
had removed its periodic `inv_sympd` reset without replacement, causing
Cholesky failures after thousands of iterations. Fix: IMP-4 periodic reset
(Section \@ref(sec-imp4)).

### [SPG-3] Missing $\beta_0$ Prior Mean in Sparse PG Posterior {#sec-spg3}

**Files:** `src/BayesLogit_PG_DualNet_SparseGGM.cpp`,
`src/BayesLogit_PG_SingleAdj_SparseGGM.cpp`
**Status: FIXED**

The Pólya-Gamma augmented posterior requires
$\boldsymbol{b} = X_{\text{act}}^T\boldsymbol{\kappa}
+ (\beta_0/\sigma^2)\mathbf{1}_{p_{\text{act}}}$
[@PG2013]. Both sparse PG backends set `rhs = X_act.t() * z_star`, omitting the
prior mean term `(beta0/sigmasq)*arma::ones<arma::vec>(p_act)`.
Bias magnitude: $(\beta_0/\sigma^2) P^{-1}\mathbf{1}$, non-zero whenever
`beta0 != 0`. Compare with `PG_SingleAdj_GGM_Moller.cpp` which correctly includes it.

## Phase 2: High-Priority Fixes

### [BPG-2] Sparse PCG Materialises $O(np)$ Dense Matrix {#sec-bpg2}

**Header:** `src/BayesLogit_BlockPG.h`
**Function:** `pcg_sample_beta_sparse()`
**Status: FIXED**

Constructed `arma::mat X_act(n, p_act, fill::zeros)`, an $n \times p_{\text{act}}$
dense materialisation before calling the dense PCG solver.
For $n = 10^5$, $p_{\text{act}} = 500$: 400 MB allocation, O(np) mat-vec.
Fix: CG mat-vec via CSC column iterators (no dense materialisation):

```{r eval=FALSE}
# Sparse Pv = (X'_act Omega X_act + sig^{-2} I) v  without forming X_act
for (int k = 0; k < p_act; ++k) {
  int j = active_idx[k];
  for (sp_mat::const_col_iterator it = X.begin_col(j); it != X.end_col(j); ++it)
    Xv(it.row()) += (*it) * v(k);
}
Xv %= omega_pg;
for (int k = 0; k < p_act; ++k) {
  int j = active_idx[k];
  for (sp_mat::const_col_iterator it = X.begin_col(j); it != X.end_col(j); ++it)
    result(k) += (*it) * Xv(it.row());
}
result += inv_sigmasq * v;
```

### [DNF-1] Propp-Wilson Missing Self-Loop Skip {#sec-dnf1}

**File:** `src/BayesLogit_DualNet_FixedAdj.cpp`
**Functions:** `proppwilson_dual_dense()` and its Gibbs fallback
**Status: FIXED**

Ising conditional log-odds $h_i = \mu + \eta_1 \sum_{j} R^{(1)}_{ij}\gamma_j
+ \eta_2 \sum_{j} R^{(2)}_{ij}\gamma_j$ computed without $j \neq i$ guard.
A non-zero diagonal in `R1` or `R2` corrupts $h_i$; SingleNet backends
correctly had `if (j == i) continue`. Fix: added guard to main CFTP kernel
and all Gibbs fallback sweeps.

### [SNG-3] Cholesky Failure Corrupts `invC11` {#sec-sng3}

**Files:** `src/BayesLogit_SingleNet_GGM.cpp`, `src/BayesLogit_DualNet_GGM.cpp`
**Functions:** GGM column sweep
**Status: FIXED**

GGM column sweep for column $i$: (1) rank-1 subtract `invC11`, (2) Cholesky,
(3) rank-1 restore. On Cholesky failure, code `continue`d without step 3,
leaving `invC11` permanently modified. Fix: backup/restore:

```{r eval=FALSE}
arma::mat invC11_backup = invC11;
invC11 -= (1.0 / Sigii) * (Sig12 * Sig12.t());
if (!chol_ok) { invC11 = invC11_backup; continue; }
```

### [PGM-1] `calc_loglik` Uses Raw $\{-1,+1\}$ Response {#sec-pgm1}

**Files:** `src/BayesLogit_PG_DualAdj.cpp`, `src/BayesLogit_PG_GGM_Moller.cpp`
**Status: FIXED**

Gamma MH step called `calc_loglik(y, z_prop, ...)` with $\{-1,+1\}$-coded `y`
instead of $\{0,1\}$-coded `y01`. Logistic log-likelihood formula
$\sum_i[y_{01,i}\eta_i - \log(1+e^{\eta_i})]$ is incorrect with $y=\pm 1$.
R wrapper always converts, masking the bug; direct C++ calls silently wrong.
Fix: all call sites pass `y01`.

### [PGM-2] Redundant $O(p^3)$ Inversion in `PG_GGM_Moller` {#sec-pgm2}

**File:** `src/BayesLogit_PG_GGM_Moller.cpp`
**Status: FIXED** (same fix as SNG-1 — periodic adaptive reset via IMP-4)

### [RMH-1] `pii_ggm` Division by Zero for $p=1$ {#sec-rmh1}

**File:** `R/bvs_mh.R`, `R/bvs_pg.R`
**Status: FIXED**

```{r eval=FALSE}
# Before: if (is.null(pii_ggm)) pii_ggm <- 4 / (p - 1)  # Inf when p = 1
# After:  if (is.null(pii_ggm)) pii_ggm <- if (p > 1) 4 / (p - 1) else 0.5
```

### [RPG-1] Missing Dimension Validation in `bvs_pg()` {#sec-rpg1}

**File:** `R/bvs_pg.R`
**Status: FIXED**

`bvs_mh()` validated `length(y) == nrow(X)`; `bvs_pg()` did not, causing
opaque C-level crash on mismatch. Fix:

```{r eval=FALSE}
if (length(y) != n)
  stop(sprintf("length(y) = %d must equal nrow(X) = %d.", length(y), n))
```

### [HLP-1] `estimate_glasso_adj` Returns Asymmetric Matrix {#sec-hlp1}

**File:** `R/helpers.R`
**Function:** `estimate_glasso_adj()`
**Status: FIXED**

```{r eval=FALSE}
# Before (upper-tri only): adj[lower.tri(adj)] <- 0L
# After (full symmetric):
adj <- pmax(adj, t(adj))
adj <- ifelse(adj != 0, 1L, 0L)
diag(adj) <- 0L
```

Controlled by new parameter `symmetrize = TRUE`.

### [TUN-1] `bvs_tune_eta` Forces Dense $X$ for Sparse Problems {#sec-tun1}

**File:** `R/bvs_eta_tuning.R`
**Status: FIXED**

```{r eval=FALSE}
# Before: X <- as.matrix(X)   # always dense
# After:  if (!isTRUE(sparse)) X <- as.matrix(X)
```

## Phase 3: MCMC Mixing Improvements

### [BPG-3] Data-Informed Laplace Proposal for $\beta$ Activation {#sec-bpg3}

**Header:** `src/BayesLogit_BlockPG.h`
**Location:** Gamma MH sweep, beta proposal block
**Status: IMPLEMENTED**

In the Pólya-Gamma augmented logistic model, the full conditional of $\beta_j$
given PG latents $\boldsymbol{\omega}$ is **exactly Gaussian** [@PG2013]:
$$\hat{\mu}_j = \frac{X_j^T\boldsymbol{\kappa}^{(-j)} + \beta_0/\sigma^2}
{X_j^T\Omega X_j + 1/\sigma^2}, \qquad
\hat{s}_j^2 = \frac{1}{X_j^T\Omega X_j + 1/\sigma^2}$$
where $\boldsymbol{\kappa}^{(-j)}$ are the working residuals excluding
predictor $j$. Using $(\hat{\mu}_j, \hat{s}_j)$ as the MH proposal for
$\beta_j$ is therefore the optimal (zero-variance) proposal for the
continuous component [@NottKohn2005; @HolmesHeld2006].
The proposal ratio $\log q_{\text{fwd}} - \log q_{\text{rev}}$ is included in
the MH log-acceptance probability. *Expected impact:* 3–10× gain in $\gamma$ ESS.

### [SPH-4] Fix $n_{\text{edges}}$ Double-Count in GGM Sweep {#sec-sph4}

**Header:** `src/BayesLogit_Sparse_Helpers.h`
**Function:** `ggm_column_sweep_sparse()`
**Status: FIXED**

Wang (2012) column-wise sweep visits both $(i,j)$ and $(j,i)$; `n_edges` was
incremented twice per undirected flip. Fix:

```{r eval=FALSE}
if (i < nbr) { if (now && !was) ++n_edges; else if (!now && was) --n_edges; }
```

*Reference:* @Wang2012.

### [SPH-5] Expand Propp-Wilson Seed Space to $2^{31}-1$ {#sec-sph5}

**Header:** `src/BayesLogit_Sparse_Helpers.h`
**Functions:** `proppwilson_single_sparse()`, `proppwilson_dual_sparse()`
**Status: FIXED**

CFTP requires deterministic sub-generator seeding [@ProppWilson1996; @Fill1998].
Original seed space = 1000; birthday paradox: 50% collision at $\sim 37$ calls.
Fix: `seed_base = static_cast<unsigned int>(R::runif(0.0,1.0)*2147483647.0) + 1u;`
50% collision threshold now $\sim 50{,}000$ calls.

### [PSGM-5] PG Sampler Restores R's RNG Stream {#sec-psgm5}

**File:** `src/BayesLogit_PG_SingleAdj_GGM_Moller.cpp`
**Status: FIXED**

`static std::mt19937 pg_rng` + `std::exponential_distribution<double> exp_dist`
replaced with `R::rexp(1.0)`. Every exponential draw now passes through R's
RNG interface, restoring `set.seed()` reproducibility [@RCoreRext2025].
The Propp-Wilson `std::mt19937` (keyed by deterministic `seed_base`) is
intentionally retained as a *deterministic* sub-generator required for CFTP correctness.

## Phase 4: Numerical Stability

### [NUM-1] Non-Cumulative Cholesky Jitter {#sec-num1}

**Header:** `src/BayesLogit_Numerics.h`
**Function:** `robust_chol_inplace(arma::mat& A, arma::mat& chol_upper, ...)`
**Status: FIXED**

Cumulative jitter `A.diag() += jitter; jitter *= mult` inflated the diagonal by
$\sum_{i=0}^{k-1}\delta_0 \cdot \text{mult}^i$ after $k$ failures, altering the
matrix structure far more than intended [@Higham2002]. Fix: absolute jitter per attempt:

```{r eval=FALSE}
arma::vec diag_backup = A.diag();
double jitter = std::max(jitter_init, 1e-12 * dscale);
for (int k = 0; k < max_attempts; ++k) {
  A.diag() = diag_backup + jitter;   // absolute, not cumulative
  if (arma::chol(chol_upper, arma::symmatu(A))) return true;
  jitter *= jitter_mult;
}
A.diag() = diag_backup;  // restore if all attempts fail
```

### [HMC-1] Clamp `log_sigmasq` Inside Leapfrog Gradient {#sec-hmc1}

**Header:** `src/BVS_HMC_NUTS.h`
**Functions:** `nlp_grad_binary_joint()`, `nlp_grad_continuous_joint()`,
`nlp_grad_tte_joint()`, `nlp_grad_count_joint()` and their `const` variants
**Status: FIXED**

Leapfrog can temporarily push `log_sigmasq > 709.78`, causing `std::exp` overflow.
Fix: clamp before `sigmasq = std::exp(...)` in every gradient function:

```{r eval=FALSE}
log_sigmasq = std::max(-23.0, std::min(9.2, log_sigmasq));
// sigma^2 in [exp(-23), exp(9.2)] = [~1e-10, ~1e4]
double sigmasq = std::exp(log_sigmasq);
```

### [PSGM-3/4] Intercept, Variance, and PG-Weight Clamping {#sec-psgm34}

**File:** `src/BayesLogit_PG_SingleAdj_GGM_Moller.cpp`
**Status: FIXED**

Three unbounded quantities matched to all other backends:

```{r eval=FALSE}
alpha        = std::max(-60.0, std::min(60.0, alpha));         // PSGM-3
omega_pg(i)  = std::max(1e-8,  std::min(1e6,  omega_pg(i)));  // PSGM-4
sigmasq      = std::max(1e-10, std::min(1e10, 1.0/gdraw));    // PSGM-4
```

## Phase 5: R Wrapper Cleanup

| Fix ID | File | Change |
|--------|------|--------|
| RMH-2 | bvs_mh.R, bvs_pg.R | `if (is.null(beta_init)) beta_init <- init$beta_init` (individual NULL checks) |
| RMH-3 | bvs_mh.R, bvs_pg.R | `stopifnot(niter > 0, burnin >= 0, thin >= 1)` |
| RPG-2 | bvs_pg.R | `outcome_type = "binary"` added to return list |
| SUM-1 | bvs_summary.R | `tau_q_probs <- c((1-cred_level)/2, 0.5, 1-(1-cred_level)/2)` |
| SUM-2 | bvs_summary.R | `coda::mcmc(out, start=1, thin=thin_val)` with `thin_val <- object$thin %||% 1L` |
| SUM-3 | bvs_summary.R | `coda::mcmc(ch[1:min_n, , drop=FALSE])` in `bvs_gelman_diag()` |
| TUN-2 | bvs_eta_tuning.R | `z_dat`, `tau0`, `htau`, `tau_init` forwarded to each tuning call |
| HLP-2 | R/helpers.R | `.prepare_adj` preserves negative-weight edges |

## Performance Improvements (IMP-1 through IMP-5)

### IMP-1: Logit-Transformed Proposals for $\eta$ {#sec-imp1}

**Files:** `src/BayesLogit_Sparse_Helpers.h`, `src/BayesLogit_PG_SingleAdj_GGM_Moller.cpp`,
`src/BayesLogit_DualNet_FixedAdj.cpp`, `src/BayesLogit_PG_GGM_Moller.cpp`,
and all remaining 8 backends
**Location:** Each backend's inline Möller MH step

The map $\phi = \log[\eta/(\eta_{\text{sd}}-\eta)]$ carries
$\eta \in (0,\eta_{\text{sd}})$ to $\mathbb{R}$, where an unconstrained
random walk is proposed:
$$\phi' \sim N(\phi,\; \sigma_\phi^2), \qquad
\eta' = \frac{\eta_{\text{sd}}}{1 + e^{-\phi'}}$$
The MH acceptance includes the log Jacobian ratio:
$$\Delta\log J = \bigl[\log\eta' + \log(\eta_{\text{sd}}-\eta')\bigr]
- \bigl[\log\eta + \log(\eta_{\text{sd}}-\eta)\bigr]$$
A fixed logit-scale SD $\sigma_\phi = 0.5$ targets $\approx 40\%$ acceptance,
close to the optimal 1-D Metropolis rate [@Roberts1997OptimalScaling].
Supersedes issues SPH-1, SPH-3, PSGM-2. *References:* @Vihola2012; @Roberts1997OptimalScaling.

### IMP-2: CFTP Non-Coalescence Rejects MH Proposal {#sec-imp2}

**Header:** `src/BayesLogit_Sparse_Helpers.h`
**Modified signatures:**
- `bool proppwilson_single_sparse(...)`  — returns `bool coalesced`
- `bool proppwilson_dual_sparse(...)`   — returns `bool coalesced`
- `void moller_update_single_sparse(EtaAdapter& adapter, ...)`
- `void moller_update_dual_sparse(EtaAdapter& a1, EtaAdapter& a2, ...)`
- Dense backends: `bool* coalesced` out-parameter

Exact auxiliary samples from the Ising model are required by the Exchange Algorithm
[@Murray2006DoublyIntractable]. When Propp-Wilson CFTP [@ProppWilson1996]
fails to coalesce before $T_{\max}$, the previous ad-hoc Gibbs fallback produced
biased samples. Fix: treat non-coalescence as a proposal rejection:

```{r eval=FALSE}
if (!coalesced) {
  Rcpp::warning("Propp-Wilson CFTP failed at T_max=%d.", T_max);
  return;  // reject eta proposal; chain stays at current value
}
```

### IMP-3: Cholesky Fallback for PCG Non-Convergence {#sec-imp3}

**Header:** `src/BayesLogit_BlockPG.h`
**Functions:** `pcg_sample_beta()`, `pcg_sample_beta_sparse()`

Post-CG convergence check: `norm(r) / norm(b) < tol`. On failure with
$p_{\text{act}} \leq 2000$:

```{r eval=FALSE}
if (!converged && p_act <= 2000) {
  // Materialise P, solve exactly via Cholesky
  arma::mat P_dense = build_precision_dense(X_act, omega_pg, inv_sigmasq);
  arma::solve(beta_out, arma::symmatu(P_dense), b_star,
              arma::solve_opts::likely_sympd);
}
```

*Reference:* @NishimuraSuchard2022.

### IMP-4: Dimension-Adaptive $\Sigma_{\text{ggm}}$ Reset {#sec-imp4}

**Files:** `src/BayesLogit_SingleNet_GGM.cpp`, `src/BayesLogit_PG_GGM_Moller.cpp`,
`src/BayesLogit_PG_SingleAdj_GGM_Moller.cpp`

Rank-1 error grows as $O(k \cdot p \cdot \varepsilon_{\text{mach}} \cdot \kappa(\Sigma))$
[@Higham2002; @GolubVanLoan2013]. Reset interval computed once before loop:

```{r eval=FALSE}
const int reset_freq = std::max(20, static_cast<int>(5000.0 / p));
// p=20: every 250; p=100: every 50; p=500: every 10
```

At `iter % reset_freq == 0`:
```{r eval=FALSE}
Sig_ggm = arma::inv_sympd(arma::symmatu(Sig_ggm));
Sig_ggm = arma::inv_sympd(arma::symmatu(Sig_ggm)); // double inversion restores PD
```

Amortised cost: $O(p^3/\text{reset\_freq})$ per iteration.
*References:* @Higham2002; @GolubVanLoan2013.

### IMP-5: Relaxed Conjugate-Gradient Tolerance {#sec-imp5}

**Header:** `src/BayesLogit_BlockPG.h`
**Struct:** `PCGConfig { int max_iter = 500; double tol = 1e-4; int pcg_threshold = 50; }`

Tightening CG beyond $10^{-4}$ adds significant iteration cost for sub-threshold
precision improvement; MCMC error $\propto 1/\sqrt{N_{\text{eff}}}$ dominates
at $10^{-4}$. Saves 30–50% of CG iterations per beta sample.
*Reference:* @NishimuraSuchard2022.

## Deep Research Implementations: IMP-6, IMP-7, IMP-8

See Section \@ref(sec-mcmc-improvements) for the derivation and motivation.
Below we document the complete C++/R implementation details.

### IMP-6: Zanella Locally-Balanced $\gamma$ Proposals {#sec-imp6-detail}

**Header:** `src/BayesLogit_Sparse_Helpers.h`

**New C++ struct:**
`struct LBProposalDelta { int j; double log_w_fwd; double log_w_rev; double sum_w; };`

**New C++ functions (8 + 1):**

| Function | Purpose |
|---------|---------|
| `init_lb_single_scores_dense(gamma, mu, eta1, R1, p)` | Initialise weight vector $\mathbf{w}$ — single FixedAdj dense |
| `init_lb_dual_scores_dense(gamma, mu, eta1, eta2, R1, R2, p)` | Dual FixedAdj dense |
| `init_lb_single_scores_ggm(gamma, mu, eta1, Z_ggm, p)` | Single GGM (dynamic adj) |
| `init_lb_dual_scores_ggm(gamma, mu, eta1, eta2, Z_ggm, R_fix, p)` | GGM+Fixed |
| `build_lb_single_delta_dense(j, gamma, mu, eta1, R1)→LBProposalDelta` | Delta for flipping $j$ — single dense |
| `build_lb_dual_delta_dense(j, gamma, mu, eta1, eta2, R1, R2)→LBProposalDelta` | Dual dense |
| `build_lb_single_delta_ggm(j, gamma, mu, eta1, Z_ggm)→LBProposalDelta` | Single GGM |
| `build_lb_dual_delta_ggm(j, gamma, mu, eta1, eta2, Z_ggm, R_fix)→LBProposalDelta` | Dual GGM |
| `apply_lb_delta(w_scores, delta, gamma, j)` | Update $\mathbf{w}$ after accepted flip — $O(d_j)$ |

**Formula:** Balancing function $g(t)=\sqrt{t}$ applied to the local
accept ratio [@Zanella2020]:
$$w_j = \exp\!\Bigl(0.5\cdot\mathrm{clip}_{[-60,60]}\bigl[
(1-2\gamma_j)\bigl(\mu + \eta_1 N_j^{(1)} + \eta_2 N_j^{(2)}\bigr)\bigr]\Bigr)$$
Coordinate $j^*$ drawn with probability $w_{j^*}/\sum_k w_k$. MH correction:
$$\log\frac{q_{\text{rev}}}{q_{\text{fwd}}} = \log w_{j^*}' - \log w_{j^*}
+ \log\textstyle\sum_k w_k - \log\textstyle\sum_k w_k'$$

For **GGM backends**, `init_lb_*_ggm()` is called *after* every GGM column sweep
(adjacency $Z_{\text{ggm}}$ changes each iteration). For **FixedAdj backends**,
initialisation is done once per gamma MH sub-loop; `apply_lb_delta()` updates
only the $O(d_j)$ neighbours of the selected $j^*$. All 12 backends accept
`bool use_lb_gamma` as the last C++ parameter (48/48 signature check passed).
*Reference:* @Zanella2020.

### IMP-7: Vihola Robust Adaptive Metropolis for $\eta$ {#sec-imp7-detail}

**Header:** `src/BayesLogit_Sparse_Helpers.h`

**New C++ struct:**

```{r eval=FALSE}
struct EtaAdapter {
  double log_sigma;   // log-scale proposal SD, clamped to [-10, 5]
  double alpha_star;  // target acceptance rate (0.44 for 1-D)
  double gamma_decay; // Robbins-Monro exponent (2/3)
  int    n_adapt;     // adaptation count

  explicit EtaAdapter(double init_log_sigma = std::log(0.5))
    : log_sigma(init_log_sigma), alpha_star(0.44),
      gamma_decay(2.0/3.0), n_adapt(0) {}

  double sigma() const {
    return std::exp(std::max(-10.0, std::min(5.0, log_sigma)));
  }
  void update(double alpha_n) {           // alpha_n in {0, 1}
    double step = 0.5 * std::min(1.0,
      std::pow(static_cast<double>(++n_adapt), -gamma_decay));
    log_sigma += step * (alpha_n - alpha_star);
  }
};
```

**1-D Robbins-Monro update** [@Vihola2012; @AndrieuThoms2008]:
$$\log\sigma_{n+1} = \log\sigma_n
+ \tfrac{1}{2}\min(1,\, n^{-2/3})(\alpha_n - 0.44)$$
Target $\alpha^* = 0.44$ is the optimal 1-D acceptance rate
[@Roberts1997OptimalScaling]. Step-size clamp $\sigma \in [\exp(-10), \exp(5)]
\approx [4.5\times10^{-5}, 148]$ prevents runaway adaptation.
The $n^{-2/3}$ schedule satisfies the diminishing-adaptation condition
(Theorem 2 of @Vihola2012), ensuring ergodicity of the adaptive chain.

One `EtaAdapter` is declared **per $\eta$ parameter before the MCMC loop**
in each of the 12 backends. After each Möller MH decision,
`adapter.update(static_cast<double>(accept_flag))` is called with the
binary indicator. Dual-network backends declare `eta1_adapter` and `eta2_adapter`
independently.

### IMP-8: Rank-Normalised $\hat{R}$ and Bulk/Tail ESS {#sec-imp8-detail}

**File:** `R/bvs_summary.R`

**Three new exported R functions:**

```{r eval=FALSE}
bvs_rhat(x)      # rank-normalised split-R-hat; x = vector or matrix (cols = chains)
bvs_ess_bulk(x)  # bulk ESS on rank-normalised split chains
bvs_ess_tail(x)  # tail ESS: min(ESS_{q=0.05}, ESS_{q=0.95})
```

**Seven new internal helpers (not exported):**

| Helper | Role |
|--------|------|
| `.ensure_matrix(x)` | Coerce input; single vector → two-column matrix (split-chain) |
| `.split_chains(X)` | Split each column in half, doubling chain count |
| `.z_scale(X)` | Rank-normalise: $z_j = \Phi^{-1}[(r_j - 3/8)/(n+1/4)]$ |
| `.fold_draws(X)` | Fold: $|x_j - \operatorname{median}(x)|$ for tail R-hat |
| `.rhat_basic(X_split)` | Classic Gelman-Rubin $\hat{R}$ on split/normalised chains |
| `.ess_multi_chain(X_split)` | Geyer's initial positive sequence estimator [@Geyer1992] |
| `.ess_quantile(x, q)` | ESS of indicator chain $\mathbf{1}[x_j \leq \hat{F}^{-1}(q)]$ |

**`bvs_rhat(x)` algorithm** [@Vehtari2021]:
1. Split each chain in half → $2M$ chains of length $\lfloor N/2\rfloor$
2. Rank-normalise combined draws: $z_j = \Phi^{-1}[(r_j-3/8)/(N_{\text{tot}}+1/4)]$
3. $\hat{R}_{\text{bulk}}$ = Gelman-Rubin on rank-normalised split chains
4. Fold: $\tilde{x}_j = |x_j - \operatorname{median}(x)|$; repeat steps 1–3 → $\hat{R}_{\text{tail}}$
5. Return $\max(\hat{R}_{\text{bulk}}, \hat{R}_{\text{tail}})$

**`bvs_ess_bulk(x)`:** Split-chain + rank-normalise, then Geyer's initial positive
sequence on the lag-$k$ autocorrelation estimator [@Geyer1992].

**`bvs_ess_tail(x)`:** $\min(\text{ESS}_{q=0.05}, \text{ESS}_{q=0.95})$ where
quantile-ESS uses the indicator-function chain $\mathbf{1}[x_j \leq \hat{F}^{-1}(q)]$.

If the `posterior` R package is installed, all three functions delegate to
`posterior::rhat()`, `posterior::ess_bulk()`, `posterior::ess_tail()` [@Vehtari2021]
for maximum numerical accuracy. `summary.bvs()` now reports both `ESS` (bulk) and
`ESS_tail` for every predictor in the `summary_beta` data frame; convergence
thresholds of $\hat{R} < 1.01$ and ESS $\geq 100$ per chain are recommended [@Vehtari2021].

## Cumulative Change Statistics

| Category | Count |
|----------|-------|
| C++ files modified | 15 (4 headers + 11 backends + RcppExports.cpp) |
| R files modified | 6 |
| New test files | 4 |
| Total issues addressed | 45 |
| New exported R functions | 3 (`bvs_rhat`, `bvs_ess_bulk`, `bvs_ess_tail`) |
| New C++ structs | 2 (`EtaAdapter`, `LBProposalDelta`) |
| New C++ helper functions | 9 (8 LB initialisers/builders + `apply_lb_delta`) |

# Acknowledgments {-}

The authors thank the MetaboAnalyst team for providing the publicly
available cachexia dataset used in the real-data illustration.

# References {-}

