---
title: "BVS.DAdj: Bayesian Variable Selection with Dual-Adjacency Network-Informed Ising Priors in R"
author:
  - name: Yubing Yao
    affiliation: Department of Biostatistics and Epidemiology, University of Massachusetts Amherst
    email: yyao@umass.edu
  - name: Mahlet Tadesse
    affiliation: Department of Mathematics & Statistics, Georgetown University
    email: mgt26@georgetown.edu
  - name: Raji Balasubramanian
    affiliation: Department of Biostatistics and Epidemiology, University of Massachusetts Amherst School of Public Health
    email: rbalasub@schoolph.umass.edu
abstract: >
  This article introduces the **BVS.DAdj** R package, which implements Bayesian
  variable selection for logistic regression models using Markov random field
  (MRF) priors that incorporate dual adjacency structures among predictors.
  The package provides two MCMC sampling strategies---Metropolis-Hastings (MH)
  and Pólya-Gamma (PG) data augmentation---each supporting five adjacency types:
  fixed, dual-fixed, graphical lasso, Bayesian GGM, and combined GGM-plus-fixed.
  Both dense ($p < n$) and sparse high-dimensional ($p \gg n$) backends are
  implemented in C++ via **Rcpp** and **RcppArmadillo** for computational
  efficiency.  We describe the underlying statistical model, the package API,
  and demonstrate its usage through simulation studies under moderate
  ($n = 500, p = 100$) and high-dimensional ($n = 200, p = 1000$) settings,
  including MCMC convergence diagnostics, variable selection performance metrics,
  and running time comparisons across sampling backends.
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: kable
vignette: >
  %\VignetteIndexEntry{BVS.DAdj: R Journal Paper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
is_truthy <- function(x) {
  tolower(trimws(as.character(x))) %in% c("1", "true", "t", "yes", "y", "on")
}

quick_env <- Sys.getenv("BVS_DADJ_QUICK_CHECK", unset = "")
quick_opt <- getOption("bvs.dadj.quick_check", FALSE)
quick_check <- is_truthy(quick_env) || isTRUE(quick_opt)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  cache = TRUE,
  cache.extra = if (quick_check) "quick-check-real-eval-v2" else "full-run-real-eval-v2"
)

# Helper for quick smoke-test runs:
# set Sys.setenv(BVS_DADJ_QUICK_CHECK = "true") before render.
mcmc_scale <- function(full_niter, full_burnin, full_thin = 1L,
                       quick_niter = 5L, quick_burnin = 2L, quick_thin = 1L) {
  if (isTRUE(quick_check)) {
    list(
      niter = as.integer(quick_niter),
      burnin = as.integer(quick_burnin),
      thin = as.integer(quick_thin)
    )
  } else {
    list(
      niter = as.integer(full_niter),
      burnin = as.integer(full_burnin),
      thin = as.integer(full_thin)
    )
  }
}

cat(sprintf("BVS_DAdj_paper quick-check mode: %s\n", if (quick_check) "ON" else "OFF"))

# Robust ACF plotting helper:
# Handles constant or non-finite chains to avoid plot failures during vignette render.
safe_acf_plot <- function(x, lag.max = 50L, main = "", col = "steelblue") {
  x <- as.numeric(x)
  x <- x[is.finite(x)]

  if (length(x) < 2L) {
    plot.new()
    title(main = main)
    mtext("Insufficient finite samples for ACF", side = 3, line = -1.5, cex = 0.85)
    return(invisible(NULL))
  }

  v <- stats::var(x)
  if (!is.finite(v) || v <= .Machine$double.eps) {
    lags <- 0:as.integer(lag.max)
    acf_vals <- c(1, rep(0, length(lags) - 1L))
    plot(lags, acf_vals,
      type = "h", lwd = 2, col = col,
      ylim = c(-0.05, 1.05), xlab = "Lag", ylab = "ACF", main = main
    )
    points(lags, acf_vals, pch = 16, cex = 0.5, col = col)
    mtext("Constant chain: ACF undefined", side = 3, line = 0.1, cex = 0.75)
    return(invisible(NULL))
  }

  stats::acf(x, lag.max = as.integer(lag.max), main = main, col = col)
}

safe_gelman_df <- function(fits, vars = NULL, autoburnin = FALSE) {
  gd <- tryCatch(
    bvs_gelman_diag(fits, vars = vars, autoburnin = autoburnin),
    error = function(e) e
  )

  if (inherits(gd, "error")) {
    return(data.frame(
      Parameter = "NA",
      Point_est = NA_real_,
      Upper_CI = NA_real_,
      Note = conditionMessage(gd),
      stringsAsFactors = FALSE
    ))
  }

  psrf <- as.data.frame(gd$psrf, stringsAsFactors = FALSE)
  if (ncol(psrf) < 2L) {
    return(data.frame(
      Parameter = "NA",
      Point_est = NA_real_,
      Upper_CI = NA_real_,
      Note = "Gelman output did not contain PSRF columns.",
      stringsAsFactors = FALSE
    ))
  }
  names(psrf)[1:2] <- c("Point_est", "Upper_CI")
  psrf$Point_est <- as.numeric(psrf$Point_est)
  psrf$Upper_CI <- as.numeric(psrf$Upper_CI)

  param <- rownames(psrf)
  if (is.null(param) || length(param) != nrow(psrf)) {
    param <- rep("", nrow(psrf))
  }
  param <- trimws(as.character(param))
  blank <- which(!nzchar(param))
  if (length(blank) > 0L) {
    if (is.null(vars)) {
      beta_labels <- paste0("Beta", seq_len(length(blank)))
    } else {
      vars <- as.integer(vars)
      beta_labels <- paste0("Beta", vars[seq_len(min(length(vars), length(blank)))])
      if (length(beta_labels) < length(blank)) {
        beta_labels <- c(
          beta_labels,
          paste0("Beta", seq_len(length(blank) - length(beta_labels)) + length(beta_labels))
        )
      }
    }
    param[blank] <- beta_labels
  }

  out <- data.frame(
    Parameter = param,
    Point_est = psrf$Point_est,
    Upper_CI = psrf$Upper_CI,
    Note = "",
    stringsAsFactors = FALSE
  )
  out
}

gelman_method_summary <- function(gelman_df, method_name, rhat_threshold = 1.1) {
  keep <- is.finite(gelman_df$Point_est)
  if (!any(keep)) {
    return(data.frame(
      Method = method_name,
      N_parameters = 0L,
      Max_Rhat = NA_real_,
      Median_Rhat = NA_real_,
      N_Rhat_gt_1_1 = NA_integer_,
      stringsAsFactors = FALSE
    ))
  }
  data.frame(
    Method = method_name,
    N_parameters = sum(keep),
    Max_Rhat = round(max(gelman_df$Point_est[keep], na.rm = TRUE), 3),
    Median_Rhat = round(stats::median(gelman_df$Point_est[keep], na.rm = TRUE), 3),
    N_Rhat_gt_1_1 = sum(gelman_df$Point_est[keep] > rhat_threshold, na.rm = TRUE),
    stringsAsFactors = FALSE
  )
}

extract_beta_chain <- function(fit, j) {
  b <- fit$beta
  if (is.matrix(b)) {
    j <- max(1L, min(as.integer(j), ncol(b)))
    return(as.numeric(b[, j]))
  }
  if (is.list(b)) {
    n_save <- length(b)
    out <- numeric(n_save)
    for (s in seq_len(n_save)) {
      bs <- b[[s]]
      if (is.null(bs) || length(bs) == 0) {
        out[s] <- 0
      } else if (is.matrix(bs) && ncol(bs) >= 2) {
        ids <- as.integer(bs[, 1])
        vals <- as.numeric(bs[, 2])
        hit <- which(ids == as.integer(j))
        out[s] <- if (length(hit) > 0) vals[hit[1]] else 0
      } else if (is.numeric(bs) && length(bs) >= as.integer(j)) {
        out[s] <- as.numeric(bs[as.integer(j)])
      } else {
        out[s] <- 0
      }
    }
    return(out)
  }
  return(numeric(0))
}

plot_two_chain_trace_panels <- function(fit1, fit2, beta_idx, method_label) {
  beta_idx <- as.integer(beta_idx)
  for (j in beta_idx) {
    ch1 <- extract_beta_chain(fit1, j)
    ch2 <- extract_beta_chain(fit2, j)
    y_lim <- range(c(ch1, ch2), finite = TRUE)
    if (length(y_lim) != 2 || !all(is.finite(y_lim))) y_lim <- c(-1, 1)
    plot(ch1,
      type = "l", col = "steelblue", lwd = 1,
      main = paste0(method_label, ": beta_", j),
      xlab = "Saved iteration", ylab = "beta",
      ylim = y_lim
    )
    lines(ch2, col = "firebrick", lwd = 1)
    legend("topright",
      legend = c("Chain 1", "Chain 2"),
      col = c("steelblue", "firebrick"),
      lty = 1, cex = 0.7, bty = "n"
    )
  }
}

# Explicit FDR-targeted hyperparameters (single and dual network Moller updates).
hpars_sim1 <- list(
  mu = qlogis(0.01),
  eta1_sd = 0.08,
  eta1_sd = 0.10, eta2_sd = 0.06,
  e_eta = 1.0, f_eta = 25.0,
  mu_tilde = -7.0,
  eta1_tilde = 0.005, eta2_tilde = 0.004
)
hpars_sim2 <- list(
  mu = qlogis(0.02),
  eta1_sd = 0.05,
  eta1_sd = 0.06, eta2_sd = 0.05,
  e_eta = 1.0, f_eta = 30.0,
  mu_tilde = qlogis(0.003),
  eta1_tilde = 0.01, eta2_tilde = 0.008
)
hpars_real <- hpars_sim1
hpars_realhd <- hpars_sim2
```

# Introduction

In metabolic case-control studies and high-throughput biological experiments, a
central goal is to identify which of many candidate variables (e.g., metabolites,
genes) are genuinely associated with a disease outcome.  These variables often
exhibit complex dependency structures arising from multiple sources: biological
pathway databases provide partial knowledge of inter-variable relationships,
while data-driven methods can reveal additional conditional dependencies.

Standard penalization methods such as LASSO [@Tib1996] and its extensions
[@ZH2005; @YL2006; @Zou2006] perform variable selection by shrinking regression
coefficients, but may miss weakly associated variables that are correlated with
stronger signals.  Bayesian variable selection with spike-and-slab priors
[@GM1993; @GM1997] provides a principled alternative that quantifies uncertainty
through posterior inclusion probabilities (PIPs).

Building on the Markov random field (MRF) framework of @LZ2010 and @STV2011,
the **BVS.DAdj** package extends the Ising prior on selection indicators
$\boldsymbol{\gamma}$ to incorporate *two independent* adjacency structures
simultaneously---one from prior knowledge and one estimated from data.  This
dual-adjacency formulation allows borrowing strength across both sources of
structural information.

The package implements 12 optimized C++ backends accessed through two
user-facing R functions, `bvs_mh()` and `bvs_pg()`, covering a complete matrix
of sampler type × adjacency source × dimensionality.

The remainder of this article is organized as follows. Section 2 describes the
statistical model and prior specifications. Section 3 outlines the MCMC sampling
strategies. Section 4 describes the package design and API. Sections 5 and 6
present simulation studies for the moderate-dimensional and high-dimensional
settings, respectively. Section 7 summarizes the results.

# Model formulation and priors {#sec-model}

## Logistic regression model

Consider a case-control study with binary outcome $\mathbf{y} = (y_1, \ldots, y_n)^\top$ where
$y_k \in \{0, 1\}$ and a design matrix $\mathbf{X} = (X_{ik})_{n \times p}$
of $p$ candidate predictors for $n$ observations.  The logistic regression model
is
$$
\text{logit}\,\Pr(y_k = 1) = \alpha + \mathbf{X}_k \boldsymbol{\beta}, \quad k = 1, \ldots, n
$$
where $\alpha$ is the intercept and $\boldsymbol{\beta} = (\beta_1, \ldots, \beta_p)^\top$ is the vector of regression coefficients.

## Spike-and-slab prior for variable selection

We introduce binary selection indicators $\boldsymbol{\gamma} = (\gamma_1, \ldots, \gamma_p)^\top \in \{0,1\}^p$ with a point-mass mixture prior on each coefficient:
$$
\beta_j \mid \gamma_j, \sigma^2 \sim \gamma_j \cdot \mathcal{N}(\beta_0, \sigma^2) + (1 - \gamma_j) \cdot \delta_0(\beta_j), \quad j = 1, \ldots, p
$$
The intercept receives a conjugate normal prior $\alpha \sim \mathcal{N}(\alpha_0, h_\alpha \sigma^2)$ with $h_\alpha$ large, and the variance component follows an inverse-gamma distribution:
$$
\sigma^2 \sim \text{Inverse-Gamma}\!\left(\frac{\nu_0}{2},\; \frac{\nu_0 \sigma_0^2}{2}\right)
$$

## Variant MRF prior with dual adjacency

Let $\mathbf{R}$ and $\mathbf{R}'$ be two $p \times p$ symmetric binary adjacency matrices encoding two independent sources of structural information among the predictors.  The prior on $\boldsymbol{\gamma}$ takes the Ising/MRF form:
$$
p(\boldsymbol{\gamma} \mid \mu, \eta_1, \eta_2) = \frac{1}{Z(\mu,\eta_1,\eta_2)}\exp\!\Big(\mu\,\mathbf{1}_p^\top \boldsymbol{\gamma} + \eta_1\, \boldsymbol{\gamma}^\top \mathbf{R}\, \boldsymbol{\gamma} + \eta_2\, \boldsymbol{\gamma}^\top \mathbf{R}'\, \boldsymbol{\gamma}\Big)
$$
where:

- $\mu$ controls the baseline sparsity level.  For a predictor with no neighbors in either graph, the marginal inclusion probability is $\text{expit}(\mu)$.
- $\eta_1, \eta_2 \geq 0$ are coupling parameters controlling how strongly a predictor's inclusion is influenced by its neighbors in $\mathbf{R}$ and $\mathbf{R}'$, respectively.
- $Z(\mu, \eta_1, \eta_2)$ is the intractable normalizing constant.

**Eta prior used in the package.**
In the implementation, each coupling parameter is bounded by a user-specified
upper limit (`eta1_sd`, `eta1_sd`, `eta2_sd`) and sampled on a scaled Beta domain:
$$
\eta_i = \eta_{i,\max} u_i,\quad u_i \sim \text{Beta}(e, f), \quad i=1,2.
$$
The helper `phase_transition()` can be used to choose practically stable values
for $\eta_{i,\max}$ before fitting the full model.

## Estimating unknown adjacency: Bayesian GGM

When the second adjacency $\mathbf{R}'$ is unknown, we model the predictors as
$\mathbf{X}_k \sim \mathcal{N}_p(\mathbf{0}, \boldsymbol{\Omega}^{-1})$ and place a continuous spike-and-slab SSVS prior on $\boldsymbol{\Omega}$ following @WH2015:
$$
p(\boldsymbol{\Omega} \mid \mathbf{R}', \theta) \propto \prod_{i < j} \mathcal{N}\!\left(\omega_{ij} \mid 0,\, v^2_{r'_{ij}}\right) \prod_i \text{Exp}\!\left(\omega_{ii} \mid \tfrac{\lambda}{2}\right) \cdot \mathbf{1}_{\boldsymbol{\Omega} \in M^+}
$$
where $v_{r'_{ij}} = v_0$ (spike) if $r'_{ij} = 0$ and $v_1$ (slab) if $r'_{ij} = 1$.  Edge indicators are updated via:
$$
\Pr(r'_{ij} = 1 \mid \boldsymbol{\Omega}) = \frac{\pi\, \mathcal{N}(\omega_{ij} \mid 0, v_1^2)}{\pi\, \mathcal{N}(\omega_{ij} \mid 0, v_1^2) + (1-\pi)\, \mathcal{N}(\omega_{ij} \mid 0, v_0^2)}
$$

# MCMC sampling strategies {#sec-mcmc}

## Metropolis-Hastings sampler

The MH sampler (`bvs_mh()`) uses random-walk proposals for $\boldsymbol{\beta}$, $\alpha$, and $\sigma^2$, and a component-wise add/delete move for $\gamma_j$.  For each $j$, a flip $\gamma_j^* = 1 - \gamma_j$ is proposed with the corresponding $\beta_j^*$ drawn from the spike-and-slab prior; the move is accepted with probability $\min(1, \exp(\Delta))$ where
$$
\Delta = \ell(\boldsymbol{\beta}^*, \boldsymbol{\gamma}^*) - \ell(\boldsymbol{\beta}, \boldsymbol{\gamma}) - \frac{(\beta_j - \beta_0)^2 - (\beta_j^* - \beta_0)^2}{2\sigma^2}
$$

## Pólya-Gamma augmented Gibbs sampler

The PG sampler (`bvs_pg()`) introduces auxiliary variables $\omega_k \sim \text{PG}(1, \psi_k)$
where $\psi_k = \alpha + \mathbf{X}_k \boldsymbol{\beta}$, rendering the logistic likelihood conditionally Gaussian [@PG2013].  In the C++ backend, PG draws are obtained through a truncated-series approximation, after which block updates for $(\boldsymbol{\beta}, \alpha)$ use Gaussian conditional forms:
$$
\boldsymbol{\beta}_\gamma \mid \cdot \sim \mathcal{N}\!\left(\mathbf{V}_\gamma \mathbf{m}_\gamma,\; \mathbf{V}_\gamma\right), \quad \mathbf{V}_\gamma = \left(\mathbf{X}_\gamma^\top \boldsymbol{\Omega}_\omega \mathbf{X}_\gamma + \sigma^{-2} \mathbf{I}\right)^{-1}
$$
where $\boldsymbol{\Omega}_\omega = \text{diag}(\omega_1, \ldots, \omega_n)$ and $\kappa_k = y_k - 1/2$.

## Möller auxiliary variable update for $\eta_1, \eta_2$

The normalizing constant $Z(\mu, \eta_1, \eta_2)$ is intractable.  Following @MB2006, we introduce auxiliary variables $\boldsymbol{\omega}$ drawn via Propp-Wilson perfect simulation from the Ising model at reference parameters $(\tilde{\mu}, \tilde{\eta}_1, \tilde{\eta}_2)$, yielding an MH ratio that cancels the intractable ratio $Z'/Z$.

## Sparse high-dimensional variant

For $p \gg n$, the package stores the empirical covariance $\mathbf{S} = \mathbf{X}^\top\mathbf{X}$ in compressed sparse column (CSC) format and uses column-wise block Gibbs steps for the precision matrix.  The selection indicator updates use a targeted MH scheme cycling through `n_mh_gamma` randomly chosen variables per iteration.

# Package design and API {#sec-api}

## Installation

```{r install, eval=FALSE}
# Install from GitHub
devtools::install_github("yuy113/BVS.DAdj")
```

```{r load-pkg}
library(BVS.DAdj)
```

## User-facing functions

The package exports two primary functions that dispatch to 12 C++ backends:

| Function   | Sampler | Primary usage |
|:-----------|:--------|:--------------|
| `bvs_mh()` | Metropolis-Hastings | All adjacency types; general purpose |
| `bvs_pg()` | Pólya-Gamma Gibbs   | All adjacency types; faster mixing for moderate $p$ |

Both accept a common interface:

```{r api-example, eval=FALSE}
fit <- bvs_mh(X, y,
  adj_type = c("fixed", "dual_fixed", "glasso", "glasso_fixed", "ggm", "ggm_fixed"),
  adj_fixed = NULL, # Required for "fixed", "dual_fixed", "glasso_fixed", "ggm_fixed"
  adj_fixed2 = NULL, # Required for "dual_fixed"
  sparse = FALSE, # TRUE for sparse-GGM backends
  niter = 60000, burnin = 10000, thin = 1,
  # ... additional hyperparameters
)
```

## Adjacency types

| `adj_type` | Adjacency source(s) | Eta | Backend count |
|:-----------|:---------------------|:----|:--------------|
| `"fixed"` | One user-supplied binary matrix | 1 | MH + PG |
| `"dual_fixed"` | Two user-supplied binary matrices | 2 | MH + PG |
| `"glasso"` | Graphical lasso estimated from $\mathbf{X}$ | 1 | MH + PG |
| `"glasso_fixed"` | Graphical lasso from $\mathbf{X}$ + one fixed matrix | 2 | MH + PG |
| `"ggm"` | Bayesian GGM SSVS (dense or sparse) | 1 | MH + PG × dense/sparse |
| `"ggm_fixed"` | GGM + one user-supplied fixed matrix | 2 | MH + PG × dense/sparse |

## Post-processing

```{r post-process, eval=FALSE}
summary(fit) # Selected variables, PIPs, posterior means/intervals
plot(fit) # PIP bar plot (optionally top_n)
```

# Simulation study 1: Moderate dimensions ($n = 500, p = 100$) {#sec-sim1}

## Data generation

We generate a logistic regression dataset with $n = 500$ observations, $p = 100$ covariates, $p^* = 10$ truly active variables, and a block-correlation structure.  Covariates within the active set have pairwise correlation $\rho = 0.5$ to emulate metabolite clustering.

```{r sim1-data}
set.seed(2024)
n <- 500
p <- 100
p_star <- 10

# Construct correlation matrix: block structure for first p_star variables
Sigma <- diag(p)
for (i in 1:p_star) {
  for (j in 1:p_star) {
    if (i != j) Sigma[i, j] <- 0.5
  }
}

# Generate covariates from multivariate normal
X <- MASS::mvrnorm(n, mu = rep(0, p), Sigma = Sigma)

# True coefficients: first p_star variables have nonzero effects
beta_true <- rep(0, p)
beta_true[1:p_star] <- c(1.5, -1.2, 1.0, -0.8, 0.6, -0.5, 0.4, -0.3, 0.25, -0.2)

# Generate binary response
eta_true <- X %*% beta_true
prob_true <- plogis(eta_true)
y <- rbinom(n, 1, prob_true)

cat(sprintf(
  "Response balance: %d cases / %d controls (%.1f%%)\n",
  sum(y), sum(1 - y), 100 * mean(y)
))
```

## Adjacency matrix construction

For the additional adjacency-type comparison, we construct:

1. A **true** adjacency matrix `R_true1` based on the simulation graph used to
   generate correlated predictors (first `p_star` variables fully connected).
2. An **external fixed** adjacency matrix `R_ext50_1` by randomly retaining
   50% of the `1` entries in `R_true1` (symmetric, zero diagonal).

For backward compatibility with the original baseline comparison in this paper,
we also keep `R_known` (partial/noisy fixed adjacency) used by the original
three-method benchmark.

```{r sim1-adj}
# True adjacency used in simulation design (first p_star block)
R_true1 <- matrix(0L, p, p)
for (i in 1:p_star) {
  for (j in (i + 1):p_star) {
    R_true1[i, j] <- R_true1[j, i] <- 1L
  }
}
diag(R_true1) <- 0L

# External fixed adjacency: random 50% of true edges
set.seed(202401)
true_pairs1 <- which(upper.tri(R_true1) & R_true1 == 1L, arr.ind = TRUE)
n_keep1 <- ceiling(0.5 * nrow(true_pairs1))
keep_idx1 <- sample(seq_len(nrow(true_pairs1)), n_keep1)
keep_pairs1 <- true_pairs1[keep_idx1, , drop = FALSE]
R_ext50_1 <- matrix(0L, p, p)
R_ext50_1[cbind(keep_pairs1[, 1], keep_pairs1[, 2])] <- 1L
R_ext50_1[cbind(keep_pairs1[, 2], keep_pairs1[, 1])] <- 1L
diag(R_ext50_1) <- 0L

# Legacy partial/noisy adjacency from earlier version of the vignette
R_known <- matrix(0L, p, p)
# Block edges among first 8 active variables
for (i in 1:8) {
  for (j in (i + 1):8) {
    R_known[i, j] <- R_known[j, i] <- 1L
  }
}
# Add a few noise edges
set.seed(123)
noise_pairs <- matrix(sample((p_star + 1):p, 10), ncol = 2)
for (k in 1:nrow(noise_pairs)) {
  R_known[noise_pairs[k, 1], noise_pairs[k, 2]] <- 1L
  R_known[noise_pairs[k, 2], noise_pairs[k, 1]] <- 1L
}
diag(R_known) <- 0L
cat(sprintf("True adjacency edges (R_true1): %d\n", sum(R_true1) / 2))
cat(sprintf("External 50%% adjacency edges (R_ext50_1): %d\n", sum(R_ext50_1) / 2))
cat(sprintf("Legacy fixed adjacency edges (R_known): %d\n", sum(R_known) / 2))
```

## Running the samplers

We run three backends on this dataset:

1. **MH with fixed adjacency** (`bvs_mh`, `adj_type = "fixed"`)
2. **PG with fixed adjacency** (`bvs_pg`, `adj_type = "fixed"`)
3. **PG with Bayesian GGM** (`bvs_pg`, `adj_type = "ggm"`, dense)

```{r sim1-run, cache=TRUE}
cfg_sim1 <- mcmc_scale(
  full_niter = 500L, full_burnin = 200L, full_thin = 1L,
  quick_niter = 5L, quick_burnin = 2L, quick_thin = 1L
)
niter_sim <- cfg_sim1$niter
burnin_sim <- cfg_sim1$burnin
thin_sim <- cfg_sim1$thin

# --- MH with fixed adjacency ---
t_mh <- system.time({
  fit_mh <- bvs_mh(X, y,
    adj_type = "fixed", adj_fixed = R_known,
    niter = niter_sim, burnin = burnin_sim,
    mu = -log(1 / 0.1 - 1), thin = thin_sim
  )
})

# --- PG with fixed adjacency ---
t_pg <- system.time({
  fit_pg <- bvs_pg(X, y,
    adj_type = "fixed", adj_fixed = R_known,
    niter = niter_sim, burnin = burnin_sim,
    mu = -log(1 / 0.1 - 1), thin = thin_sim
  )
})

# --- PG with dense Bayesian GGM ---
t_pg_ggm <- system.time({
  fit_pg_ggm <- bvs_pg(X, y,
    adj_type = "ggm", sparse = FALSE,
    niter = niter_sim, burnin = burnin_sim,
    mu = -log(1 / 0.1 - 1), thin = thin_sim
  )
})
```

## Running time comparison

```{r sim1-timing}
timing_df <- data.frame(
  Method = c("MH (fixed adj)", "PG (fixed adj)", "PG (Bayesian GGM)"),
  Elapsed_sec = round(c(t_mh["elapsed"], t_pg["elapsed"], t_pg_ggm["elapsed"]), 1),
  Iter_per_sec = round(niter_sim / c(t_mh["elapsed"], t_pg["elapsed"], t_pg_ggm["elapsed"]), 0)
)
knitr::kable(
  timing_df,
  caption = sprintf(
    "Running time comparison ($n = 500, p = 100$, %s post-burn-in iterations, burn-in %s).",
    format(niter_sim, big.mark = ","),
    format(burnin_sim, big.mark = ",")
  ),
  col.names = c("Method", "Elapsed (sec)", "Iter/sec")
)
```

## MCMC convergence diagnostics

### Trace plots

```{r sim1-trace, fig.height=8, fig.width=8, fig.cap="Trace plots of selected regression coefficients (beta_1 through beta_4) from the three samplers. The dashed red lines indicate true coefficient values."}
par(mfrow = c(3, 4), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
vars_to_plot <- 1:4

for (method_name in c("MH (fixed)", "PG (fixed)", "PG (GGM)")) {
  fit_cur <- switch(method_name,
    "MH (fixed)" = fit_mh,
    "PG (fixed)" = fit_pg,
    "PG (GGM)"   = fit_pg_ggm
  )
  for (j in vars_to_plot) {
    plot(fit_cur$beta[, j],
      type = "l", col = "steelblue",
      main = paste0(method_name, ": beta_", j),
      xlab = "Iteration", ylab = expression(beta[j]),
      cex.main = 0.85
    )
    abline(h = beta_true[j], col = "red", lty = 2, lwd = 1.5)
  }
}
```

### Running mean plots

```{r sim1-runmean, fig.height=6, fig.width=8, fig.cap="Running mean of beta_1 and beta_2 across the three sampling methods, demonstrating convergence to the true value (dashed red line)."}
running_mean <- function(x) cumsum(x) / seq_along(x)

par(mfrow = c(2, 3), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (j in 1:2) {
  for (method_name in c("MH (fixed)", "PG (fixed)", "PG (GGM)")) {
    fit_cur <- switch(method_name,
      "MH (fixed)" = fit_mh,
      "PG (fixed)" = fit_pg,
      "PG (GGM)"   = fit_pg_ggm
    )
    rm_vals <- running_mean(fit_cur$beta[, j])
    plot(rm_vals,
      type = "l", col = "darkgreen", lwd = 1.2,
      main = paste0(method_name, ": beta_", j),
      xlab = "Iteration", ylab = "Running mean",
      cex.main = 0.85
    )
    abline(h = beta_true[j], col = "red", lty = 2, lwd = 1.5)
  }
}
```

### Autocorrelation

```{r sim1-acf, fig.height=4, fig.width=8, fig.cap="Autocorrelation function of beta_1 for the three sampling methods."}
par(mfrow = c(1, 3), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
safe_acf_plot(fit_mh$beta[, 1], lag.max = 50, main = "MH (fixed): beta_1", col = "steelblue")
safe_acf_plot(fit_pg$beta[, 1], lag.max = 50, main = "PG (fixed): beta_1", col = "steelblue")
safe_acf_plot(fit_pg_ggm$beta[, 1], lag.max = 50, main = "PG (GGM): beta_1", col = "steelblue")
```

### Gelman-Rubin diagnostics (2 chains, 2 seeds)

```{r sim1-gelman, cache=TRUE}
vars_gr1 <- 1:4
seed_sim1 <- c(20261101L, 20261102L)

set.seed(seed_sim1[1])
fit_mh_gr1 <- bvs_mh(
  X, y,
  adj_type = "fixed", adj_fixed = R_known,
  niter = niter_sim, burnin = burnin_sim,
  mu = -log(1 / 0.1 - 1), thin = thin_sim
)
set.seed(seed_sim1[2])
fit_mh_gr2 <- bvs_mh(
  X, y,
  adj_type = "fixed", adj_fixed = R_known,
  niter = niter_sim, burnin = burnin_sim,
  mu = -log(1 / 0.1 - 1), thin = thin_sim
)

set.seed(seed_sim1[1])
fit_pg_gr1 <- bvs_pg(
  X, y,
  adj_type = "fixed", adj_fixed = R_known,
  niter = niter_sim, burnin = burnin_sim,
  mu = -log(1 / 0.1 - 1), thin = thin_sim
)
set.seed(seed_sim1[2])
fit_pg_gr2 <- bvs_pg(
  X, y,
  adj_type = "fixed", adj_fixed = R_known,
  niter = niter_sim, burnin = burnin_sim,
  mu = -log(1 / 0.1 - 1), thin = thin_sim
)

set.seed(seed_sim1[1])
fit_pg_ggm_gr1 <- bvs_pg(
  X, y,
  adj_type = "ggm", sparse = FALSE,
  niter = niter_sim, burnin = burnin_sim,
  mu = -log(1 / 0.1 - 1), thin = thin_sim
)
set.seed(seed_sim1[2])
fit_pg_ggm_gr2 <- bvs_pg(
  X, y,
  adj_type = "ggm", sparse = FALSE,
  niter = niter_sim, burnin = burnin_sim,
  mu = -log(1 / 0.1 - 1), thin = thin_sim
)

gelman_sim1 <- list(
  "MH (fixed adj)" = safe_gelman_df(list(fit_mh_gr1, fit_mh_gr2), vars = vars_gr1, autoburnin = FALSE),
  "PG (fixed adj)" = safe_gelman_df(list(fit_pg_gr1, fit_pg_gr2), vars = vars_gr1, autoburnin = FALSE),
  "PG (Bayesian GGM)" = safe_gelman_df(list(fit_pg_ggm_gr1, fit_pg_ggm_gr2), vars = vars_gr1, autoburnin = FALSE)
)

gelman_sim1_summary <- do.call(rbind, lapply(names(gelman_sim1), function(mn) {
  gelman_method_summary(gelman_sim1[[mn]], mn)
}))

gelman_sim1_detail <- do.call(rbind, lapply(names(gelman_sim1), function(mn) {
  df <- gelman_sim1[[mn]]
  data.frame(
    Method = mn,
    Parameter = df$Parameter,
    Point_Rhat = round(df$Point_est, 3),
    Upper_CI_Rhat = round(df$Upper_CI, 3),
    stringsAsFactors = FALSE
  )
}))

knitr::kable(
  gelman_sim1_summary,
  caption = "Simulation 1 Gelman-Rubin diagnostics using two chains (two seeds) per method."
)
knitr::kable(
  gelman_sim1_detail,
  caption = "Simulation 1 PSRF details (Alpha and beta_1 to beta_4)."
)
```

```{r sim1-gelman-trace-2chain, fig.height=10, fig.width=12, fig.cap="Simulation 1: two-chain trace plots for representative coefficients across Gelman-Rubin diagnostic runs."}
beta_idx_gr1_trace <- unique(c(1L, 2L, 5L, min(p, p_star + 5L)))
par(
  mfrow = c(3, length(beta_idx_gr1_trace)),
  mar = c(3, 3, 2, 1),
  mgp = c(1.8, 0.5, 0)
)
plot_two_chain_trace_panels(fit_mh_gr1, fit_mh_gr2, beta_idx_gr1_trace, "MH (fixed)")
plot_two_chain_trace_panels(fit_pg_gr1, fit_pg_gr2, beta_idx_gr1_trace, "PG (fixed)")
plot_two_chain_trace_panels(fit_pg_ggm_gr1, fit_pg_ggm_gr2, beta_idx_gr1_trace, "PG (Bayesian GGM)")
```

## Variable selection performance

### Posterior inclusion probabilities

```{r sim1-pip, fig.height=5, fig.width=9, fig.cap="Posterior inclusion probabilities (PIPs) across all 100 variables. Red bars indicate truly active variables (1--10). The dashed line marks the 0.5 median probability threshold."}
compute_pip <- function(fit) {
  if (!is.null(fit$gamma_pip)) {
    return(as.numeric(fit$gamma_pip))
  }
  if (is.matrix(fit$gamma)) {
    return(colMeans(fit$gamma))
  }
  if (is.list(fit$gamma)) {
    p <- fit$p
    n_save <- length(fit$gamma)
    pip <- numeric(p)
    if (n_save > 0) {
      for (s in seq_len(n_save)) {
        idx <- as.integer(fit$gamma[[s]])
        idx <- idx[idx >= 1L & idx <= p]
        if (length(idx) > 0) pip[idx] <- pip[idx] + 1
      }
      pip <- pip / n_save
    }
    return(pip)
  }
  stop("No gamma samples available for PIP computation.")
}

get_gamma_matrix <- function(fit) {
  if (is.matrix(fit$gamma)) {
    return(fit$gamma)
  }
  if (is.list(fit$gamma)) {
    p <- fit$p
    n_save <- length(fit$gamma)
    G <- matrix(0, n_save, p)
    if (n_save > 0) {
      for (s in seq_len(n_save)) {
        idx <- as.integer(fit$gamma[[s]])
        idx <- idx[idx >= 1L & idx <= p]
        if (length(idx) > 0) G[s, idx] <- 1
      }
    }
    return(G)
  }
  stop("No gamma samples available.")
}

get_beta_matrix <- function(fit) {
  if (is.matrix(fit$beta)) {
    return(fit$beta)
  }
  if (is.list(fit$beta)) {
    p <- fit$p
    n_save <- length(fit$beta)
    B <- matrix(0, n_save, p)
    if (n_save > 0) {
      for (s in seq_len(n_save)) {
        bm <- fit$beta[[s]]
        if (is.null(bm) || length(bm) == 0) next
        bm <- as.matrix(bm)
        if (nrow(bm) > 0 && ncol(bm) >= 2) {
          idx <- as.integer(bm[, 1]) # C++ sparse backends emit 1-based indices
          valid <- idx >= 1L & idx <= p
          B[s, idx[valid]] <- as.numeric(bm[valid, 2])
        }
      }
    }
    return(B)
  }
  stop("No beta samples available.")
}

model_size_chain <- function(fit) {
  if (is.matrix(fit$gamma)) {
    return(rowSums(fit$gamma))
  }
  if (is.list(fit$gamma)) {
    return(vapply(fit$gamma, length, integer(1)))
  }
  stop("No gamma samples available.")
}

pip_mh <- compute_pip(fit_mh)
pip_pg <- compute_pip(fit_pg)
pip_ggm <- compute_pip(fit_pg_ggm)

par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
cols <- ifelse(1:p <= p_star, "firebrick", "grey60")

barplot(pip_mh,
  col = cols, border = NA, main = "MH (fixed adj)",
  xlab = "Variable index", ylab = "PIP", ylim = c(0, 1)
)
abline(h = 0.5, lty = 2, col = "black")

barplot(pip_pg,
  col = cols, border = NA, main = "PG (fixed adj)",
  xlab = "Variable index", ylab = "PIP", ylim = c(0, 1)
)
abline(h = 0.5, lty = 2, col = "black")

barplot(pip_ggm,
  col = cols, border = NA, main = "PG (Bayesian GGM)",
  xlab = "Variable index", ylab = "PIP", ylim = c(0, 1)
)
abline(h = 0.5, lty = 2, col = "black")
```

### Selection metrics

```{r sim1-metrics}
compute_metrics <- function(pip, threshold = 0.5, p_star = 10) {
  selected <- which(pip >= threshold)
  tp <- sum(selected <= p_star)
  fp <- sum(selected > p_star)
  fn <- p_star - tp
  sensitivity <- tp / p_star
  fdr <- if (length(selected) > 0) fp / length(selected) else 0
  f1 <- if (tp + fp + fn > 0) 2 * tp / (2 * tp + fp + fn) else 0
  data.frame(
    TP = tp, FP = fp, FN = fn,
    Sensitivity = round(sensitivity, 3),
    FDR = round(fdr, 3),
    F1 = round(f1, 3),
    N_selected = length(selected)
  )
}

metrics <- rbind(
  cbind(Method = "MH (fixed adj)", compute_metrics(pip_mh)),
  cbind(Method = "PG (fixed adj)", compute_metrics(pip_pg)),
  cbind(Method = "PG (Bayesian GGM)", compute_metrics(pip_ggm))
)
knitr::kable(metrics,
  caption = "Variable selection performance at PIP threshold 0.5 ($n = 500, p = 100$).",
  row.names = FALSE
)
```

### Eta coupling parameter traces

```{r sim1-eta, fig.height=4, fig.width=8, fig.cap="Trace plots of the Ising coupling parameter eta for the three sampling methods."}
par(mfrow = c(1, 3), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
if (!is.null(fit_mh$eta1)) {
  plot(fit_mh$eta1,
    type = "l", col = "purple",
    main = "MH: eta1", xlab = "Iteration", ylab = expression(eta)
  )
}
if (!is.null(fit_pg$eta1)) {
  plot(fit_pg$eta1,
    type = "l", col = "purple",
    main = "PG: eta1", xlab = "Iteration", ylab = expression(eta)
  )
}
if (!is.null(fit_pg_ggm$eta1)) {
  plot(fit_pg_ggm$eta1,
    type = "l", col = "purple",
    main = "PG GGM: eta1", xlab = "Iteration", ylab = expression(eta)
  )
}
```

## Additional adjacency-type comparison (5 methods) using 50% true external edges

To compare adjacency **types** directly, we run five additional methods using a
common MH sampler and the same simulated data. For fixed-network inputs, we use
`R_ext50_1` (random 50% of true adjacency edges):

1. `adj_type = "fixed"` (external `R_ext50_1` only),
2. `adj_type = "ggm"` (Bayesian GGM),
3. `adj_type = "glasso"` (data-driven glasso only),
4. `adj_type = "glasso_fixed"` (external `R_ext50_1` + glasso),
5. `adj_type = "ggm_fixed"` (external `R_ext50_1` + Bayesian GGM).

```{r sim1-adjtype-run, cache=TRUE}
sim1_adj_dir <- "sim1_adjtype_saved_runs"
if (!dir.exists(sim1_adj_dir)) dir.create(sim1_adj_dir, recursive = TRUE)

niter_adj1 <- niter_sim
burnin_adj1 <- burnin_sim
thin_adj1 <- thin_sim

sim1_f_fixed <- file.path(sim1_adj_dir, sprintf("sim1_fixed50_n%d_b%d_t%d.rds", niter_adj1, burnin_adj1, thin_adj1))
sim1_f_ggm <- file.path(sim1_adj_dir, sprintf("sim1_ggm_n%d_b%d_t%d.rds", niter_adj1, burnin_adj1, thin_adj1))
sim1_f_gls <- file.path(sim1_adj_dir, sprintf("sim1_glasso_n%d_b%d_t%d.rds", niter_adj1, burnin_adj1, thin_adj1))
sim1_f_gf <- file.path(sim1_adj_dir, sprintf("sim1_glasso_fixed_n%d_b%d_t%d.rds", niter_adj1, burnin_adj1, thin_adj1))
sim1_f_ggmf <- file.path(sim1_adj_dir, sprintf("sim1_ggm_fixed_n%d_b%d_t%d.rds", niter_adj1, burnin_adj1, thin_adj1))

if (file.exists(sim1_f_fixed)) {
  o <- readRDS(sim1_f_fixed)
  fit1_fixed50 <- o$fit
  t1_fixed50 <- o$time
} else {
  t1_fixed50 <- system.time({
    fit1_fixed50 <- bvs_mh(
      X, y,
      adj_type = "fixed", adj_fixed = R_ext50_1,
      niter = niter_adj1, burnin = burnin_adj1, thin = thin_adj1,
      mu = -log(1 / 0.1 - 1)
    )
  })
  saveRDS(list(fit = fit1_fixed50, time = t1_fixed50), sim1_f_fixed)
}

if (file.exists(sim1_f_ggm)) {
  o <- readRDS(sim1_f_ggm)
  fit1_ggm <- o$fit
  t1_ggm <- o$time
} else {
  t1_ggm <- system.time({
    fit1_ggm <- bvs_mh(
      X, y,
      adj_type = "ggm", sparse = FALSE,
      niter = niter_adj1, burnin = burnin_adj1, thin = thin_adj1,
      mu = -log(1 / 0.1 - 1)
    )
  })
  saveRDS(list(fit = fit1_ggm, time = t1_ggm), sim1_f_ggm)
}

if (file.exists(sim1_f_gls)) {
  o <- readRDS(sim1_f_gls)
  fit1_glasso <- o$fit
  t1_glasso <- o$time
} else {
  t1_glasso <- system.time({
    fit1_glasso <- bvs_mh(
      X, y,
      adj_type = "glasso", glasso_criterion = "ebic",
      niter = niter_adj1, burnin = burnin_adj1, thin = thin_adj1,
      mu = -log(1 / 0.1 - 1)
    )
  })
  saveRDS(list(fit = fit1_glasso, time = t1_glasso), sim1_f_gls)
}

if (file.exists(sim1_f_gf)) {
  o <- readRDS(sim1_f_gf)
  fit1_glasso_fixed <- o$fit
  t1_glasso_fixed <- o$time
} else {
  t1_glasso_fixed <- system.time({
    fit1_glasso_fixed <- bvs_mh(
      X, y,
      adj_type = "glasso_fixed", adj_fixed = R_ext50_1,
      glasso_criterion = "ebic",
      niter = niter_adj1, burnin = burnin_adj1, thin = thin_adj1,
      mu = hpars_sim1$mu,
      eta1_sd = hpars_sim1$eta1_sd, eta2_sd = hpars_sim1$eta2_sd,
      e_eta = hpars_sim1$e_eta, f_eta = hpars_sim1$f_eta,
      mu_tilde = hpars_sim1$mu_tilde,
      eta1_tilde = hpars_sim1$eta1_tilde, eta2_tilde = hpars_sim1$eta2_tilde
    )
  })
  saveRDS(list(fit = fit1_glasso_fixed, time = t1_glasso_fixed), sim1_f_gf)
}

if (file.exists(sim1_f_ggmf)) {
  o <- readRDS(sim1_f_ggmf)
  fit1_ggm_fixed <- o$fit
  t1_ggm_fixed <- o$time
} else {
  t1_ggm_fixed <- system.time({
    fit1_ggm_fixed <- bvs_mh(
      X, y,
      adj_type = "ggm_fixed", adj_fixed = R_ext50_1, sparse = FALSE,
      niter = niter_adj1, burnin = burnin_adj1, thin = thin_adj1,
      mu = hpars_sim1$mu,
      eta1_sd = hpars_sim1$eta1_sd, eta2_sd = hpars_sim1$eta2_sd,
      e_eta = hpars_sim1$e_eta, f_eta = hpars_sim1$f_eta,
      mu_tilde = hpars_sim1$mu_tilde,
      eta1_tilde = hpars_sim1$eta1_tilde, eta2_tilde = hpars_sim1$eta2_tilde
    )
  })
  saveRDS(list(fit = fit1_ggm_fixed, time = t1_ggm_fixed), sim1_f_ggmf)
}
```

```{r sim1-adjtype-timing}
timing_adj1 <- data.frame(
  Method = c("Fixed (50% true edges)", "GGM", "Glasso", "Glasso + Fixed", "GGM + Fixed"),
  Elapsed_sec = round(c(
    t1_fixed50["elapsed"], t1_ggm["elapsed"], t1_glasso["elapsed"],
    t1_glasso_fixed["elapsed"], t1_ggm_fixed["elapsed"]
  ), 2),
  Iter_per_sec = round(niter_adj1 / c(
    t1_fixed50["elapsed"], t1_ggm["elapsed"], t1_glasso["elapsed"],
    t1_glasso_fixed["elapsed"], t1_ggm_fixed["elapsed"]
  ), 1)
)
knitr::kable(
  timing_adj1,
  caption = "Simulation 1 adjacency-type comparison timing (MH sampler; fixed input uses random 50% of true adjacency edges).",
  row.names = FALSE
)
```

```{r sim1-adjtype-pip, fig.height=8, fig.width=10, fig.cap="Simulation 1: PIP comparison across adjacency types (fixed, ggm, glasso, glasso_fixed, ggm_fixed)."}
pip1_fixed50 <- compute_pip(fit1_fixed50)
pip1_ggm <- compute_pip(fit1_ggm)
pip1_glasso <- compute_pip(fit1_glasso)
pip1_gf <- compute_pip(fit1_glasso_fixed)
pip1_ggmf <- compute_pip(fit1_ggm_fixed)

show_vars1 <- unique(c(
  1:40,
  which(pip1_fixed50 > 0.1),
  which(pip1_ggm > 0.1),
  which(pip1_glasso > 0.1),
  which(pip1_gf > 0.1),
  which(pip1_ggmf > 0.1)
))
show_vars1 <- sort(show_vars1)
cols1 <- ifelse(show_vars1 <= p_star, "firebrick", "grey65")

pip_list1 <- list(
  "Fixed (50% true)" = pip1_fixed50,
  "GGM" = pip1_ggm,
  "Glasso" = pip1_glasso,
  "Glasso + Fixed" = pip1_gf,
  "GGM + Fixed" = pip1_ggmf
)

par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))
for (mn in names(pip_list1)) {
  barplot(pip_list1[[mn]][show_vars1],
    col = cols1, border = NA, ylim = c(0, 1),
    main = mn, xlab = "Variable", ylab = "PIP",
    names.arg = show_vars1, las = 2, cex.names = 0.5
  )
  abline(h = 0.5, lty = 2)
}
plot.new()
```

```{r sim1-adjtype-metrics}
metrics_adj1 <- rbind(
  cbind(Method = "Fixed (50% true)", compute_metrics(pip1_fixed50, p_star = p_star)),
  cbind(Method = "GGM", compute_metrics(pip1_ggm, p_star = p_star)),
  cbind(Method = "Glasso", compute_metrics(pip1_glasso, p_star = p_star)),
  cbind(Method = "Glasso + Fixed", compute_metrics(pip1_gf, p_star = p_star)),
  cbind(Method = "GGM + Fixed", compute_metrics(pip1_ggmf, p_star = p_star))
)
knitr::kable(
  metrics_adj1,
  caption = "Simulation 1 variable-selection performance for five adjacency-type methods.",
  row.names = FALSE
)
```

```{r sim1-adjtype-diagnostics}
safe_ess_sim <- function(x, lag.max = 200L) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  n <- length(x)
  if (n < 3L) {
    return(NA_real_)
  }
  v <- stats::var(x)
  if (!is.finite(v) || v <= .Machine$double.eps) {
    return(n)
  }
  ac <- as.numeric(stats::acf(x, lag.max = min(lag.max, n - 1L), plot = FALSE)$acf)[-1L]
  if (!length(ac)) {
    return(n)
  }
  first_nonpos <- which(ac <= 0)
  m <- if (length(first_nonpos)) first_nonpos[1L] - 1L else length(ac)
  if (m < 1L) {
    return(n)
  }
  tau <- 1 + 2 * sum(ac[seq_len(m)])
  if (!is.finite(tau) || tau <= 0) {
    return(n)
  }
  min(n, n / tau)
}

lag1_acf_sim <- function(x) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  if (length(x) < 3L) {
    return(NA_real_)
  }
  v <- stats::var(x)
  if (!is.finite(v) || v <= .Machine$double.eps) {
    return(NA_real_)
  }
  as.numeric(stats::acf(x, lag.max = 1L, plot = FALSE)$acf[2L])
}

fits_adj1 <- list(
  "Fixed (50% true)" = fit1_fixed50,
  "GGM" = fit1_ggm,
  "Glasso" = fit1_glasso,
  "Glasso + Fixed" = fit1_glasso_fixed,
  "GGM + Fixed" = fit1_ggm_fixed
)
times_adj1 <- c(
  "Fixed (50% true)" = as.numeric(t1_fixed50["elapsed"]),
  "GGM" = as.numeric(t1_ggm["elapsed"]),
  "Glasso" = as.numeric(t1_glasso["elapsed"]),
  "Glasso + Fixed" = as.numeric(t1_glasso_fixed["elapsed"]),
  "GGM + Fixed" = as.numeric(t1_ggm_fixed["elapsed"])
)

pip_mat_adj1 <- sapply(fits_adj1, compute_pip)
top_idx_adj1 <- order(rowMeans(pip_mat_adj1), decreasing = TRUE)[1L]
top_name_adj1 <- if (!is.null(colnames(X))) colnames(X)[top_idx_adj1] else paste0("V", top_idx_adj1)

diag_adj1 <- do.call(rbind, lapply(names(fits_adj1), function(mn) {
  fit <- fits_adj1[[mn]]
  pip <- compute_pip(fit)
  beta_chain <- get_beta_matrix(fit)[, top_idx_adj1]
  ms <- model_size_chain(fit)
  ess <- safe_ess_sim(beta_chain)
  n_saved <- length(ms)
  iter_sec <- n_saved / max(as.numeric(times_adj1[mn]), 1e-8)
  data.frame(
    Method = mn,
    Runtime_sec = round(times_adj1[mn], 2),
    N_saved = n_saved,
    Iter_per_sec = round(iter_sec, 2),
    Mean_model_size = round(mean(ms), 2),
    SD_model_size = round(stats::sd(ms), 2),
    Median_model_size = round(stats::median(ms), 2),
    Q05_model_size = round(as.numeric(stats::quantile(ms, 0.05)), 2),
    Q95_model_size = round(as.numeric(stats::quantile(ms, 0.95)), 2),
    Top_gene_for_diag = top_name_adj1,
    PIP_top_gene = round(pip[top_idx_adj1], 3),
    Lag1_ACF_top_beta = round(lag1_acf_sim(beta_chain), 3),
    ESS_top_beta = round(ess, 1),
    ESS_per_sec = round(ess / max(as.numeric(times_adj1[mn]), 1e-8), 3),
    MCSE_top_beta = round(stats::sd(beta_chain) / sqrt(max(1, ess)), 4),
    stringsAsFactors = FALSE
  )
}))

knitr::kable(
  diag_adj1,
  caption = "Simulation 1 MCMC convergence diagnostics and sampling performance for five adjacency-type methods.",
  row.names = FALSE
)
```

```{r sim1-adjtype-trace, fig.height=8, fig.width=10, fig.cap="Simulation 1 adjacency-type comparison: trace plots of the representative top-PIP coefficient."}
par(mfrow = c(2, 3), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (mn in names(fits_adj1)) {
  beta_chain <- get_beta_matrix(fits_adj1[[mn]])[, top_idx_adj1]
  plot(beta_chain,
    type = "l", col = "steelblue",
    main = paste0(mn, ": ", top_name_adj1),
    xlab = "Iteration", ylab = "beta", cex.main = 0.85
  )
}
plot.new()
```

```{r sim1-adjtype-acf, fig.height=8, fig.width=10, fig.cap="Simulation 1 adjacency-type comparison: ACF of the representative top-PIP coefficient."}
par(mfrow = c(2, 3), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (mn in names(fits_adj1)) {
  beta_chain <- get_beta_matrix(fits_adj1[[mn]])[, top_idx_adj1]
  safe_acf_plot(beta_chain,
    lag.max = 50,
    main = paste0(mn, ": ACF(", top_name_adj1, ")"),
    col = "steelblue"
  )
}
plot.new()
```


## Software availability

The **BVS.DAdj** package (version 0.1.0) is available at
[https://github.com/yuy113/BVS.DAdj](https://github.com/yuy113/BVS.DAdj).
It requires R $\geq$ 3.5.0 with **Rcpp**, **RcppArmadillo**, and **Matrix**.
The **huge** package is optional for graphical lasso adjacency estimation.

# References {-}
