---
title: "BVS.DAdj: Bayesian Variable Selection with Dual-Adjacency Network-Informed Ising Priors in R"
author:
  - "Yubing Yao"
  - "Mahlet G. Tadesse"
  - "Raji Balasubramanian"
abstract: >
  We introduce **BVS.DAdj**, an R package for Bayesian variable selection in
  generalized regression models using Markov random field (MRF) priors that
  incorporate dual adjacency structures among predictors. The package provides
  two MCMC interfaces---`bvs_mh()` and `bvs_pg()`---covering six outcome
  types (binary logistic, continuous Gaussian, right-censored time-to-event
  via Cox partial likelihood, overdispersed count via a negative-binomial
  representation, zero-inflated count via a point-mass mixture, and
  imbalanced binary) and six adjacency specifications (fixed, dual-fixed,
  graphical lasso, Bayesian GGM, and combined GGM-plus-fixed). Within
  `bvs_mh()`, Metropolis-Hastings (MH), Hamiltonian Monte Carlo (HMC), and
  No-U-Turn Sampling (NUTS) are available for non-conjugate binary and
  time-to-event parameter blocks. `bvs_pg()` employs Pólya-Gamma data
  augmentation for exact conjugate updates in binary logistic models. Coupling
  parameters in the Ising/MRF prior are updated via the Möller auxiliary
  variable method, which bypasses the intractable normalizing constant. Dense
  ($p \leq n$) and sparse high-dimensional ($p \gg n$) C++ backends are
  implemented via **Rcpp** and **RcppArmadillo**. We describe the statistical
  model, prior specifications, MCMC algorithms, and package API, and
  demonstrate usage through a moderate-dimensional simulation example
  ($n=500$, $p=100$) focused on the `glasso_fixed` method.
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    self_contained: true
    fig_caption: true
    theme: flatly
    highlight: null
    df_print: kable
    mathjax: "file:///opt/homebrew/anaconda3/lib/python3.12/site-packages/spyder/plugins/help/utils/js/mathjax/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    fig_crop: false
    highlight: null
vignette: >
  %\VignetteIndexEntry{BVS.DAdj: R Journal Article}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{=html}
$$
\require{AMSmath}
\require{AMSsymbols}
\newcommand{\N}{\mathcal{N}}
\newcommand{\bfbeta}{\boldsymbol{\beta}}
\newcommand{\bfgamma}{\boldsymbol{\gamma}}
\newcommand{\bftau}{\boldsymbol{\tau}}
\newcommand{\bfOmega}{\boldsymbol{\Omega}}
\newcommand{\calR}{\mathcal{R}}
$$
```

```{=latex}
\newcommand{\N}{\mathcal{N}}
\newcommand{\bfbeta}{\boldsymbol{\beta}}
\newcommand{\bfgamma}{\boldsymbol{\gamma}}
\newcommand{\bftau}{\boldsymbol{\tau}}
\newcommand{\bfOmega}{\boldsymbol{\Omega}}
\newcommand{\calR}{\mathcal{R}}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval    = TRUE,
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction {#sec-intro}

In metabolomics, genomics, and other high-throughput biological studies, a
central question is which of many candidate variables---metabolites, genes, or
other molecular features---are genuinely associated with a disease outcome or
phenotype. Structural relationships among these variables, arising from
biological pathways, co-expression networks, or data-driven graphical models,
carry substantial prior information that standard penalization methods largely
ignore.

Classical penalization approaches including the lasso [@Tib1996], elastic net
[@ZH2005], group lasso [@YL2006], and adaptive lasso [@Zou2006] shrink
regression coefficients to zero but treat predictors independently and may
miss weakly active variables embedded in correlated networks. Bayesian
variable selection with spike-and-slab priors [@GM1993; @GM1997] offers a
principled alternative, quantifying uncertainty through posterior inclusion
probabilities (PIPs). A key extension is to encode predictor-level graph
structure through a Markov random field (MRF) prior on the binary selection
indicators, so that adjacent variables in a known or estimated graph tend to
be co-selected [@LZ2010; @STV2011].

The **BVS.DAdj** package extends this Ising/MRF framework to *two
simultaneous* and *independent* adjacency sources: one based on external
domain knowledge (e.g., a biological pathway graph) and one estimated from
the data themselves (e.g., via Bayesian graphical model selection or
graphical lasso). This dual-adjacency formulation allows the model to borrow
strength from both sources while remaining agnostic about their relative
importance, with coupling strengths $\eta_1$ and $\eta_2$ inferred from data
via the Möller auxiliary variable approach [@MB2006].

**Package contributions.** The **BVS.DAdj** package provides:

1. A unified R interface (`bvs_mh` and `bvs_pg`) dispatching to 12
   optimized C++ backends via **Rcpp**/**RcppArmadillo**.
2. Support for *six* outcome types: binary logistic, continuous Gaussian,
   right-censored time-to-event (Cox partial likelihood), overdispersed
   count (negative-binomial via Poisson-Gamma augmentation), zero-inflated
   count (ZIC), and imbalanced binary.
3. Three within-model samplers for non-conjugate parameter blocks:
   random-walk MH, HMC [@Neal2011], and NUTS [@Hoffman2014].
4. *Six* adjacency specifications covering fixed, data-driven (graphical
   lasso, Bayesian GGM), and dual combinations.
5. Scalable sparse C++ backends for $p \gg n$ with compressed-sparse-column
   (CSC) matrix arithmetic and $O(|\mathcal{E}|)$ neighbourhood sums.
6. An $\eta$ hyperparameter tuning workflow (`phase_transition`,
   `bvs_eta_grid_single`, `bvs_tune_eta`) for choosing Möller reference
   parameters.
7. Post-processing utilities including `summary.bvs`, `plot.bvs`,
   `as.mcmc.bvs`, and `bvs_gelman_diag`.

The remainder of this article is organized as follows.
Section\ 2 describes the statistical model and prior specifications.
Section\ 3 presents the MCMC sampling algorithms.
Section\ 4 describes the package design and API.
Section\ 5 presents Simulation Study 1 (moderate dimensions, $n=500$,
$p=100$).
Section\ 6 concludes with discussion.

# Statistical Model and Prior Specifications {#sec-model}

## Linear predictor and notation

Let $\mathbf{X} \in \mathbb{R}^{n \times p}$ be the predictor matrix, where
column $j$ is the $j$-th candidate variable subject to selection, and let
$\mathbf{Z} \in \mathbb{R}^{n \times q}$ be an optional matrix of
always-included covariates. Define the linear predictor for observation $i$:

$$
\eta_i = \alpha + \mathbf{x}_i^\top \boldsymbol{\beta}_{\boldsymbol{\gamma}}
        + \mathbf{z}_i^\top \boldsymbol{\tau},
\quad i = 1, \ldots, n,
$$

where $\alpha$ is an intercept, $\boldsymbol{\gamma} = (\gamma_1, \ldots,
\gamma_p)^\top \in \{0,1\}^p$ is the binary selection indicator vector,
$\boldsymbol{\beta}_{\boldsymbol{\gamma}}$ is the vector of regression
coefficients with $\beta_j = 0$ imposed whenever $\gamma_j = 0$, and
$\boldsymbol{\tau} \in \mathbb{R}^q$ are the always-included covariate
coefficients.

## Outcome likelihoods {#sec-likelihood}

**Binary logistic outcome** (`outcome_type = "binary"`)
$$
y_i \mid \eta_i \sim \mathrm{Bernoulli}\!\left(\mathrm{logit}^{-1}(\eta_i)\right),
\quad y_i \in \{0, 1\}.
$$

**Continuous Gaussian outcome** (`outcome_type = "continuous"`)
$$
y_i \mid \eta_i, \sigma^2 \sim \mathcal{N}(\eta_i,\, \sigma^2).
$$

**Right-censored time-to-event outcome** (`outcome_type = "TTE"`)

Under the Cox proportional hazard model with Breslow tie handling
[@Cox1972], the partial log-likelihood is

$$
\log L(\boldsymbol{\beta}, \boldsymbol{\tau}) =
\sum_{i:\,\delta_i = 1}
\left[\eta_i - \log\!\left(\sum_{j \in \mathcal{R}(t_i)} e^{\eta_j}\right)\right],
$$

where $\delta_i \in \{0,1\}$ is the event indicator and
$\mathcal{R}(t_i) = \{j : t_j \geq t_i\}$ is the risk set at event time
$t_i$. The intercept $\alpha$ is not identifiable under the partial
likelihood and is fixed at zero for this outcome type.

**Overdispersed count outcome** (`outcome_type = "count"`)

A negative-binomial representation via Gamma mixing:

$$
y_i \mid w_i, \eta_i \sim \mathrm{Poisson}\!\left(w_i\, e^{\eta_i}\right),
\qquad
w_i \sim \mathrm{Gamma}(r,\, r),
$$

where $r > 0$ is the dispersion parameter. Marginalizing over $w_i$ gives
a negative-binomial distribution for $y_i$ with mean $e^{\eta_i}$ and
variance $e^{\eta_i}(1 + e^{\eta_i}/r)$.

**Zero-inflated count outcome** (`outcome_type = "ZIC"`)

Introduces a structural zero component via probability $\pi$:

$$
Z_i \sim \mathrm{Bernoulli}(\pi), \quad
y_i \mid Z_i=1 \sim \delta_{0}(y), \quad
y_i \mid Z_i=0, w_i, \eta_i \sim \mathrm{Poisson}(w_i e^{\eta_i}),
$$

with $w_i \sim \mathrm{Gamma}(r, r)$, $\pi \sim \mathrm{Beta}(a_\pi, b_\pi)$, and optionally $r \sim \mathrm{Gamma}(a_r, b_r)$ [@Lambert1992; @GhoshZINB2006].

**Imbalanced binary outcome** (`outcome_type = "imbalanced_binary"`)

Robust models for imbalanced data, offering either a `logit_t` link (Student-t prior on coefficients via local adaptive variance $\lambda_j \sim \mathrm{IG}(df/2, df/2)$; @Gelman2008) or a `cloglog` link addressing class asymmetry via the complementary log-log link [@KingZeng2001].

## Spike-and-slab prior on regression coefficients {#sec-spikeslab}

Conditional on the selection indicators and the variance $\sigma^2$ (or its
analogue), we use the continuous spike-and-slab formulation:

$$
\beta_j \mid \gamma_j, \sigma^2 \;\sim\;
\gamma_j\,\mathcal{N}(\beta_0,\, \sigma^2)
\;+\;
(1 - \gamma_j)\,\delta_0(\beta_j),
\quad j = 1, \ldots, p.
$$

Always-included covariate coefficients receive an independent normal prior:

$$
\tau_k \mid \sigma^2 \sim \mathcal{N}(\tau_0,\, h_\tau\,\sigma^2),
\quad k = 1, \ldots, q.
$$

The intercept (when estimable) is given:

$$
\alpha \sim \mathcal{N}(\alpha_0,\, h_\alpha\,\sigma^2),
$$

and the residual variance (used in continuous outcomes and as the prior
scale for spike-and-slab) follows:

$$
\sigma^2 \sim \mathrm{Inverse\text{-}Gamma}\!\left(\frac{\nu_0}{2},\;
\frac{\nu_0\,\sigma_0^2}{2}\right).
$$

## Dual-adjacency Ising/MRF prior on selection indicators {#sec-ising}

Let $\mathbf{R}^{(1)}, \mathbf{R}^{(2)} \in \{0,1\}^{p \times p}$ be two
symmetric binary adjacency matrices (zero diagonal) encoding two independent
sources of structural information among the $p$ predictors. The joint prior
on $\boldsymbol{\gamma}$ takes the Ising/MRF form:

$$
p\!\left(\boldsymbol{\gamma} \mid \mu, \eta_1, \eta_2\right)
=
\frac{1}{Z(\mu,\eta_1,\eta_2)}
\exp\!\left(
\mu \sum_{j=1}^p \gamma_j
+ \eta_1 \sum_{j < k} R^{(1)}_{jk}\,\gamma_j\gamma_k
+ \eta_2 \sum_{j < k} R^{(2)}_{jk}\,\gamma_j\gamma_k
\right),
$$

where $\mu \in \mathbb{R}$ is the external field controlling baseline
sparsity and $\eta_1, \eta_2 \geq 0$ are non-negative coupling parameters.
For a predictor with no neighbours in either graph, the marginal prior
inclusion probability is $\mathrm{expit}(\mu)$, so $\mu = \mathrm{logit}(\pi_0)$
for target sparsity $\pi_0$.

The normalizing constant $Z(\mu, \eta_1, \eta_2)$ is intractable for general
$p$ and is handled via the Möller auxiliary variable method (Section\ 3.6).
Single-adjacency models use $\eta_2 = 0$ (and hence $\mathbf{R}^{(2)}$
plays no role); the prior then reduces to the standard single-network Ising
form of @LZ2010 and @STV2011.

## Prior on coupling parameters $\eta_1$, $\eta_2$ {#sec-etaprior}

Each coupling parameter is bounded above by a user-specified maximum
(parameters `eta1_sd` and `eta2_sd`) and follows a scaled Beta distribution:

$$
\eta_\ell = \eta_{\ell,\max}\,u_\ell, \quad
u_\ell \sim \mathrm{Beta}(e_\eta,\, f_\eta), \quad
\ell = 1, 2.
$$

The shape parameters $e_\eta$ and $f_\eta$ and the upper bound
$\eta_{\ell,\max}$ are chosen to keep the prior mass below the phase
transition of the Ising model on the given adjacency graph; the
`phase_transition()` helper function profiles this transition before fitting.

## Estimating unknown adjacency via Bayesian GGM {#sec-ggm}

When an adjacency matrix is unknown, the package optionally estimates it
jointly with the regression model by placing a continuous spike-and-slab SSVS
prior on the precision matrix $\boldsymbol{\Omega}$ of the predictors
$\mathbf{X}$, following the Bayesian GGM approach of @WH2015:

$$
p(\boldsymbol{\Omega} \mid \mathbf{R}^{\mathrm{ggm}}, \boldsymbol{\theta})
\propto
\prod_{i < j}
\mathcal{N}\!\left(\omega_{ij} \mid 0,\, v^2_{r^{\mathrm{ggm}}_{ij}}\right)
\prod_i \mathrm{Exp}\!\left(\omega_{ii} \mid \tfrac{\lambda}{2}\right)
\cdot \mathbf{1}\!\left(\boldsymbol{\Omega} \succ 0\right),
$$

where $v_{r^{\mathrm{ggm}}_{ij}} = v_0$ (spike) when $r^{\mathrm{ggm}}_{ij}=0$
(no edge) and $v_1 \gg v_0$ (slab) when $r^{\mathrm{ggm}}_{ij}=1$ (edge
present). The binary edge indicator $r^{\mathrm{ggm}}_{ij}$ is updated via
Bayes' theorem at each MCMC iteration:

$$
\Pr\!\left(r^{\mathrm{ggm}}_{ij} = 1 \mid \omega_{ij}\right)
=
\frac{\pi\,\mathcal{N}(\omega_{ij} \mid 0, v_1^2)}
     {\pi\,\mathcal{N}(\omega_{ij} \mid 0, v_1^2)
      + (1-\pi)\,\mathcal{N}(\omega_{ij} \mid 0, v_0^2)},
$$

where $\pi$ is the prior edge inclusion probability (default
$4 / (p-1)$, targeting sparse graphs). The precision matrix is updated by
the block Gibbs column sweep of @WH2015; in the sparse backend this is
performed in CSC format.

# MCMC Sampling Algorithms {#sec-mcmc}

## Per-iteration structure

Each iteration of either sampler updates in the following block order:

| Step | Update | Both / MH only / PG only |
|:-----|:-------|:------------------------:|
| 1 | Binary selection indicators $\gamma_j$ (add/delete MH move) | both |
| 2 | Active regression coefficients $\boldsymbol{\beta}_\gamma$ | both |
| 3 | Intercept $\alpha$ and covariate coefficients $\boldsymbol{\tau}$ | both |
| 4 | Residual variance $\sigma^2$ | both |
| 5 | Coupling parameters $\eta_1$ (and $\eta_2$ in dual models) | both |
| 6 | GGM precision matrix $\boldsymbol{\Omega}$ (column sweep) | when `adj_type` $\in$ `{"ggm","ggm_fixed"}` |
| 7 | PG latent variables $\omega_i$ | PG only |

In `bvs_mh()`, Steps\ 1–2 are optionally repeated `n_thin_gb` times per
stored sample (inner gamma/beta thinning), improving mixing of the discrete
selection indicators without inflating output storage.

## Metropolis-Hastings sampler (`alg_type = "MH"`) {#sec-mh}

**Selection indicator update (Step\ 1).** For each candidate variable $j$,
a flip $\gamma_j^* = 1 - \gamma_j$ is proposed. If activating the variable
($\gamma_j = 0 \to 1$), a proposal $\beta_j^* \sim \mathcal{N}(\beta_0, \sigma^2)$
is drawn from the slab; if deactivating ($\gamma_j = 1 \to 0$), $\beta_j^*$ is
set to zero. The move is accepted with probability
$\min\{1, \exp(\Delta)\}$ where

$$
\Delta = \ell(\boldsymbol{\beta}^*, \boldsymbol{\gamma}^*) -
         \ell(\boldsymbol{\beta},  \boldsymbol{\gamma})
       + \log p(\boldsymbol{\gamma}^* \mid \mu, \eta_1, \eta_2)
       - \log p(\boldsymbol{\gamma}  \mid \mu, \eta_1, \eta_2)
       + \log p(\beta_j^*) - \log p(\beta_j).
$$

**Regression coefficient update (Step\ 2).**

- *Binary / count outcomes*: random-walk MH perturbation of the active
  $\boldsymbol{\beta}_\gamma$ block.
- *Continuous outcome*: exact conjugate Gibbs draw for active $\boldsymbol{\beta}_\gamma$
  from the Gaussian conditional posterior.
- *TTE outcome*: Fisher-scoring random-walk MH perturbation of the active
  block.

All MH acceptance steps use the numerically robust helper
`safe_mh_accept(log_ratio)` which avoids the classic `log(runif(0,1)) < log_ratio`
deadlock at $-\infty$.

## Hamiltonian Monte Carlo (`alg_type = "HMC"`) {#sec-hmc}

For binary and TTE outcomes, `bvs_mh(..., alg_type = "HMC")` replaces the
random-walk MH update on the active parameter block with Hamiltonian
dynamics [@Neal2011]. Let $\mathbf{q}$ denote the active parameter vector
$(\boldsymbol{\beta}_\gamma, \alpha, \boldsymbol{\tau})$ (with $\alpha$
excluded for TTE) and $U(\mathbf{q}) = -\log\pi(\mathbf{q} \mid \cdot)$
the potential energy. With auxiliary momentum $\mathbf{p} \sim \mathcal{N}(\mathbf{0}, \mathbf{M})$
(mass matrix $\mathbf{M}$), the Hamiltonian is

$$
H(\mathbf{q}, \mathbf{p})
= U(\mathbf{q}) + \tfrac{1}{2}\mathbf{p}^\top \mathbf{M}^{-1} \mathbf{p}.
$$

Each HMC sub-step:

1. Sample momentum $\mathbf{p} \sim \mathcal{N}(\mathbf{0}, \mathbf{M})$.
2. Run $L$ leapfrog steps of size $\epsilon$:
   $\mathbf{p} \leftarrow \mathbf{p} - \frac{\epsilon}{2}\nabla_{\mathbf{q}} U(\mathbf{q})$;
   $\mathbf{q} \leftarrow \mathbf{q} + \epsilon\, \mathbf{M}^{-1}\mathbf{p}$;
   $\mathbf{p} \leftarrow \mathbf{p} - \frac{\epsilon}{2}\nabla_{\mathbf{q}} U(\mathbf{q})$.
3. Accept/reject: $\alpha_{\mathrm{HMC}} = \min\{1, \exp[-H(\mathbf{q}', \mathbf{p}') + H(\mathbf{q}, \mathbf{p})]\}$.

The gradient $\nabla_{\mathbf{q}} U$ is computed analytically for each
outcome type. For binary logistic:
$\nabla_{\beta_j} U = -\sum_i x_{ij}(y_i - \hat{\pi}_i) + \beta_j/\sigma^2$
where $\hat{\pi}_i = \mathrm{logit}^{-1}(\eta_i)$.
For Cox TTE:
$\nabla_{\beta_j} U = -\sum_{i:\delta_i=1} x_{ij} + \sum_{i:\delta_i=1} \bar{x}_j(t_i) + \beta_j/\sigma^2$
where $\bar{x}_j(t_i) = \sum_{k\in\mathcal{R}(t_i)} x_{kj} e^{\eta_k} / \sum_{k\in\mathcal{R}(t_i)} e^{\eta_k}$.

## No-U-Turn Sampler (`alg_type = "NUTS"`) {#sec-nuts}

NUTS [@Hoffman2014] removes the need to specify the number of leapfrog steps
$L$ by adaptively growing a balanced binary tree of trajectory
states until a U-turn is detected:

$$
({\mathbf{q}}^+ - {\mathbf{q}}^-)^\top \mathbf{M}^{-1} {\mathbf{p}}^- < 0
\quad \text{or} \quad
({\mathbf{q}}^+ - {\mathbf{q}}^-)^\top \mathbf{M}^{-1} {\mathbf{p}}^+ < 0,
$$

where $\mathbf{q}^-$, $\mathbf{q}^+$ are the leftmost/rightmost tree
positions and $\mathbf{p}^-$, $\mathbf{p}^+$ the corresponding momenta.
A proposal is drawn uniformly from the built trajectory with detailed
balance, and the step size $\epsilon$ is tuned during burn-in via dual
averaging. The maximum tree depth is controlled by `nuts_max_treedepth`
(default 10).

Compared with fixed-step HMC, NUTS reduces sensitivity to $L$ specification
and often achieves higher effective sample size per second in strongly
correlated parameter spaces.

## Pólya-Gamma augmented Gibbs (`bvs_pg`) {#sec-pg}

For binary logistic outcomes, `bvs_pg()` introduces Pólya-Gamma latent
variables $\omega_i \sim \mathrm{PG}(1, \eta_i)$ [@PG2013]. Conditional
on $\boldsymbol{\omega} = (\omega_1, \ldots, \omega_n)^\top$, the logistic
likelihood becomes conjugate Gaussian in $(\alpha, \boldsymbol{\beta}_\gamma)$,
enabling the exact Gibbs draw:

$$
\begin{pmatrix}\alpha \\ \boldsymbol{\beta}_\gamma\end{pmatrix}
\;\Bigg|\; \boldsymbol{\omega}, \boldsymbol{\gamma}, \sigma^2
\;\sim\;
\mathcal{N}\!\left(\mathbf{V}_\gamma \mathbf{m}_\gamma,\; \mathbf{V}_\gamma\right),
$$

where
$\mathbf{V}_\gamma =
  \left(\tilde{\mathbf{X}}_\gamma^\top \mathrm{diag}(\boldsymbol{\omega})\,
        \tilde{\mathbf{X}}_\gamma + \sigma^{-2}\mathbf{I}\right)^{-1}$,
$\mathbf{m}_\gamma = \tilde{\mathbf{X}}_\gamma^\top
  \boldsymbol{\kappa}$,
$\kappa_i = y_i - 1/2$,
and $\tilde{\mathbf{X}}_\gamma = [\mathbf{1}, \mathbf{X}_\gamma]$ includes
the intercept column. PG draws are obtained via the truncated-series
approximation of @PG2013 implemented in C++.

## Möller auxiliary variable update for $\eta_1$, $\eta_2$ {#sec-moller}

The normalizing constant $Z(\mu, \eta_1, \eta_2)$ of the Ising prior is
intractable for general $p$. Following @MB2006, we introduce an auxiliary
variable $\boldsymbol{\omega}^*$ drawn via Propp-Wilson perfect simulation
from the Ising model at fixed reference parameters
$(\tilde{\mu}, \tilde{\eta}_1, \tilde{\eta}_2)$. The resulting MH ratio
for a proposal $\eta_\ell' \sim q(\eta_\ell' \mid \eta_\ell)$ is:

$$
\alpha_{\mathrm{Moller}} = \min\!\left\{1,\;
\frac{p(\boldsymbol{\gamma} \mid \mu, \eta_1', \eta_2')\,
      p(\eta_\ell')\,
      f(\boldsymbol{\omega}^* \mid \tilde{\mu}, \tilde{\eta}_1', \tilde{\eta}_2')}
     {p(\boldsymbol{\gamma} \mid \mu, \eta_1,  \eta_2)\,
      p(\eta_\ell)\,
      f(\boldsymbol{\omega}^* \mid \tilde{\mu}, \tilde{\eta}_1,  \tilde{\eta}_2)}\right\},
$$

where the intractable $Z$ cancels exactly because the same auxiliary sample
$\boldsymbol{\omega}^*$ is used in numerator and denominator. Propp-Wilson
simulation terminates in finite time and the maximum doubling horizon is
controlled by `Tmax` (default 64).

## Sparse high-dimensional backend {#sec-sparse}

For $p \gg n$, the package stores the scatter matrix
$\mathbf{S} = \mathbf{X}^\top\mathbf{X}$ in compressed sparse column (CSC)
format (via `prepare_sparse_S()`), thresholding near-zero entries for
storage efficiency. Ising neighbourhood sums
$\sum_k R^{(1)}_{jk} \gamma_k$ are computed in $O(\deg_j)$ time, where
$\deg_j$ is the degree of variable $j$ in the adjacency graph. The GGM
column sweep updates one precision-matrix column at a time using only the
non-zero entries of the current column of $\mathbf{S}$. The selection
indicator update cycles through `n_mh_gamma` randomly chosen predictors per
outer iteration to reduce per-iteration cost.

## MCMC mixing and convergence improvements {#sec-mcmc-improvements}

Three improvements are implemented uniformly across all 12 C++ backends and the R
post-processing layer.

### Locally-balanced $\gamma$ proposals {#sec-lb-gamma}

Uniform-coordinate random-walk proposals for the binary selection indicators
$\gamma$ can mix slowly when most proposed flips have low acceptance
probability. Following @Zanella2020, the package replaces uniform coordinate
selection with a *locally balanced* distribution. For each candidate variable
$j$, the unnormalised weight

$$
w_j = \sqrt{ \pi(\boldsymbol{\gamma}^{(j)}) / \pi(\boldsymbol{\gamma}) }
     = \exp\!\bigl( 0.5 \cdot \mathrm{clip}_{[-60,60]}\bigl[
       (1 - 2\gamma_j)(\mu + \eta_1 N_j^{(1)} + \eta_2 N_j^{(2)}) \bigr] \bigr),
$$

where $N_j^{(\ell)} = \sum_k R^{(\ell)}_{jk}\gamma_k$ is the active-neighbour count
in adjacency $\ell$ and $\mathrm{clip}$ prevents numerical overflow.
The balancing function $g(t) = \sqrt{t}$ corresponds to the geometric
mean of the target ratios, yielding the "square-root" or Barker kernel in
the sense of @Zanella2020. Coordinate $j$ is selected with probability
proportional to $w_j$, and the MH ratio is corrected by the log
proposal ratio $\log(q_\mathrm{rev}/q_\mathrm{fwd})$.

Activated via `use_lb_gamma = TRUE` (default). Supported by all 12 backends.

**Reference:** @Zanella2020.

### Robust Adaptive Metropolis for $\eta$ {#sec-ram}

The Möller MH proposal SD for the coupling parameters $\eta_1$, $\eta_2$
is adapted online using the Robust Adaptive Metropolis (RAM) algorithm
of @Vihola2012. Letting $\log\sigma_n$ denote the log-scale proposal
standard deviation after $n$ adaptations, the Robbins-Monro update is

$$
\log\sigma_{n+1} = \log\sigma_n + \tfrac{1}{2} \min(1, n^{-2/3})
                   \bigl(\hat{\alpha}_n - \alpha^*\bigr),
$$

where $\hat{\alpha}_n \in \{0,1\}$ is the accept-indicator for iteration $n$
and $\alpha^* = 0.44$ is the optimal 1D acceptance rate
[@Roberts1997OptimalScaling]. The $n^{-2/3}$ diminishing step satisfies
the diminishing-adaptation condition (Theorem 2 of @Vihola2012), ensuring
ergodicity.

An `EtaAdapter` struct in `BayesLogit_Sparse_Helpers.h` encapsulates the
log-sigma state and the update rule; each backend declares one adapter per
$\eta$ parameter.

**References:** @Vihola2012; @AndrieuThoms2008.

### Rank-normalised $\hat{R}$ and bulk/tail ESS {#sec-rhat}

Classical Gelman-Rubin $\hat{R}$ and `effectiveSize` (ESS) assume approximately
Gaussian chains. @Vehtari2021 propose three improvements:

1. **Split chains**: each chain is split in half before computing $\hat{R}$,
   detecting within-chain non-stationarity.
2. **Rank normalisation**: draws are replaced by their normal quantiles,
   making the diagnostic robust to heavy tails.
3. **Bulk and tail ESS**: bulk ESS (via Geyer's initial positive sequence
   estimator) measures efficiency for the body of the distribution; tail ESS
   = $\min(\mathrm{ESS}_{0.05}, \mathrm{ESS}_{0.95})$ measures tail efficiency.

These diagnostics are available as exported R functions:

```{r diag-functions, eval=FALSE}
bvs_rhat(x)       # rank-normalised split-Rhat; target < 1.01
bvs_ess_bulk(x)   # bulk ESS
bvs_ess_tail(x)   # tail ESS = min(ESS_q=0.05, ESS_q=0.95)
```

`x` can be a numeric vector (single chain, split in half automatically) or
a matrix with one column per chain. When the `posterior` package is
installed, these functions delegate to `posterior::rhat()`,
`posterior::ess_bulk()`, and `posterior::ess_tail()`; otherwise a pure-R
fallback implementing the same algorithm is used.
`summary.bvs()` reports `ESS` (bulk) and `ESS_tail` for each predictor.

**Reference:** @Vehtari2021.

# Package Design and API {#sec-api}

## Installation

```{r install, eval=FALSE}
# From GitHub:
devtools::install_github("yuy113/BVS.DAdj")
```

```{r load-pkg}
library(BVS.DAdj)
```

## User-facing functions

**BVS.DAdj** exports two primary sampling functions that automatically
dispatch to 12 C++ backends based on `sampler × adj_type × sparse`:

| Function   | Sampler | Outcome types | Notes |
|:-----------|:--------|:--------------|:------|
| `bvs_mh()` | MH / HMC / NUTS | binary, continuous, TTE, count | HMC/NUTS for binary & TTE |
| `bvs_pg()` | Pólya-Gamma Gibbs | binary only | Exact beta/alpha updates |

## Adjacency types {#sec-adj}

| `adj_type` | Adjacency source(s) | Free $\eta$ | C++ backends |
|:-----------|:--------------------|:-----------:|:-------------|
| `"fixed"` | One user-supplied binary matrix `adj_fixed` | $\eta_1$ | MH + PG, dense |
| `"dual_fixed"` | Two user-supplied matrices (`adj_fixed`, `adj_fixed2`) | $\eta_1, \eta_2$ | MH + PG, dense |
| `"glasso"` | Graphical lasso from $\mathbf{X}$ (EBIC or RIC) | $\eta_1$ | MH + PG, dense |
| `glasso_fixed` | Graphical lasso + one fixed matrix | $\eta_1, \eta_2$ | MH + PG, dense |
| `"ggm"` | Bayesian GGM SSVS, updated at every iteration | $\eta_1$ | MH + PG, dense/sparse |
| `"ggm_fixed"` | Bayesian GGM + one fixed matrix | $\eta_1, \eta_2$ | MH + PG, dense/sparse |

For `glasso` and `glasso_fixed`, the adjacency is estimated once
before MCMC using the **huge** package [@FHT2008] and held fixed
throughout. For `"ggm"` and `"ggm_fixed"`, the adjacency is updated at
every MCMC iteration via the GGM column sweep.

## Core function signatures

```{r api-mh, eval=FALSE}
bvs_mh(
  X, y,
  event        = NULL,            # event indicator for TTE (1=event, 0=censored)
  outcome_type = "binary",        # "binary" | "continuous" | "TTE" | "count"
  adj_type     = "fixed",         # see adjacency type table above
  adj_fixed    = NULL,            # p x p binary adjacency matrix
  adj_fixed2   = NULL,            # second p x p matrix (dual_fixed only)
  sparse       = FALSE,           # TRUE for sparse GGM backend (p >> n)
  S_ggm        = NULL,            # pre-computed CSC scatter matrix
  niter        = 60000L, burnin = 10000L, thin = 1L,
  n_thin_gb    = 3L,              # inner gamma/beta sub-iterations (dense)
  n_mh_gamma   = 3L,              # gamma flip proposals per iteration (sparse)
  # -- Sparsity / variance priors --
  mu       = -2.197,              # logit(pi_0); default pi_0 = 10%
  nu0      = 2,   sigmasq0 = 1.5,
  h        = 1.5, alpha0 = 0, beta0 = 0,
  # -- Eta / Moller parameters --
  eta1_sd  = 0.5, eta2_sd  = 0.5,
  mu_tilde = -4,  eta1_tilde = 0.5, eta2_tilde = 0.5,
  e_eta    = 1,   f_eta = 1, Tmax = 64L,
  # -- GGM SSVS priors --
  v0_ggm = 0.015^2, v1_ggm = NULL, pii_ggm = NULL, lambda_ggm = 1,
  # -- Always-included covariates --
  z_dat = NULL, tau0 = 0, htau = 1.5, tau_init = NULL,
  # -- HMC/NUTS --
  alg_type = "MH", hmc_step_size = 0.1,
  hmc_n_leapfrog = 10L, nuts_max_treedepth = 10L,
  # -- Storage (sparse mode) --
  store_beta = FALSE, store_gamma = FALSE,
  store_Z_pip = TRUE, store_Z_list = FALSE,
  # -- Initialisation --
  beta_init = NULL, gamma_init = NULL, alpha_init = NULL
)
```

```{r api-pg, eval=FALSE}
bvs_pg(
  X, y,
  adj_type  = "fixed",
  adj_fixed = NULL, adj_fixed2 = NULL,
  sparse    = FALSE, S_ggm = NULL,
  niter = 60000L, burnin = 10000L, thin = 1L,
  n_mh_gamma = 3L,
  mu = -2.197, nu0 = 2, sigmasq0 = 1.5, h = 1.5,
  alpha0 = 0, beta0 = 0,
  eta1_sd = 0.5, eta2_sd = 0.5,
  mu_tilde = -4, eta1_tilde = 0.5, eta2_tilde = 0.5,
  e_eta = 1, f_eta = 1, Tmax = 64L,
  v0_ggm = 0.015^2, v1_ggm = NULL, pii_ggm = NULL, lambda_ggm = 1,
  z_dat = NULL, tau0 = 0, htau = 1.5, tau_init = NULL,
  store_beta = FALSE, store_gamma = FALSE,
  store_Z_pip = TRUE, store_Z_list = FALSE,
  beta_init = NULL, gamma_init = NULL, alpha_init = NULL
)
```

## Key hyperparameter guide

| Parameter | Default | Interpretation |
|:----------|:--------|:---------------|
| `mu` | $\mathrm{logit}(0.10) \approx -2.197$ | Baseline log-odds of inclusion; set to $\mathrm{logit}(\pi_0)$ for target sparsity $\pi_0$ |
| `eta1_sd`, `eta2_sd` | `0.5` | Upper bound of $\eta_\ell$; keep below phase-transition critical value |
| `mu_tilde`, `eta1_tilde`, `eta2_tilde` | `-4`, `0.5`, `0.5` | Möller reference parameters; should correspond to a sparser reference model |
| `e_eta`, `f_eta` | `1`, `1` | Beta shape for $u_\ell \sim \mathrm{Beta}(e,f)$; increase `f_eta` to shrink $\eta_\ell$ toward zero |
| `n_thin_gb` | `3` | Inner gamma/beta sub-iterations per stored sample (dense MH only) |
| `v0_ggm` | `0.015^2` | GGM spike variance (small = strong near-zero shrinkage for non-edges) |
| `pii_ggm` | `4/(p-1)` | GGM prior edge inclusion probability (sparse default) |

## Return value

Both functions return an S3 object of class `"bvs"`:

| Field | Description |
|:------|:------------|
| `gamma_pip` | Length-$p$ vector of posterior inclusion probabilities |
| `beta` | $N_{\text{saved}} \times p$ matrix (dense) or sparse list (sparse) of $\beta$ draws |
| `gamma` | $N_{\text{saved}} \times p$ matrix (dense) or sparse index list of $\gamma$ draws |
| `alpha` | Length-$N_{\text{saved}}$ intercept chain |
| `sigmasq` | Length-$N_{\text{saved}}$ variance chain |
| `eta1`, `eta2` | Coupling parameter chains |
| `Z_pip` | GGM edge PIP matrix (when `store_Z_pip = TRUE`) |
| `sampler`, `adj_type`, `outcome_type` | Metadata strings |
| `p`, `n`, `niter`, `burnin` | Dimension / MCMC metadata |

## Post-processing utilities

```{r postproc, eval=FALSE}
# Summarise: PIPs, posterior means, credible intervals, HPD, ESS
s <- summary(fit, pip_threshold = 0.5, cred_level = 0.95)
cat("Selected variables:", s$selected, "\n")
cat("Model size:", s$nselected, "\n")
print(s$summary_beta[s$selected, ])

# Plot: PIP bar chart, trace plots, ACF
plot(fit, type = "pip", pip_threshold = 0.5)
plot(fit, type = "trace", top_n = 6)
plot(fit, type = "acf",   vars = c(1:5))

# Convert to coda mcmc for diagnostics
library(coda)
mc <- as.mcmc.bvs(fit, vars = 1:20)
effectiveSize(mc)

# Gelman-Rubin PSRF across two independent chains
diag <- bvs_gelman_diag(list(fit1, fit2), vars = 1:20)
print(diag$psrf[1:10, ])   # per-variable PSRF (target < 1.1)
```

## Eta hyperparameter tuning workflow

```{r eta-tuning, eval=FALSE}
# Stage 1: profile the phase transition of the Ising prior
pt <- phase_transition(R,
        mu = -2.197, min_eta = 0, max_eta = 1.5,
        step_size = 0.05, num_rep = 10, Tmax = 64L)
matplot(pt, type = "l", xlab = "eta index", ylab = "model size",
        main = "Ising phase transition")

# Stage 2: build candidate hyperparameter grid
grid <- bvs_eta_grid_single(
  adj       = R,
  eta1_frac = seq(0.2, 0.8, by = 0.2),
  mu_tilde  = c(-4, -5),
  eta1_tilde = c(0.05, 0.10)
)

# Stage 3: select by FDR-controlled PIP threshold
best <- bvs_tune_eta_fixed_pip(
  X, y,
  adj_type    = "fixed", adj_fixed = R,
  sampler     = "pg",
  eta_grid    = grid,
  niter       = 3000L, burnin = 500L,
  pip_threshold_fixed = 0.5,
  target_fdr  = 0.05
)
cat("Best eta1_sd:", best$best_row$eta1_sd, "\n")
```

# Simulation Study 1: Moderate Dimensions ($n = 500$, $p = 100$) {#sec-sim1}

## Study Design {#sec-sim1-design}

We evaluate BVS-DAdj under a moderate-dimensional binary logistic regression
setting with $n = 500$ observations, $p = 100$ covariates, and $p^* = 30$
truly active predictors. The active set is partitioned into two signal groups:
10 weak-signal predictors ($j = 1, \ldots, 10$) with coefficients drawn from
$\mathrm{Uniform}(0.4, 0.6)$ and 20 strong-signal predictors
($j = 11, \ldots, 30$) with coefficients from
$\mathrm{Uniform}(1.05, 1.15)$, all with alternating signs.
The remaining $p - p^* = 70$ predictors are null ($\beta_j = 0$).

The covariate matrix $\mathbf{X}$ is generated from a multivariate normal
distribution with block-equicorrelation structure: predictors $1$ through $30$
share pairwise correlation $\rho = 0.35$, while predictors $31$ through $100$
are independent. The binary response is generated as
$y_i \sim \mathrm{Bernoulli}(\mathrm{logistic}(\mathbf{x}_i^\top \boldsymbol{\beta}))$.

```{r sim1-data}
set.seed(2024)
n      <- 500L
p      <- 100L
p_star <- 30L

# Block covariance for active predictors
Sigma <- diag(p)
Sigma[1:p_star, 1:p_star] <- 0.35
diag(Sigma) <- 1.0

X <- MASS::mvrnorm(n, mu = rep(0, p), Sigma = Sigma)

beta_true <- rep(0, p)
weak_mag   <- runif(10L, min = 0.4, max = 0.6)
strong_mag <- runif(20L, min = 1.05, max = 1.15)
sgn        <- rep(c(1, -1), length.out = p_star)

beta_true[1:10]   <- sgn[1:10]   * weak_mag
beta_true[11:30]  <- sgn[11:30]  * strong_mag

linpred <- X %*% beta_true
y <- rbinom(n, size = 1L, prob = plogis(linpred))

cat(sprintf("Cases: %d  Controls: %d  (%.1f%% cases)\n",
            sum(y), sum(1 - y), 100 * mean(y)))
```

## Adjacency construction {#sec-sim1-adj}

For this vignette version, we keep only the `glasso_fixed` method.
The fixed external adjacency is formed by randomly retaining 50% of true
active-block edges and is then combined with a glasso-estimated adjacency
inside `adj_type = "glasso_fixed"`.

```{r sim1-adj}
# True active-block adjacency
R_true1 <- matrix(0L, p, p)
R_true1[1:p_star, 1:p_star] <- 1L
diag(R_true1) <- 0L

# External fixed adjacency: random 50% of true edges
set.seed(202401)
true_pairs <- which(upper.tri(R_true1) & R_true1 == 1L, arr.ind = TRUE)
keep_n     <- ceiling(0.5 * nrow(true_pairs))
keep_pairs <- true_pairs[sample(nrow(true_pairs), keep_n), , drop = FALSE]

R_ext50 <- matrix(0L, p, p)
R_ext50[cbind(keep_pairs[, 1], keep_pairs[, 2])] <- 1L
R_ext50[cbind(keep_pairs[, 2], keep_pairs[, 1])] <- 1L
diag(R_ext50) <- 0L

cat("True edges:", sum(R_true1) / 2,
    "  External-50% edges:", sum(R_ext50) / 2, "\n")
```

## MCMC analysis (glasso\_fixed only) {#sec-sim1-run}

For illustrative runtime only, we set `niter = 10000` and `runin = 5000`.
All other hyperparameters are kept the same as before.

```{r sim1-run}
# Hyperparameters (kept unchanged except niter/runin)
MU    <- qlogis(0.01)
ETA1  <- 0.08
ETA2  <- 0.06
E_ETA <- 1.0
F_ETA <- 25.0
MU_T  <- -7.0
E1T   <- 0.005
E2T   <- 0.004

niter_s1 <- 10000L
runin_s1 <- 5000L
thin_s1  <- 1L

set.seed(2024)
fit_glasso_fixed_1 <- bvs_mh(
  X, y,
  adj_type = "glasso_fixed",
  adj_fixed = R_ext50,
  glasso_criterion = "ebic",
  niter = niter_s1,
  burnin = runin_s1,
  thin = thin_s1,
  mu = MU,
  eta1_sd = ETA1,
  eta2_sd = ETA2,
  mu_tilde = MU_T,
  eta1_tilde = E1T,
  eta2_tilde = E2T,
  e_eta = E_ETA,
  f_eta = F_ETA,
  n_thin_gb = 3L
)

set.seed(9999)
fit_glasso_fixed_2 <- bvs_mh(
  X, y,
  adj_type = "glasso_fixed",
  adj_fixed = R_ext50,
  glasso_criterion = "ebic",
  niter = niter_s1,
  burnin = runin_s1,
  thin = thin_s1,
  mu = MU,
  eta1_sd = ETA1,
  eta2_sd = ETA2,
  mu_tilde = MU_T,
  eta1_tilde = E1T,
  eta2_tilde = E2T,
  e_eta = E_ETA,
  f_eta = F_ETA,
  n_thin_gb = 3L
)
```

## Results {#sec-sim1-results}

We summarize posterior inclusion probabilities (PIPs), selected predictors
(PIP $> 0.5$), and selection metrics including TPR, FPR, F1, FDR,
precision, and model size.

```{r sim1-results}
compute_pip <- function(fit) {
  if (!is.null(fit$gamma_pip)) return(as.numeric(fit$gamma_pip))
  if (is.matrix(fit$gamma))    return(colMeans(fit$gamma))
  stop("No gamma samples available.")
}

compute_metrics <- function(pip, threshold = 0.5, p_star = 30L) {
  sel <- which(pip > threshold)
  tp  <- sum(sel <= p_star)
  fp  <- sum(sel > p_star)
  fn  <- p_star - tp
  tn  <- (length(pip) - p_star) - fp

  precision <- if ((tp + fp) > 0) tp / (tp + fp) else 0
  tpr       <- if ((tp + fn) > 0) tp / (tp + fn) else 0
  fpr       <- if ((fp + tn) > 0) fp / (fp + tn) else 0
  fdr       <- if ((tp + fp) > 0) fp / (tp + fp) else 0
  f1        <- if ((2 * tp + fp + fn) > 0) (2 * tp) / (2 * tp + fp + fn) else 0

  data.frame(
    TP = tp,
    FP = fp,
    FN = fn,
    TN = tn,
    TPR = round(tpr, 3),
    FPR = round(fpr, 3),
    Precision = round(precision, 3),
    FDR = round(fdr, 3),
    F1 = round(f1, 3),
    N_selected = length(sel)
  )
}

pip_s1 <- compute_pip(fit_glasso_fixed_1)
selected_idx <- which(pip_s1 > 0.5)

selected_tab <- data.frame(
  Predictor = selected_idx,
  PIP = round(pip_s1[selected_idx], 3)
)

if (nrow(selected_tab) == 0L) {
  cat("No predictors selected at PIP > 0.5\n")
} else {
  knitr::kable(
    selected_tab,
    row.names = FALSE,
    caption = "Predictors selected by glasso-fixed with PIP > 0.5 (seed 2024 chain)."
  )
}

metrics_s1 <- compute_metrics(pip_s1, threshold = 0.5, p_star = p_star)
knitr::kable(
  metrics_s1,
  row.names = FALSE,
  caption = "Simulation 1 (glasso-fixed): variable-selection metrics at PIP threshold 0.5."
)
```

```{r sim1-pip-plot, fig.height=4.8, fig.width=6.4, out.width="95%", fig.align='center', fig.cap="Simulation 1 (glasso-fixed): PIPs for all 100 predictors. Red bars are true active predictors (1--30); grey bars are null predictors; dashed line is PIP = 0.5."}
idx <- 1:p
bar_col <- ifelse(idx <= p_star, "firebrick", "grey70")
op <- par(no.readonly = TRUE)
on.exit(par(op), add = TRUE)
par(mar = c(4.2, 4.2, 2.2, 1.0))
barplot(
  pip_s1[idx],
  col = bar_col,
  border = NA,
  ylim = c(0, 1),
  ylab = "PIP",
  xlab = "Predictor index",
  main = "glasso-fixed PIP profile"
)
abline(h = 0.5, lty = 2)
```

## MCMC Convergence Diagnostics {#sec-sim1-diag}

We report trace plot, ACF plot, Gelman--Rubin $\hat{R}$, and effective sample
size (ESS) using two independent chains.

```{r sim1-trace, fig.height=4.2, fig.width=6.4, out.width="90%", fig.align='center', fig.cap="Trace plot of the first active coefficient $\\beta_1$ for glasso-fixed (seed 2024 chain)."}
beta1_chain <- as.numeric(fit_glasso_fixed_1$beta[, 1L])
op <- par(no.readonly = TRUE)
on.exit(par(op), add = TRUE)
par(mar = c(4.2, 4.2, 2.2, 1.0))
plot(
  beta1_chain,
  type = "l",
  col = "steelblue4",
  xlab = "Iteration (post-runin)",
  ylab = expression(beta[1]),
  main = expression("Trace plot for " * beta[1])
)
```

```{r sim1-acf, fig.height=4.2, fig.width=6.4, out.width="90%", fig.align='center', fig.cap="ACF plot of $\\beta_1$ for glasso-fixed (seed 2024 chain)."}
acf(beta1_chain, main = expression("ACF for " * beta[1]))
```

```{r sim1-rhat-ess}
beta_chain_1 <- as.matrix(fit_glasso_fixed_1$beta)
beta_chain_2 <- as.matrix(fit_glasso_fixed_2$beta)

n_keep <- min(nrow(beta_chain_1), nrow(beta_chain_2))
mcmc_1 <- coda::mcmc(beta_chain_1[1:n_keep, , drop = FALSE])
mcmc_2 <- coda::mcmc(beta_chain_2[1:n_keep, , drop = FALSE])
mcmc_list <- coda::mcmc.list(mcmc_1, mcmc_2)

rhat_vec <- coda::gelman.diag(mcmc_list, autoburnin = FALSE, multivariate = FALSE)$psrf[, "Point est."]
ess_vec  <- as.numeric(coda::effectiveSize(mcmc_list))

diag_tab <- data.frame(
  Predictor = 1:p,
  Rhat = round(rhat_vec, 3),
  ESS = round(ess_vec, 1)
)

knitr::kable(
  diag_tab[1:20, ],
  row.names = FALSE,
  caption = "First 20 predictors: R-hat and effective sample size (ESS) from two chains."
)

diag_summary <- data.frame(
  Metric = c("Max Rhat", "Median Rhat", "Min ESS", "Median ESS"),
  Value = c(
    round(max(rhat_vec, na.rm = TRUE), 3),
    round(median(rhat_vec, na.rm = TRUE), 3),
    round(min(ess_vec, na.rm = TRUE), 1),
    round(median(ess_vec, na.rm = TRUE), 1)
  )
)

knitr::kable(
  diag_summary,
  row.names = FALSE,
  caption = "Overall convergence summary for glasso-fixed."
)
```
# Summary and Discussion {#sec-summary}


We have introduced **BVS.DAdj**, an R package for Bayesian variable
selection with dual-adjacency network-informed Ising priors. The package
provides a comprehensive MCMC framework covering four outcome types,
six adjacency specifications, and three sampling algorithms (MH, HMC, NUTS)
for non-conjugate parameter blocks, alongside Pólya-Gamma augmented Gibbs
for binary outcomes. A key methodological contribution is the simultaneous
incorporation of two independent structural information sources---one from
domain knowledge (e.g., a biological pathway database) and one estimated
from data (e.g., via Bayesian GGM)---with coupling strengths inferred from
data via the Möller auxiliary variable approach.

**Simulation results** in this vignette focus on the first moderate-dimensional
example ($n=500$, $p=100$) with the `glasso_fixed` method only. We provide
code to summarize PIPs, selected predictors at PIP $> 0.5$, and
variable-selection metrics (TPR, FPR, F1, FDR, precision), together with
trace/ACF diagnostics and two-chain convergence summaries via $\\hat{R}$ and ESS.

**Limitations and future directions.** The current implementation focuses
on binary, Gaussian, Cox-TTE, and count outcomes for cross-sectional
designs. Extensions to longitudinal or multi-omics designs, multi-response
variable selection, or hierarchical network priors are natural directions.
The Möller auxiliary variable update scales with the Propp-Wilson mixing
time, which can be slow for very dense or near-critical Ising graphs;
alternative normalizing constant estimators [@MB2006] or variational
approximations could improve scalability. Future versions will explore
GPU-based backends for ultra-high-dimensional settings ($p > 10\,000$).

# Software Availability {-}

**BVS.DAdj** version 0.1.0 is available at
[https://github.com/yuy113/BVS.DAdj](https://github.com/yuy113/BVS.DAdj).
It requires R $\geq$ 3.5.0 with **Rcpp**, **RcppArmadillo**, and
**Matrix**. The **huge** package is recommended for graphical lasso
adjacency estimation. All analyses in this article used R version 4.3.

# References {-}
