---
title: "BVS.DAdj: Bayesian Variable Selection with Dual-Adjacency Network-Informed Ising Priors in R"
author:
  - name: Yubing Yao
    affiliation: Department of Biostatistics and Epidemiology, University of Massachusetts Amherst
    email: yyao@umass.edu
  - name: Mahlet Tadesse
    affiliation: Department of Mathematics & Statistics, Georgetown University
    email: mgt26@georgetown.edu
  - name: Raji Balasubramanian
    affiliation: Department of Biostatistics and Epidemiology, University of Massachusetts Amherst School of Public Health
    email: rbalasub@schoolph.umass.edu
abstract: >
  This article introduces the **BVS.DAdj** R package, which implements Bayesian
  variable selection for logistic regression models using Markov random field
  (MRF) priors that incorporate dual adjacency structures among predictors.
  The package provides two MCMC sampling strategies---Metropolis-Hastings (MH)
  and Pólya-Gamma (PG) data augmentation---each supporting five adjacency types:
  fixed, dual-fixed, graphical lasso, Bayesian GGM, and combined GGM-plus-fixed.
  Both dense ($p < n$) and sparse high-dimensional ($p \gg n$) backends are
  implemented in C++ via **Rcpp** and **RcppArmadillo** for computational
  efficiency.  We describe the underlying statistical model, the package API,
  and demonstrate its usage through simulation studies under moderate
  ($n = 500, p = 100$) and high-dimensional ($n = 200, p = 1000$) settings,
  including MCMC convergence diagnostics, variable selection performance metrics,
  and running time comparisons across sampling backends.
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: kable
vignette: >
  %\VignetteIndexEntry{BVS.DAdj: R Journal Paper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
is_truthy <- function(x) {
  tolower(trimws(as.character(x))) %in% c("1", "true", "t", "yes", "y", "on")
}

quick_env <- Sys.getenv("BVS_DADJ_QUICK_CHECK", unset = "")
quick_opt <- getOption("bvs.dadj.quick_check", FALSE)
quick_check <- is_truthy(quick_env) || isTRUE(quick_opt)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  cache = TRUE,
  cache.extra = if (quick_check) "quick-check-real-eval-v2" else "full-run-real-eval-v2"
)

# Helper for quick smoke-test runs:
# set Sys.setenv(BVS_DADJ_QUICK_CHECK = "true") before render.
mcmc_scale <- function(full_niter, full_burnin, full_thin = 1L,
                       quick_niter = 5L, quick_burnin = 2L, quick_thin = 1L) {
  if (isTRUE(quick_check)) {
    list(
      niter = as.integer(quick_niter),
      burnin = as.integer(quick_burnin),
      thin = as.integer(quick_thin)
    )
  } else {
    list(
      niter = as.integer(full_niter),
      burnin = as.integer(full_burnin),
      thin = as.integer(full_thin)
    )
  }
}

cat(sprintf("BVS_DAdj_paper quick-check mode: %s\n", if (quick_check) "ON" else "OFF"))

# Robust ACF plotting helper:
# Handles constant or non-finite chains to avoid plot failures during vignette render.
safe_acf_plot <- function(x, lag.max = 50L, main = "", col = "steelblue") {
  x <- as.numeric(x)
  x <- x[is.finite(x)]

  if (length(x) < 2L) {
    plot.new()
    title(main = main)
    mtext("Insufficient finite samples for ACF", side = 3, line = -1.5, cex = 0.85)
    return(invisible(NULL))
  }

  v <- stats::var(x)
  if (!is.finite(v) || v <= .Machine$double.eps) {
    lags <- 0:as.integer(lag.max)
    acf_vals <- c(1, rep(0, length(lags) - 1L))
    plot(lags, acf_vals,
      type = "h", lwd = 2, col = col,
      ylim = c(-0.05, 1.05), xlab = "Lag", ylab = "ACF", main = main
    )
    points(lags, acf_vals, pch = 16, cex = 0.5, col = col)
    mtext("Constant chain: ACF undefined", side = 3, line = 0.1, cex = 0.75)
    return(invisible(NULL))
  }

  stats::acf(x, lag.max = as.integer(lag.max), main = main, col = col)
}

safe_gelman_df <- function(fits, vars = NULL, autoburnin = FALSE) {
  gd <- tryCatch(
    bvs_gelman_diag(fits, vars = vars, autoburnin = autoburnin),
    error = function(e) e
  )

  if (inherits(gd, "error")) {
    return(data.frame(
      Parameter = "NA",
      Point_est = NA_real_,
      Upper_CI = NA_real_,
      Note = conditionMessage(gd),
      stringsAsFactors = FALSE
    ))
  }

  psrf <- as.data.frame(gd$psrf, stringsAsFactors = FALSE)
  if (ncol(psrf) < 2L) {
    return(data.frame(
      Parameter = "NA",
      Point_est = NA_real_,
      Upper_CI = NA_real_,
      Note = "Gelman output did not contain PSRF columns.",
      stringsAsFactors = FALSE
    ))
  }
  names(psrf)[1:2] <- c("Point_est", "Upper_CI")
  psrf$Point_est <- as.numeric(psrf$Point_est)
  psrf$Upper_CI <- as.numeric(psrf$Upper_CI)

  param <- rownames(psrf)
  if (is.null(param) || length(param) != nrow(psrf)) {
    param <- rep("", nrow(psrf))
  }
  param <- trimws(as.character(param))
  blank <- which(!nzchar(param))
  if (length(blank) > 0L) {
    if (is.null(vars)) {
      beta_labels <- paste0("Beta", seq_len(length(blank)))
    } else {
      vars <- as.integer(vars)
      beta_labels <- paste0("Beta", vars[seq_len(min(length(vars), length(blank)))])
      if (length(beta_labels) < length(blank)) {
        beta_labels <- c(
          beta_labels,
          paste0("Beta", seq_len(length(blank) - length(beta_labels)) + length(beta_labels))
        )
      }
    }
    param[blank] <- beta_labels
  }

  out <- data.frame(
    Parameter = param,
    Point_est = psrf$Point_est,
    Upper_CI = psrf$Upper_CI,
    Note = "",
    stringsAsFactors = FALSE
  )
  out
}

gelman_method_summary <- function(gelman_df, method_name, rhat_threshold = 1.1) {
  keep <- is.finite(gelman_df$Point_est)
  if (!any(keep)) {
    return(data.frame(
      Method = method_name,
      N_parameters = 0L,
      Max_Rhat = NA_real_,
      Median_Rhat = NA_real_,
      N_Rhat_gt_1_1 = NA_integer_,
      stringsAsFactors = FALSE
    ))
  }
  data.frame(
    Method = method_name,
    N_parameters = sum(keep),
    Max_Rhat = round(max(gelman_df$Point_est[keep], na.rm = TRUE), 3),
    Median_Rhat = round(stats::median(gelman_df$Point_est[keep], na.rm = TRUE), 3),
    N_Rhat_gt_1_1 = sum(gelman_df$Point_est[keep] > rhat_threshold, na.rm = TRUE),
    stringsAsFactors = FALSE
  )
}

extract_beta_chain <- function(fit, j) {
  b <- fit$beta
  if (is.matrix(b)) {
    j <- max(1L, min(as.integer(j), ncol(b)))
    return(as.numeric(b[, j]))
  }
  if (is.list(b)) {
    n_save <- length(b)
    out <- numeric(n_save)
    for (s in seq_len(n_save)) {
      bs <- b[[s]]
      if (is.null(bs) || length(bs) == 0) {
        out[s] <- 0
      } else if (is.matrix(bs) && ncol(bs) >= 2) {
        ids <- as.integer(bs[, 1])
        vals <- as.numeric(bs[, 2])
        hit <- which(ids == as.integer(j))
        out[s] <- if (length(hit) > 0) vals[hit[1]] else 0
      } else if (is.numeric(bs) && length(bs) >= as.integer(j)) {
        out[s] <- as.numeric(bs[as.integer(j)])
      } else {
        out[s] <- 0
      }
    }
    return(out)
  }
  return(numeric(0))
}

plot_two_chain_trace_panels <- function(fit1, fit2, beta_idx, method_label) {
  beta_idx <- as.integer(beta_idx)
  for (j in beta_idx) {
    ch1 <- extract_beta_chain(fit1, j)
    ch2 <- extract_beta_chain(fit2, j)
    y_lim <- range(c(ch1, ch2), finite = TRUE)
    if (length(y_lim) != 2 || !all(is.finite(y_lim))) y_lim <- c(-1, 1)
    plot(ch1,
      type = "l", col = "steelblue", lwd = 1,
      main = paste0(method_label, ": beta_", j),
      xlab = "Saved iteration", ylab = "beta",
      ylim = y_lim
    )
    lines(ch2, col = "firebrick", lwd = 1)
    legend("topright",
      legend = c("Chain 1", "Chain 2"),
      col = c("steelblue", "firebrick"),
      lty = 1, cex = 0.7, bty = "n"
    )
  }
}

# Explicit FDR-targeted hyperparameters (single and dual network Moller updates).
hpars_sim1 <- list(
  mu = qlogis(0.01),
  eta1_sd = 0.08,
  eta1_sd = 0.10, eta2_sd = 0.06,
  e_eta = 1.0, f_eta = 25.0,
  mu_tilde = -7.0,
  eta1_tilde = 0.005, eta2_tilde = 0.004
)
hpars_sim2 <- list(
  mu = qlogis(0.02),
  eta1_sd = 0.05,
  eta1_sd = 0.06, eta2_sd = 0.05,
  e_eta = 1.0, f_eta = 30.0,
  mu_tilde = qlogis(0.003),
  eta1_tilde = 0.01, eta2_tilde = 0.008
)
hpars_real <- hpars_sim1
hpars_realhd <- hpars_sim2
```

# Introduction

In metabolic case-control studies and high-throughput biological experiments, a
central goal is to identify which of many candidate variables (e.g., metabolites,
genes) are genuinely associated with a disease outcome.  These variables often
exhibit complex dependency structures arising from multiple sources: biological
pathway databases provide partial knowledge of inter-variable relationships,
while data-driven methods can reveal additional conditional dependencies.

Standard penalization methods such as LASSO [@Tib1996] and its extensions
[@ZH2005; @YL2006; @Zou2006] perform variable selection by shrinking regression
coefficients, but may miss weakly associated variables that are correlated with
stronger signals.  Bayesian variable selection with spike-and-slab priors
[@GM1993; @GM1997] provides a principled alternative that quantifies uncertainty
through posterior inclusion probabilities (PIPs).

Building on the Markov random field (MRF) framework of @LZ2010 and @STV2011,
the **BVS.DAdj** package extends the Ising prior on selection indicators
$\boldsymbol{\gamma}$ to incorporate *two independent* adjacency structures
simultaneously---one from prior knowledge and one estimated from data.  This
dual-adjacency formulation allows borrowing strength across both sources of
structural information.

The package implements 12 optimized C++ backends accessed through two
user-facing R functions, `bvs_mh()` and `bvs_pg()`, covering a complete matrix
of sampler type × adjacency source × dimensionality.

The remainder of this article is organized as follows. Section 2 describes the
statistical model and prior specifications. Section 3 outlines the MCMC sampling
strategies. Section 4 describes the package design and API. Sections 5 and 6
present simulation studies for the moderate-dimensional and high-dimensional
settings, respectively. Section 7 summarizes the results.

# Model formulation and priors {#sec-model}

## Logistic regression model

Consider a case-control study with binary outcome $\mathbf{y} = (y_1, \ldots, y_n)^\top$ where
$y_k \in \{0, 1\}$ and a design matrix $\mathbf{X} = (X_{ik})_{n \times p}$
of $p$ candidate predictors for $n$ observations.  The logistic regression model
is
$$
\text{logit}\,\Pr(y_k = 1) = \alpha + \mathbf{X}_k \boldsymbol{\beta}, \quad k = 1, \ldots, n
$$
where $\alpha$ is the intercept and $\boldsymbol{\beta} = (\beta_1, \ldots, \beta_p)^\top$ is the vector of regression coefficients.

## Spike-and-slab prior for variable selection

We introduce binary selection indicators $\boldsymbol{\gamma} = (\gamma_1, \ldots, \gamma_p)^\top \in \{0,1\}^p$ with a point-mass mixture prior on each coefficient:
$$
\beta_j \mid \gamma_j, \sigma^2 \sim \gamma_j \cdot \mathcal{N}(\beta_0, \sigma^2) + (1 - \gamma_j) \cdot \delta_0(\beta_j), \quad j = 1, \ldots, p
$$
The intercept receives a conjugate normal prior $\alpha \sim \mathcal{N}(\alpha_0, h_\alpha \sigma^2)$ with $h_\alpha$ large, and the variance component follows an inverse-gamma distribution:
$$
\sigma^2 \sim \text{Inverse-Gamma}\!\left(\frac{\nu_0}{2},\; \frac{\nu_0 \sigma_0^2}{2}\right)
$$

## Variant MRF prior with dual adjacency

Let $\mathbf{R}$ and $\mathbf{R}'$ be two $p \times p$ symmetric binary adjacency matrices encoding two independent sources of structural information among the predictors.  The prior on $\boldsymbol{\gamma}$ takes the Ising/MRF form:
$$
p(\boldsymbol{\gamma} \mid \mu, \eta_1, \eta_2) = \frac{1}{Z(\mu,\eta_1,\eta_2)}\exp\!\Big(\mu\,\mathbf{1}_p^\top \boldsymbol{\gamma} + \eta_1\, \boldsymbol{\gamma}^\top \mathbf{R}\, \boldsymbol{\gamma} + \eta_2\, \boldsymbol{\gamma}^\top \mathbf{R}'\, \boldsymbol{\gamma}\Big)
$$
where:

- $\mu$ controls the baseline sparsity level.  For a predictor with no neighbors in either graph, the marginal inclusion probability is $\text{expit}(\mu)$.
- $\eta_1, \eta_2 \geq 0$ are coupling parameters controlling how strongly a predictor's inclusion is influenced by its neighbors in $\mathbf{R}$ and $\mathbf{R}'$, respectively.
- $Z(\mu, \eta_1, \eta_2)$ is the intractable normalizing constant.

**Eta prior used in the package.**
In the implementation, each coupling parameter is bounded by a user-specified
upper limit (`eta1_sd`, `eta1_sd`, `eta2_sd`) and sampled on a scaled Beta domain:
$$
\eta_i = \eta_{i,\max} u_i,\quad u_i \sim \text{Beta}(e, f), \quad i=1,2.
$$
The helper `phase_transition()` can be used to choose practically stable values
for $\eta_{i,\max}$ before fitting the full model.

## Estimating unknown adjacency: Bayesian GGM

When the second adjacency $\mathbf{R}'$ is unknown, we model the predictors as
$\mathbf{X}_k \sim \mathcal{N}_p(\mathbf{0}, \boldsymbol{\Omega}^{-1})$ and place a continuous spike-and-slab SSVS prior on $\boldsymbol{\Omega}$ following @WH2015:
$$
p(\boldsymbol{\Omega} \mid \mathbf{R}', \theta) \propto \prod_{i < j} \mathcal{N}\!\left(\omega_{ij} \mid 0,\, v^2_{r'_{ij}}\right) \prod_i \text{Exp}\!\left(\omega_{ii} \mid \tfrac{\lambda}{2}\right) \cdot \mathbf{1}_{\boldsymbol{\Omega} \in M^+}
$$
where $v_{r'_{ij}} = v_0$ (spike) if $r'_{ij} = 0$ and $v_1$ (slab) if $r'_{ij} = 1$.  Edge indicators are updated via:
$$
\Pr(r'_{ij} = 1 \mid \boldsymbol{\Omega}) = \frac{\pi\, \mathcal{N}(\omega_{ij} \mid 0, v_1^2)}{\pi\, \mathcal{N}(\omega_{ij} \mid 0, v_1^2) + (1-\pi)\, \mathcal{N}(\omega_{ij} \mid 0, v_0^2)}
$$

# MCMC sampling strategies {#sec-mcmc}

## Metropolis-Hastings sampler

The MH sampler (`bvs_mh()`) uses random-walk proposals for $\boldsymbol{\beta}$, $\alpha$, and $\sigma^2$, and a component-wise add/delete move for $\gamma_j$.  For each $j$, a flip $\gamma_j^* = 1 - \gamma_j$ is proposed with the corresponding $\beta_j^*$ drawn from the spike-and-slab prior; the move is accepted with probability $\min(1, \exp(\Delta))$ where
$$
\Delta = \ell(\boldsymbol{\beta}^*, \boldsymbol{\gamma}^*) - \ell(\boldsymbol{\beta}, \boldsymbol{\gamma}) - \frac{(\beta_j - \beta_0)^2 - (\beta_j^* - \beta_0)^2}{2\sigma^2}
$$

## Pólya-Gamma augmented Gibbs sampler

The PG sampler (`bvs_pg()`) introduces auxiliary variables $\omega_k \sim \text{PG}(1, \psi_k)$
where $\psi_k = \alpha + \mathbf{X}_k \boldsymbol{\beta}$, rendering the logistic likelihood conditionally Gaussian [@PG2013].  In the C++ backend, PG draws are obtained through a truncated-series approximation, after which block updates for $(\boldsymbol{\beta}, \alpha)$ use Gaussian conditional forms:
$$
\boldsymbol{\beta}_\gamma \mid \cdot \sim \mathcal{N}\!\left(\mathbf{V}_\gamma \mathbf{m}_\gamma,\; \mathbf{V}_\gamma\right), \quad \mathbf{V}_\gamma = \left(\mathbf{X}_\gamma^\top \boldsymbol{\Omega}_\omega \mathbf{X}_\gamma + \sigma^{-2} \mathbf{I}\right)^{-1}
$$
where $\boldsymbol{\Omega}_\omega = \text{diag}(\omega_1, \ldots, \omega_n)$ and $\kappa_k = y_k - 1/2$.

## Möller auxiliary variable update for $\eta_1, \eta_2$

The normalizing constant $Z(\mu, \eta_1, \eta_2)$ is intractable.  Following @MB2006, we introduce auxiliary variables $\boldsymbol{\omega}$ drawn via Propp-Wilson perfect simulation from the Ising model at reference parameters $(\tilde{\mu}, \tilde{\eta}_1, \tilde{\eta}_2)$, yielding an MH ratio that cancels the intractable ratio $Z'/Z$.

## Sparse high-dimensional variant

For $p \gg n$, the package stores the empirical covariance $\mathbf{S} = \mathbf{X}^\top\mathbf{X}$ in compressed sparse column (CSC) format and uses column-wise block Gibbs steps for the precision matrix.  The selection indicator updates use a targeted MH scheme cycling through `n_mh_gamma` randomly chosen variables per iteration.

# Package design and API {#sec-api}

## Installation

```{r install, eval=FALSE}
# Install from GitHub
devtools::install_github("yuy113/BVS.DAdj")
```

```{r load-pkg}
library(BVS.DAdj)
```

## User-facing functions

The package exports two primary functions that dispatch to 12 C++ backends:

| Function   | Sampler | Primary usage |
|:-----------|:--------|:--------------|
| `bvs_mh()` | Metropolis-Hastings | All adjacency types; general purpose |
| `bvs_pg()` | Pólya-Gamma Gibbs   | All adjacency types; faster mixing for moderate $p$ |

Both accept a common interface:

```{r api-example, eval=FALSE}
fit <- bvs_mh(X, y,
  adj_type = c("fixed", "dual_fixed", "glasso", "glasso_fixed", "ggm", "ggm_fixed"),
  adj_fixed = NULL, # Required for "fixed", "dual_fixed", "glasso_fixed", "ggm_fixed"
  adj_fixed2 = NULL, # Required for "dual_fixed"
  sparse = FALSE, # TRUE for sparse-GGM backends
  niter = 60000, burnin = 10000, thin = 1,
  # ... additional hyperparameters
)
```

## Adjacency types

| `adj_type` | Adjacency source(s) | Eta | Backend count |
|:-----------|:---------------------|:----|:--------------|
| `"fixed"` | One user-supplied binary matrix | 1 | MH + PG |
| `"dual_fixed"` | Two user-supplied binary matrices | 2 | MH + PG |
| `"glasso"` | Graphical lasso estimated from $\mathbf{X}$ | 1 | MH + PG |
| `"glasso_fixed"` | Graphical lasso from $\mathbf{X}$ + one fixed matrix | 2 | MH + PG |
| `"ggm"` | Bayesian GGM SSVS (dense or sparse) | 1 | MH + PG × dense/sparse |
| `"ggm_fixed"` | GGM + one user-supplied fixed matrix | 2 | MH + PG × dense/sparse |

## Post-processing

```{r post-process, eval=FALSE}
summary(fit) # Selected variables, PIPs, posterior means/intervals
plot(fit) # PIP bar plot (optionally top_n)
```

# Simulation study 1: Moderate dimensions ($n = 500, p = 100$) {#sec-sim1}

## Data generation

We generate a logistic regression dataset with $n = 500$ observations, $p = 100$ covariates, $p^* = 10$ truly active variables, and a block-correlation structure.  Covariates within the active set have pairwise correlation $\rho = 0.5$ to emulate metabolite clustering.

```{r sim1-data}
set.seed(2024)
n <- 500
p <- 100
p_star <- 10

# Construct correlation matrix: block structure for first p_star variables
Sigma <- diag(p)
for (i in 1:p_star) {
  for (j in 1:p_star) {
    if (i != j) Sigma[i, j] <- 0.5
  }
}

# Generate covariates from multivariate normal
X <- MASS::mvrnorm(n, mu = rep(0, p), Sigma = Sigma)

# True coefficients: first p_star variables have nonzero effects
beta_true <- rep(0, p)
beta_true[1:p_star] <- c(1.5, -1.2, 1.0, -0.8, 0.6, -0.5, 0.4, -0.3, 0.25, -0.2)

# Generate binary response
eta_true <- X %*% beta_true
prob_true <- plogis(eta_true)
y <- rbinom(n, 1, prob_true)

cat(sprintf(
  "Response balance: %d cases / %d controls (%.1f%%)\n",
  sum(y), sum(1 - y), 100 * mean(y)
))
```

## Adjacency matrix construction

For the additional adjacency-type comparison, we construct:

1. A **true** adjacency matrix `R_true1` based on the simulation graph used to
   generate correlated predictors (first `p_star` variables fully connected).
2. An **external fixed** adjacency matrix `R_ext50_1` by randomly retaining
   50% of the `1` entries in `R_true1` (symmetric, zero diagonal).

For backward compatibility with the original baseline comparison in this paper,
we also keep `R_known` (partial/noisy fixed adjacency) used by the original
three-method benchmark.

```{r sim1-adj}
# True adjacency used in simulation design (first p_star block)
R_true1 <- matrix(0L, p, p)
for (i in 1:p_star) {
  for (j in (i + 1):p_star) {
    R_true1[i, j] <- R_true1[j, i] <- 1L
  }
}
diag(R_true1) <- 0L

# External fixed adjacency: random 50% of true edges
set.seed(202401)
true_pairs1 <- which(upper.tri(R_true1) & R_true1 == 1L, arr.ind = TRUE)
n_keep1 <- ceiling(0.5 * nrow(true_pairs1))
keep_idx1 <- sample(seq_len(nrow(true_pairs1)), n_keep1)
keep_pairs1 <- true_pairs1[keep_idx1, , drop = FALSE]
R_ext50_1 <- matrix(0L, p, p)
R_ext50_1[cbind(keep_pairs1[, 1], keep_pairs1[, 2])] <- 1L
R_ext50_1[cbind(keep_pairs1[, 2], keep_pairs1[, 1])] <- 1L
diag(R_ext50_1) <- 0L

# Legacy partial/noisy adjacency from earlier version of the vignette
R_known <- matrix(0L, p, p)
# Block edges among first 8 active variables
for (i in 1:8) {
  for (j in (i + 1):8) {
    R_known[i, j] <- R_known[j, i] <- 1L
  }
}
# Add a few noise edges
set.seed(123)
noise_pairs <- matrix(sample((p_star + 1):p, 10), ncol = 2)
for (k in 1:nrow(noise_pairs)) {
  R_known[noise_pairs[k, 1], noise_pairs[k, 2]] <- 1L
  R_known[noise_pairs[k, 2], noise_pairs[k, 1]] <- 1L
}
diag(R_known) <- 0L
cat(sprintf("True adjacency edges (R_true1): %d\n", sum(R_true1) / 2))
cat(sprintf("External 50%% adjacency edges (R_ext50_1): %d\n", sum(R_ext50_1) / 2))
cat(sprintf("Legacy fixed adjacency edges (R_known): %d\n", sum(R_known) / 2))
```

## Running the samplers

We run three backends on this dataset:

1. **MH with fixed adjacency** (`bvs_mh`, `adj_type = "fixed"`)
2. **PG with fixed adjacency** (`bvs_pg`, `adj_type = "fixed"`)
3. **PG with Bayesian GGM** (`bvs_pg`, `adj_type = "ggm"`, dense)

```{r sim1-run, cache=TRUE}
cfg_sim1 <- mcmc_scale(
  full_niter = 500L, full_burnin = 200L, full_thin = 1L,
  quick_niter = 5L, quick_burnin = 2L, quick_thin = 1L
)
niter_sim <- cfg_sim1$niter
burnin_sim <- cfg_sim1$burnin
thin_sim <- cfg_sim1$thin

# --- MH with fixed adjacency ---
t_mh <- system.time({
  fit_mh <- bvs_mh(X, y,
    adj_type = "fixed", adj_fixed = R_known,
    niter = niter_sim, burnin = burnin_sim,
    mu = -log(1 / 0.1 - 1), thin = thin_sim
  )
})

# --- PG with fixed adjacency ---
t_pg <- system.time({
  fit_pg <- bvs_pg(X, y,
    adj_type = "fixed", adj_fixed = R_known,
    niter = niter_sim, burnin = burnin_sim,
    mu = -log(1 / 0.1 - 1), thin = thin_sim
  )
})

# --- PG with dense Bayesian GGM ---
t_pg_ggm <- system.time({
  fit_pg_ggm <- bvs_pg(X, y,
    adj_type = "ggm", sparse = FALSE,
    niter = niter_sim, burnin = burnin_sim,
    mu = -log(1 / 0.1 - 1), thin = thin_sim
  )
})
```

## Running time comparison

```{r sim1-timing}
timing_df <- data.frame(
  Method = c("MH (fixed adj)", "PG (fixed adj)", "PG (Bayesian GGM)"),
  Elapsed_sec = round(c(t_mh["elapsed"], t_pg["elapsed"], t_pg_ggm["elapsed"]), 1),
  Iter_per_sec = round(niter_sim / c(t_mh["elapsed"], t_pg["elapsed"], t_pg_ggm["elapsed"]), 0)
)
knitr::kable(
  timing_df,
  caption = sprintf(
    "Running time comparison ($n = 500, p = 100$, %s post-burn-in iterations, burn-in %s).",
    format(niter_sim, big.mark = ","),
    format(burnin_sim, big.mark = ",")
  ),
  col.names = c("Method", "Elapsed (sec)", "Iter/sec")
)
```

## MCMC convergence diagnostics

### Trace plots

```{r sim1-trace, fig.height=8, fig.width=8, fig.cap="Trace plots of selected regression coefficients (beta_1 through beta_4) from the three samplers. The dashed red lines indicate true coefficient values."}
par(mfrow = c(3, 4), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
vars_to_plot <- 1:4

for (method_name in c("MH (fixed)", "PG (fixed)", "PG (GGM)")) {
  fit_cur <- switch(method_name,
    "MH (fixed)" = fit_mh,
    "PG (fixed)" = fit_pg,
    "PG (GGM)"   = fit_pg_ggm
  )
  for (j in vars_to_plot) {
    plot(fit_cur$beta[, j],
      type = "l", col = "steelblue",
      main = paste0(method_name, ": beta_", j),
      xlab = "Iteration", ylab = expression(beta[j]),
      cex.main = 0.85
    )
    abline(h = beta_true[j], col = "red", lty = 2, lwd = 1.5)
  }
}
```

### Running mean plots

```{r sim1-runmean, fig.height=6, fig.width=8, fig.cap="Running mean of beta_1 and beta_2 across the three sampling methods, demonstrating convergence to the true value (dashed red line)."}
running_mean <- function(x) cumsum(x) / seq_along(x)

par(mfrow = c(2, 3), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (j in 1:2) {
  for (method_name in c("MH (fixed)", "PG (fixed)", "PG (GGM)")) {
    fit_cur <- switch(method_name,
      "MH (fixed)" = fit_mh,
      "PG (fixed)" = fit_pg,
      "PG (GGM)"   = fit_pg_ggm
    )
    rm_vals <- running_mean(fit_cur$beta[, j])
    plot(rm_vals,
      type = "l", col = "darkgreen", lwd = 1.2,
      main = paste0(method_name, ": beta_", j),
      xlab = "Iteration", ylab = "Running mean",
      cex.main = 0.85
    )
    abline(h = beta_true[j], col = "red", lty = 2, lwd = 1.5)
  }
}
```

### Autocorrelation

```{r sim1-acf, fig.height=4, fig.width=8, fig.cap="Autocorrelation function of beta_1 for the three sampling methods."}
par(mfrow = c(1, 3), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
safe_acf_plot(fit_mh$beta[, 1], lag.max = 50, main = "MH (fixed): beta_1", col = "steelblue")
safe_acf_plot(fit_pg$beta[, 1], lag.max = 50, main = "PG (fixed): beta_1", col = "steelblue")
safe_acf_plot(fit_pg_ggm$beta[, 1], lag.max = 50, main = "PG (GGM): beta_1", col = "steelblue")
```

### Gelman-Rubin diagnostics (2 chains, 2 seeds)

```{r sim1-gelman, cache=TRUE}
vars_gr1 <- 1:4
seed_sim1 <- c(20261101L, 20261102L)

set.seed(seed_sim1[1])
fit_mh_gr1 <- bvs_mh(
  X, y,
  adj_type = "fixed", adj_fixed = R_known,
  niter = niter_sim, burnin = burnin_sim,
  mu = -log(1 / 0.1 - 1), thin = thin_sim
)
set.seed(seed_sim1[2])
fit_mh_gr2 <- bvs_mh(
  X, y,
  adj_type = "fixed", adj_fixed = R_known,
  niter = niter_sim, burnin = burnin_sim,
  mu = -log(1 / 0.1 - 1), thin = thin_sim
)

set.seed(seed_sim1[1])
fit_pg_gr1 <- bvs_pg(
  X, y,
  adj_type = "fixed", adj_fixed = R_known,
  niter = niter_sim, burnin = burnin_sim,
  mu = -log(1 / 0.1 - 1), thin = thin_sim
)
set.seed(seed_sim1[2])
fit_pg_gr2 <- bvs_pg(
  X, y,
  adj_type = "fixed", adj_fixed = R_known,
  niter = niter_sim, burnin = burnin_sim,
  mu = -log(1 / 0.1 - 1), thin = thin_sim
)

set.seed(seed_sim1[1])
fit_pg_ggm_gr1 <- bvs_pg(
  X, y,
  adj_type = "ggm", sparse = FALSE,
  niter = niter_sim, burnin = burnin_sim,
  mu = -log(1 / 0.1 - 1), thin = thin_sim
)
set.seed(seed_sim1[2])
fit_pg_ggm_gr2 <- bvs_pg(
  X, y,
  adj_type = "ggm", sparse = FALSE,
  niter = niter_sim, burnin = burnin_sim,
  mu = -log(1 / 0.1 - 1), thin = thin_sim
)

gelman_sim1 <- list(
  "MH (fixed adj)" = safe_gelman_df(list(fit_mh_gr1, fit_mh_gr2), vars = vars_gr1, autoburnin = FALSE),
  "PG (fixed adj)" = safe_gelman_df(list(fit_pg_gr1, fit_pg_gr2), vars = vars_gr1, autoburnin = FALSE),
  "PG (Bayesian GGM)" = safe_gelman_df(list(fit_pg_ggm_gr1, fit_pg_ggm_gr2), vars = vars_gr1, autoburnin = FALSE)
)

gelman_sim1_summary <- do.call(rbind, lapply(names(gelman_sim1), function(mn) {
  gelman_method_summary(gelman_sim1[[mn]], mn)
}))

gelman_sim1_detail <- do.call(rbind, lapply(names(gelman_sim1), function(mn) {
  df <- gelman_sim1[[mn]]
  data.frame(
    Method = mn,
    Parameter = df$Parameter,
    Point_Rhat = round(df$Point_est, 3),
    Upper_CI_Rhat = round(df$Upper_CI, 3),
    stringsAsFactors = FALSE
  )
}))

knitr::kable(
  gelman_sim1_summary,
  caption = "Simulation 1 Gelman-Rubin diagnostics using two chains (two seeds) per method."
)
knitr::kable(
  gelman_sim1_detail,
  caption = "Simulation 1 PSRF details (Alpha and beta_1 to beta_4)."
)
```

```{r sim1-gelman-trace-2chain, fig.height=10, fig.width=12, fig.cap="Simulation 1: two-chain trace plots for representative coefficients across Gelman-Rubin diagnostic runs."}
beta_idx_gr1_trace <- unique(c(1L, 2L, 5L, min(p, p_star + 5L)))
par(
  mfrow = c(3, length(beta_idx_gr1_trace)),
  mar = c(3, 3, 2, 1),
  mgp = c(1.8, 0.5, 0)
)
plot_two_chain_trace_panels(fit_mh_gr1, fit_mh_gr2, beta_idx_gr1_trace, "MH (fixed)")
plot_two_chain_trace_panels(fit_pg_gr1, fit_pg_gr2, beta_idx_gr1_trace, "PG (fixed)")
plot_two_chain_trace_panels(fit_pg_ggm_gr1, fit_pg_ggm_gr2, beta_idx_gr1_trace, "PG (Bayesian GGM)")
```

## Variable selection performance

### Posterior inclusion probabilities

```{r sim1-pip, fig.height=5, fig.width=9, fig.cap="Posterior inclusion probabilities (PIPs) across all 100 variables. Red bars indicate truly active variables (1--10). The dashed line marks the 0.5 median probability threshold."}
compute_pip <- function(fit) {
  if (!is.null(fit$gamma_pip)) {
    return(as.numeric(fit$gamma_pip))
  }
  if (is.matrix(fit$gamma)) {
    return(colMeans(fit$gamma))
  }
  if (is.list(fit$gamma)) {
    p <- fit$p
    n_save <- length(fit$gamma)
    pip <- numeric(p)
    if (n_save > 0) {
      for (s in seq_len(n_save)) {
        idx <- as.integer(fit$gamma[[s]])
        idx <- idx[idx >= 1L & idx <= p]
        if (length(idx) > 0) pip[idx] <- pip[idx] + 1
      }
      pip <- pip / n_save
    }
    return(pip)
  }
  stop("No gamma samples available for PIP computation.")
}

get_gamma_matrix <- function(fit) {
  if (is.matrix(fit$gamma)) {
    return(fit$gamma)
  }
  if (is.list(fit$gamma)) {
    p <- fit$p
    n_save <- length(fit$gamma)
    G <- matrix(0, n_save, p)
    if (n_save > 0) {
      for (s in seq_len(n_save)) {
        idx <- as.integer(fit$gamma[[s]])
        idx <- idx[idx >= 1L & idx <= p]
        if (length(idx) > 0) G[s, idx] <- 1
      }
    }
    return(G)
  }
  stop("No gamma samples available.")
}

get_beta_matrix <- function(fit) {
  if (is.matrix(fit$beta)) {
    return(fit$beta)
  }
  if (is.list(fit$beta)) {
    p <- fit$p
    n_save <- length(fit$beta)
    B <- matrix(0, n_save, p)
    if (n_save > 0) {
      for (s in seq_len(n_save)) {
        bm <- fit$beta[[s]]
        if (is.null(bm) || length(bm) == 0) next
        bm <- as.matrix(bm)
        if (nrow(bm) > 0 && ncol(bm) >= 2) {
          idx <- as.integer(bm[, 1]) # C++ sparse backends emit 1-based indices
          valid <- idx >= 1L & idx <= p
          B[s, idx[valid]] <- as.numeric(bm[valid, 2])
        }
      }
    }
    return(B)
  }
  stop("No beta samples available.")
}

model_size_chain <- function(fit) {
  if (is.matrix(fit$gamma)) {
    return(rowSums(fit$gamma))
  }
  if (is.list(fit$gamma)) {
    return(vapply(fit$gamma, length, integer(1)))
  }
  stop("No gamma samples available.")
}

pip_mh <- compute_pip(fit_mh)
pip_pg <- compute_pip(fit_pg)
pip_ggm <- compute_pip(fit_pg_ggm)

par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
cols <- ifelse(1:p <= p_star, "firebrick", "grey60")

barplot(pip_mh,
  col = cols, border = NA, main = "MH (fixed adj)",
  xlab = "Variable index", ylab = "PIP", ylim = c(0, 1)
)
abline(h = 0.5, lty = 2, col = "black")

barplot(pip_pg,
  col = cols, border = NA, main = "PG (fixed adj)",
  xlab = "Variable index", ylab = "PIP", ylim = c(0, 1)
)
abline(h = 0.5, lty = 2, col = "black")

barplot(pip_ggm,
  col = cols, border = NA, main = "PG (Bayesian GGM)",
  xlab = "Variable index", ylab = "PIP", ylim = c(0, 1)
)
abline(h = 0.5, lty = 2, col = "black")
```

### Selection metrics

```{r sim1-metrics}
compute_metrics <- function(pip, threshold = 0.5, p_star = 10) {
  selected <- which(pip >= threshold)
  tp <- sum(selected <= p_star)
  fp <- sum(selected > p_star)
  fn <- p_star - tp
  sensitivity <- tp / p_star
  fdr <- if (length(selected) > 0) fp / length(selected) else 0
  f1 <- if (tp + fp + fn > 0) 2 * tp / (2 * tp + fp + fn) else 0
  data.frame(
    TP = tp, FP = fp, FN = fn,
    Sensitivity = round(sensitivity, 3),
    FDR = round(fdr, 3),
    F1 = round(f1, 3),
    N_selected = length(selected)
  )
}

metrics <- rbind(
  cbind(Method = "MH (fixed adj)", compute_metrics(pip_mh)),
  cbind(Method = "PG (fixed adj)", compute_metrics(pip_pg)),
  cbind(Method = "PG (Bayesian GGM)", compute_metrics(pip_ggm))
)
knitr::kable(metrics,
  caption = "Variable selection performance at PIP threshold 0.5 ($n = 500, p = 100$).",
  row.names = FALSE
)
```

### Eta coupling parameter traces

```{r sim1-eta, fig.height=4, fig.width=8, fig.cap="Trace plots of the Ising coupling parameter eta for the three sampling methods."}
par(mfrow = c(1, 3), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
if (!is.null(fit_mh$eta1)) {
  plot(fit_mh$eta1,
    type = "l", col = "purple",
    main = "MH: eta1", xlab = "Iteration", ylab = expression(eta)
  )
}
if (!is.null(fit_pg$eta1)) {
  plot(fit_pg$eta1,
    type = "l", col = "purple",
    main = "PG: eta1", xlab = "Iteration", ylab = expression(eta)
  )
}
if (!is.null(fit_pg_ggm$eta1)) {
  plot(fit_pg_ggm$eta1,
    type = "l", col = "purple",
    main = "PG GGM: eta1", xlab = "Iteration", ylab = expression(eta)
  )
}
```

## Additional adjacency-type comparison (5 methods) using 50% true external edges

To compare adjacency **types** directly, we run five additional methods using a
common MH sampler and the same simulated data. For fixed-network inputs, we use
`R_ext50_1` (random 50% of true adjacency edges):

1. `adj_type = "fixed"` (external `R_ext50_1` only),
2. `adj_type = "ggm"` (Bayesian GGM),
3. `adj_type = "glasso"` (data-driven glasso only),
4. `adj_type = "glasso_fixed"` (external `R_ext50_1` + glasso),
5. `adj_type = "ggm_fixed"` (external `R_ext50_1` + Bayesian GGM).

```{r sim1-adjtype-run, cache=TRUE}
sim1_adj_dir <- "sim1_adjtype_saved_runs"
if (!dir.exists(sim1_adj_dir)) dir.create(sim1_adj_dir, recursive = TRUE)

niter_adj1 <- niter_sim
burnin_adj1 <- burnin_sim
thin_adj1 <- thin_sim

sim1_f_fixed <- file.path(sim1_adj_dir, sprintf("sim1_fixed50_n%d_b%d_t%d.rds", niter_adj1, burnin_adj1, thin_adj1))
sim1_f_ggm <- file.path(sim1_adj_dir, sprintf("sim1_ggm_n%d_b%d_t%d.rds", niter_adj1, burnin_adj1, thin_adj1))
sim1_f_gls <- file.path(sim1_adj_dir, sprintf("sim1_glasso_n%d_b%d_t%d.rds", niter_adj1, burnin_adj1, thin_adj1))
sim1_f_gf <- file.path(sim1_adj_dir, sprintf("sim1_glasso_fixed_n%d_b%d_t%d.rds", niter_adj1, burnin_adj1, thin_adj1))
sim1_f_ggmf <- file.path(sim1_adj_dir, sprintf("sim1_ggm_fixed_n%d_b%d_t%d.rds", niter_adj1, burnin_adj1, thin_adj1))

if (file.exists(sim1_f_fixed)) {
  o <- readRDS(sim1_f_fixed)
  fit1_fixed50 <- o$fit
  t1_fixed50 <- o$time
} else {
  t1_fixed50 <- system.time({
    fit1_fixed50 <- bvs_mh(
      X, y,
      adj_type = "fixed", adj_fixed = R_ext50_1,
      niter = niter_adj1, burnin = burnin_adj1, thin = thin_adj1,
      mu = -log(1 / 0.1 - 1)
    )
  })
  saveRDS(list(fit = fit1_fixed50, time = t1_fixed50), sim1_f_fixed)
}

if (file.exists(sim1_f_ggm)) {
  o <- readRDS(sim1_f_ggm)
  fit1_ggm <- o$fit
  t1_ggm <- o$time
} else {
  t1_ggm <- system.time({
    fit1_ggm <- bvs_mh(
      X, y,
      adj_type = "ggm", sparse = FALSE,
      niter = niter_adj1, burnin = burnin_adj1, thin = thin_adj1,
      mu = -log(1 / 0.1 - 1)
    )
  })
  saveRDS(list(fit = fit1_ggm, time = t1_ggm), sim1_f_ggm)
}

if (file.exists(sim1_f_gls)) {
  o <- readRDS(sim1_f_gls)
  fit1_glasso <- o$fit
  t1_glasso <- o$time
} else {
  t1_glasso <- system.time({
    fit1_glasso <- bvs_mh(
      X, y,
      adj_type = "glasso", glasso_criterion = "ebic",
      niter = niter_adj1, burnin = burnin_adj1, thin = thin_adj1,
      mu = -log(1 / 0.1 - 1)
    )
  })
  saveRDS(list(fit = fit1_glasso, time = t1_glasso), sim1_f_gls)
}

if (file.exists(sim1_f_gf)) {
  o <- readRDS(sim1_f_gf)
  fit1_glasso_fixed <- o$fit
  t1_glasso_fixed <- o$time
} else {
  t1_glasso_fixed <- system.time({
    fit1_glasso_fixed <- bvs_mh(
      X, y,
      adj_type = "glasso_fixed", adj_fixed = R_ext50_1,
      glasso_criterion = "ebic",
      niter = niter_adj1, burnin = burnin_adj1, thin = thin_adj1,
      mu = hpars_sim1$mu,
      eta1_sd = hpars_sim1$eta1_sd, eta2_sd = hpars_sim1$eta2_sd,
      e_eta = hpars_sim1$e_eta, f_eta = hpars_sim1$f_eta,
      mu_tilde = hpars_sim1$mu_tilde,
      eta1_tilde = hpars_sim1$eta1_tilde, eta2_tilde = hpars_sim1$eta2_tilde
    )
  })
  saveRDS(list(fit = fit1_glasso_fixed, time = t1_glasso_fixed), sim1_f_gf)
}

if (file.exists(sim1_f_ggmf)) {
  o <- readRDS(sim1_f_ggmf)
  fit1_ggm_fixed <- o$fit
  t1_ggm_fixed <- o$time
} else {
  t1_ggm_fixed <- system.time({
    fit1_ggm_fixed <- bvs_mh(
      X, y,
      adj_type = "ggm_fixed", adj_fixed = R_ext50_1, sparse = FALSE,
      niter = niter_adj1, burnin = burnin_adj1, thin = thin_adj1,
      mu = hpars_sim1$mu,
      eta1_sd = hpars_sim1$eta1_sd, eta2_sd = hpars_sim1$eta2_sd,
      e_eta = hpars_sim1$e_eta, f_eta = hpars_sim1$f_eta,
      mu_tilde = hpars_sim1$mu_tilde,
      eta1_tilde = hpars_sim1$eta1_tilde, eta2_tilde = hpars_sim1$eta2_tilde
    )
  })
  saveRDS(list(fit = fit1_ggm_fixed, time = t1_ggm_fixed), sim1_f_ggmf)
}
```

```{r sim1-adjtype-timing}
timing_adj1 <- data.frame(
  Method = c("Fixed (50% true edges)", "GGM", "Glasso", "Glasso + Fixed", "GGM + Fixed"),
  Elapsed_sec = round(c(
    t1_fixed50["elapsed"], t1_ggm["elapsed"], t1_glasso["elapsed"],
    t1_glasso_fixed["elapsed"], t1_ggm_fixed["elapsed"]
  ), 2),
  Iter_per_sec = round(niter_adj1 / c(
    t1_fixed50["elapsed"], t1_ggm["elapsed"], t1_glasso["elapsed"],
    t1_glasso_fixed["elapsed"], t1_ggm_fixed["elapsed"]
  ), 1)
)
knitr::kable(
  timing_adj1,
  caption = "Simulation 1 adjacency-type comparison timing (MH sampler; fixed input uses random 50% of true adjacency edges).",
  row.names = FALSE
)
```

```{r sim1-adjtype-pip, fig.height=8, fig.width=10, fig.cap="Simulation 1: PIP comparison across adjacency types (fixed, ggm, glasso, glasso_fixed, ggm_fixed)."}
pip1_fixed50 <- compute_pip(fit1_fixed50)
pip1_ggm <- compute_pip(fit1_ggm)
pip1_glasso <- compute_pip(fit1_glasso)
pip1_gf <- compute_pip(fit1_glasso_fixed)
pip1_ggmf <- compute_pip(fit1_ggm_fixed)

show_vars1 <- unique(c(
  1:40,
  which(pip1_fixed50 > 0.1),
  which(pip1_ggm > 0.1),
  which(pip1_glasso > 0.1),
  which(pip1_gf > 0.1),
  which(pip1_ggmf > 0.1)
))
show_vars1 <- sort(show_vars1)
cols1 <- ifelse(show_vars1 <= p_star, "firebrick", "grey65")

pip_list1 <- list(
  "Fixed (50% true)" = pip1_fixed50,
  "GGM" = pip1_ggm,
  "Glasso" = pip1_glasso,
  "Glasso + Fixed" = pip1_gf,
  "GGM + Fixed" = pip1_ggmf
)

par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))
for (mn in names(pip_list1)) {
  barplot(pip_list1[[mn]][show_vars1],
    col = cols1, border = NA, ylim = c(0, 1),
    main = mn, xlab = "Variable", ylab = "PIP",
    names.arg = show_vars1, las = 2, cex.names = 0.5
  )
  abline(h = 0.5, lty = 2)
}
plot.new()
```

```{r sim1-adjtype-metrics}
metrics_adj1 <- rbind(
  cbind(Method = "Fixed (50% true)", compute_metrics(pip1_fixed50, p_star = p_star)),
  cbind(Method = "GGM", compute_metrics(pip1_ggm, p_star = p_star)),
  cbind(Method = "Glasso", compute_metrics(pip1_glasso, p_star = p_star)),
  cbind(Method = "Glasso + Fixed", compute_metrics(pip1_gf, p_star = p_star)),
  cbind(Method = "GGM + Fixed", compute_metrics(pip1_ggmf, p_star = p_star))
)
knitr::kable(
  metrics_adj1,
  caption = "Simulation 1 variable-selection performance for five adjacency-type methods.",
  row.names = FALSE
)
```

```{r sim1-adjtype-diagnostics}
safe_ess_sim <- function(x, lag.max = 200L) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  n <- length(x)
  if (n < 3L) {
    return(NA_real_)
  }
  v <- stats::var(x)
  if (!is.finite(v) || v <= .Machine$double.eps) {
    return(n)
  }
  ac <- as.numeric(stats::acf(x, lag.max = min(lag.max, n - 1L), plot = FALSE)$acf)[-1L]
  if (!length(ac)) {
    return(n)
  }
  first_nonpos <- which(ac <= 0)
  m <- if (length(first_nonpos)) first_nonpos[1L] - 1L else length(ac)
  if (m < 1L) {
    return(n)
  }
  tau <- 1 + 2 * sum(ac[seq_len(m)])
  if (!is.finite(tau) || tau <= 0) {
    return(n)
  }
  min(n, n / tau)
}

lag1_acf_sim <- function(x) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  if (length(x) < 3L) {
    return(NA_real_)
  }
  v <- stats::var(x)
  if (!is.finite(v) || v <= .Machine$double.eps) {
    return(NA_real_)
  }
  as.numeric(stats::acf(x, lag.max = 1L, plot = FALSE)$acf[2L])
}

fits_adj1 <- list(
  "Fixed (50% true)" = fit1_fixed50,
  "GGM" = fit1_ggm,
  "Glasso" = fit1_glasso,
  "Glasso + Fixed" = fit1_glasso_fixed,
  "GGM + Fixed" = fit1_ggm_fixed
)
times_adj1 <- c(
  "Fixed (50% true)" = as.numeric(t1_fixed50["elapsed"]),
  "GGM" = as.numeric(t1_ggm["elapsed"]),
  "Glasso" = as.numeric(t1_glasso["elapsed"]),
  "Glasso + Fixed" = as.numeric(t1_glasso_fixed["elapsed"]),
  "GGM + Fixed" = as.numeric(t1_ggm_fixed["elapsed"])
)

pip_mat_adj1 <- sapply(fits_adj1, compute_pip)
top_idx_adj1 <- order(rowMeans(pip_mat_adj1), decreasing = TRUE)[1L]
top_name_adj1 <- if (!is.null(colnames(X))) colnames(X)[top_idx_adj1] else paste0("V", top_idx_adj1)

diag_adj1 <- do.call(rbind, lapply(names(fits_adj1), function(mn) {
  fit <- fits_adj1[[mn]]
  pip <- compute_pip(fit)
  beta_chain <- get_beta_matrix(fit)[, top_idx_adj1]
  ms <- model_size_chain(fit)
  ess <- safe_ess_sim(beta_chain)
  n_saved <- length(ms)
  iter_sec <- n_saved / max(as.numeric(times_adj1[mn]), 1e-8)
  data.frame(
    Method = mn,
    Runtime_sec = round(times_adj1[mn], 2),
    N_saved = n_saved,
    Iter_per_sec = round(iter_sec, 2),
    Mean_model_size = round(mean(ms), 2),
    SD_model_size = round(stats::sd(ms), 2),
    Median_model_size = round(stats::median(ms), 2),
    Q05_model_size = round(as.numeric(stats::quantile(ms, 0.05)), 2),
    Q95_model_size = round(as.numeric(stats::quantile(ms, 0.95)), 2),
    Top_gene_for_diag = top_name_adj1,
    PIP_top_gene = round(pip[top_idx_adj1], 3),
    Lag1_ACF_top_beta = round(lag1_acf_sim(beta_chain), 3),
    ESS_top_beta = round(ess, 1),
    ESS_per_sec = round(ess / max(as.numeric(times_adj1[mn]), 1e-8), 3),
    MCSE_top_beta = round(stats::sd(beta_chain) / sqrt(max(1, ess)), 4),
    stringsAsFactors = FALSE
  )
}))

knitr::kable(
  diag_adj1,
  caption = "Simulation 1 MCMC convergence diagnostics and sampling performance for five adjacency-type methods.",
  row.names = FALSE
)
```

```{r sim1-adjtype-trace, fig.height=8, fig.width=10, fig.cap="Simulation 1 adjacency-type comparison: trace plots of the representative top-PIP coefficient."}
par(mfrow = c(2, 3), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (mn in names(fits_adj1)) {
  beta_chain <- get_beta_matrix(fits_adj1[[mn]])[, top_idx_adj1]
  plot(beta_chain,
    type = "l", col = "steelblue",
    main = paste0(mn, ": ", top_name_adj1),
    xlab = "Iteration", ylab = "beta", cex.main = 0.85
  )
}
plot.new()
```

```{r sim1-adjtype-acf, fig.height=8, fig.width=10, fig.cap="Simulation 1 adjacency-type comparison: ACF of the representative top-PIP coefficient."}
par(mfrow = c(2, 3), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (mn in names(fits_adj1)) {
  beta_chain <- get_beta_matrix(fits_adj1[[mn]])[, top_idx_adj1]
  safe_acf_plot(beta_chain,
    lag.max = 50,
    main = paste0(mn, ": ACF(", top_name_adj1, ")"),
    col = "steelblue"
  )
}
plot.new()
```

# Simulation study 2: High-dimensional sparse setting ($n = 200, p = 1000$) {#sec-sim2}

## Data generation

We now consider a challenging high-dimensional setting with $n = 200$ and $p = 1000$, where $p \gg n$.  Only $p^* = 15$ variables are truly active with a decay pattern in effect sizes.

```{r sim2-data}
set.seed(2025)
n2 <- 200
p2 <- 1000
p_star2 <- 15

# Independent covariates (sparse regime)
X2 <- matrix(rnorm(n2 * p2), n2, p2)

# Add mild correlation within the first 15 active variables
rho2 <- 0.3
for (i in 1:p_star2) {
  for (j in 1:p_star2) {
    if (i != j) {
      X2[, j] <- rho2 * X2[, i] + sqrt(1 - rho2^2) * X2[, j]
    }
  }
}

# True coefficients with decaying magnitudes
beta_true2 <- rep(0, p2)
signal_sizes <- seq(1.5, 0.3, length.out = p_star2)
signs <- rep(c(1, -1), length.out = p_star2)
beta_true2[1:p_star2] <- signs * signal_sizes

# Generate binary response
eta_true2 <- X2 %*% beta_true2
prob_true2 <- plogis(eta_true2)
y2 <- rbinom(n2, 1, prob_true2)
cat(sprintf("Response balance: %d / %d (%.1f%%)\n", sum(y2), n2 - sum(y2), 100 * mean(y2)))
```

## Adjacency matrix

```{r sim2-adj}
# True adjacency used by simulation design: fully connected active block
R_true2 <- matrix(0L, p2, p2)
for (i in 1:p_star2) {
  for (j in (i + 1):p_star2) {
    R_true2[i, j] <- R_true2[j, i] <- 1L
  }
}
diag(R_true2) <- 0L

# External fixed adjacency: random 50% of true edges
set.seed(202502)
true_pairs2 <- which(upper.tri(R_true2) & R_true2 == 1L, arr.ind = TRUE)
n_keep2 <- ceiling(0.5 * nrow(true_pairs2))
keep_idx2 <- sample(seq_len(nrow(true_pairs2)), n_keep2)
keep_pairs2 <- true_pairs2[keep_idx2, , drop = FALSE]
R_ext50_2 <- matrix(0L, p2, p2)
if (nrow(keep_pairs2) > 0L) {
  R_ext50_2[cbind(keep_pairs2[, 1], keep_pairs2[, 2])] <- 1L
  R_ext50_2[cbind(keep_pairs2[, 2], keep_pairs2[, 1])] <- 1L
}
diag(R_ext50_2) <- 0L

# Legacy fixed adjacency used by the original 3-method benchmark
R_known2 <- matrix(0L, p2, p2)
for (i in 1:10) {
  for (j in (i + 1):min(10, p2)) {
    R_known2[i, j] <- R_known2[j, i] <- 1L
  }
}
diag(R_known2) <- 0L

cat(sprintf("True adjacency edges (R_true2): %d\n", sum(R_true2) / 2))
cat(sprintf("External 50%% adjacency edges (R_ext50_2): %d\n", sum(R_ext50_2) / 2))
cat(sprintf("Legacy fixed adjacency edges (R_known2): %d\n", sum(R_known2) / 2))
```

## Running the sparse samplers

In the high-dimensional regime, we use the **sparse** backends designed for
$p \gg n$. To keep memory usage stable in vignette rendering, we use
memory-safe MCMC settings and run each sampler in a separate chunk with
intermediate results saved to disk.

```{r sim2-run-setup, cache=TRUE}
cfg_sim2 <- mcmc_scale(
  full_niter = 50000L, full_burnin = 20000L, full_thin = 2L,
  quick_niter = 5L, quick_burnin = 2L, quick_thin = 1L
)
niter_sim2 <- cfg_sim2$niter
burnin_sim2 <- cfg_sim2$burnin
thin_sim2 <- cfg_sim2$thin
sim2_n_mh_gamma <- if (isTRUE(quick_check)) 10L else 75L
pip_threshold_sim2 <- 0.10

# Force sparse GGM backends for high-dimensional examples
hd_sparse <- TRUE

sim2_dir <- "sim2_saved_runs"
if (!dir.exists(sim2_dir)) dir.create(sim2_dir, recursive = TRUE)

sim2_mh_file <- file.path(
  sim2_dir, sprintf("sim2_mh_sparse_n%d_b%d_t%d.rds", niter_sim2, burnin_sim2, thin_sim2)
)
sim2_pg_file <- file.path(
  sim2_dir, sprintf("sim2_pg_sparse_n%d_b%d_t%d.rds", niter_sim2, burnin_sim2, thin_sim2)
)
sim2_fx_file <- file.path(
  sim2_dir, sprintf("sim2_mh_fixed_n%d_b%d_t%d.rds", niter_sim2, burnin_sim2, thin_sim2)
)
```

```{r sim2-run-mh-sparse, cache=FALSE}
if (file.exists(sim2_mh_file)) {
  mh_obj <- readRDS(sim2_mh_file)
  t_mh_sparse <- mh_obj$time
} else {
  t_mh_sparse <- system.time({
    fit_mh_sp <- bvs_mh(
      X2, y2,
      adj_type = "ggm", sparse = hd_sparse,
      niter = niter_sim2, burnin = burnin_sim2,
      mu = hpars_sim2$mu, n_mh_gamma = sim2_n_mh_gamma, thin = thin_sim2,
      store_beta = TRUE, store_gamma = TRUE, store_Z_list = FALSE, store_Z_pip = TRUE
    )
  })
  saveRDS(list(fit = fit_mh_sp, time = t_mh_sparse), sim2_mh_file)
  rm(fit_mh_sp)
}
gc()
```

```{r sim2-run-pg-sparse, cache=FALSE}
if (file.exists(sim2_pg_file)) {
  pg_obj <- readRDS(sim2_pg_file)
  t_pg_sparse <- pg_obj$time
} else {
  t_pg_sparse <- system.time({
    fit_pg_sp <- bvs_pg(
      X2, y2,
      adj_type = "ggm", sparse = hd_sparse,
      niter = niter_sim2, burnin = burnin_sim2,
      mu = hpars_sim2$mu, n_mh_gamma = sim2_n_mh_gamma, thin = thin_sim2,
      store_beta = TRUE, store_gamma = TRUE, store_Z_list = FALSE, store_Z_pip = TRUE
    )
  })
  saveRDS(list(fit = fit_pg_sp, time = t_pg_sparse), sim2_pg_file)
  rm(fit_pg_sp)
}
gc()
```

```{r sim2-run-mh-fixed, cache=FALSE}
if (file.exists(sim2_fx_file)) {
  fx_obj <- readRDS(sim2_fx_file)
  t_mh_fixed2 <- fx_obj$time
} else {
  t_mh_fixed2 <- system.time({
    fit_mh_fixed2 <- bvs_mh(
      X2, y2,
      adj_type = "fixed", adj_fixed = R_known2,
      niter = niter_sim2, burnin = burnin_sim2,
      mu = hpars_sim2$mu, thin = thin_sim2
    )
  })
  saveRDS(list(fit = fit_mh_fixed2, time = t_mh_fixed2), sim2_fx_file)
  rm(fit_mh_fixed2)
}
gc()
```

```{r sim2-load-fits, cache=FALSE}
mh_obj <- readRDS(sim2_mh_file)
pg_obj <- readRDS(sim2_pg_file)
fx_obj <- readRDS(sim2_fx_file)

fit_mh_sp <- mh_obj$fit
fit_pg_sp <- pg_obj$fit
fit_mh_fixed2 <- fx_obj$fit
t_mh_sparse <- mh_obj$time
t_pg_sparse <- pg_obj$time
t_mh_fixed2 <- fx_obj$time

rm(mh_obj, pg_obj, fx_obj)
gc()
```

## Running time comparison

```{r sim2-timing}
timing_df2 <- data.frame(
  Method     = c("MH (sparse GGM)", "PG (sparse GGM)", "MH (fixed adj)"),
  Elapsed    = round(c(t_mh_sparse["elapsed"], t_pg_sparse["elapsed"], t_mh_fixed2["elapsed"]), 1),
  Iter_sec   = round(niter_sim2 / c(t_mh_sparse["elapsed"], t_pg_sparse["elapsed"], t_mh_fixed2["elapsed"]), 0)
)
knitr::kable(
  timing_df2,
  caption = sprintf(
    "Running time comparison ($n = 200, p = 1000$, %s post-burn-in iterations, burn-in %s, thin = %s).",
    format(niter_sim2, big.mark = ","),
    format(burnin_sim2, big.mark = ","),
    thin_sim2
  ),
  col.names = c("Method", "Elapsed (sec)", "Iter/sec")
)
```

## MCMC diagnostics

### Trace plots for strongest signals

```{r sim2-trace, fig.height=8, fig.width=8, fig.cap="Trace plots of the top 4 coefficients from the sparse high-dimensional simulation. Dashed red lines show true values."}
par(mfrow = c(3, 4), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
vars2 <- 1:4
for (method_name in c("MH (sparse GGM)", "PG (sparse GGM)", "MH (fixed)")) {
  fit_cur <- switch(method_name,
    "MH (sparse GGM)" = fit_mh_sp,
    "PG (sparse GGM)" = fit_pg_sp,
    "MH (fixed)" = fit_mh_fixed2
  )
  beta_cur <- get_beta_matrix(fit_cur)
  for (j in vars2) {
    plot(beta_cur[, j],
      type = "l", col = "steelblue",
      main = paste0(method_name, ": beta_", j),
      xlab = "Iteration", ylab = expression(beta[j]),
      cex.main = 0.8
    )
    abline(h = beta_true2[j], col = "red", lty = 2, lwd = 1.5)
  }
}
```

### Model size (number of included variables per iteration)

```{r sim2-modelsize, fig.height=4, fig.width=8, fig.cap="Trace of model size (number of included variables) per MCMC iteration in the high-dimensional setting."}
par(mfrow = c(1, 3), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
plot(model_size_chain(fit_mh_sp),
  type = "l", col = "darkblue",
  main = "MH (sparse GGM)", xlab = "Iteration", ylab = "Model size"
)
abline(h = p_star2, col = "red", lty = 2)

plot(model_size_chain(fit_pg_sp),
  type = "l", col = "darkblue",
  main = "PG (sparse GGM)", xlab = "Iteration", ylab = "Model size"
)
abline(h = p_star2, col = "red", lty = 2)

plot(model_size_chain(fit_mh_fixed2),
  type = "l", col = "darkblue",
  main = "MH (fixed adj)", xlab = "Iteration", ylab = "Model size"
)
abline(h = p_star2, col = "red", lty = 2)
```

### ACF for sparse sampler

```{r sim2-acf, fig.height=4, fig.width=8, fig.cap="ACF of beta_1 for the three methods in the high-dimensional setting."}
par(mfrow = c(1, 3), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
safe_acf_plot(get_beta_matrix(fit_mh_sp)[, 1], lag.max = 50, main = "MH (sparse GGM)", col = "steelblue")
safe_acf_plot(get_beta_matrix(fit_pg_sp)[, 1], lag.max = 50, main = "PG (sparse GGM)", col = "steelblue")
safe_acf_plot(get_beta_matrix(fit_mh_fixed2)[, 1], lag.max = 50, main = "MH (fixed adj)", col = "steelblue")
```

### Gelman-Rubin diagnostics (2 chains, 2 seeds)

```{r sim2-gelman, cache=TRUE}
vars_gr2 <- 1:4
seed_sim2 <- c(20261201L, 20261202L)

# Use a quick sparse approximation of S_ggm for two-chain diagnostics in p >> n.
S_ggm_gr2 <- Matrix::crossprod(X2)
S2_dense <- as.matrix(S_ggm_gr2)
diag2 <- diag(S2_dense)
thr2 <- as.numeric(stats::quantile(abs(S2_dense[upper.tri(S2_dense)]), probs = 0.90, na.rm = TRUE))
S2_dense[abs(S2_dense) < thr2] <- 0
diag(S2_dense) <- diag2
S_ggm_gr2 <- Matrix::Matrix(S2_dense, sparse = TRUE)

set.seed(seed_sim2[1])
fit_mh_sp_gr1 <- bvs_mh(
  X2, y2,
  adj_type = "ggm", sparse = TRUE, S_ggm = S_ggm_gr2,
  niter = niter_sim2, burnin = burnin_sim2,
  mu = hpars_sim2$mu, n_mh_gamma = sim2_n_mh_gamma, thin = thin_sim2,
  store_beta = TRUE, store_gamma = TRUE, store_Z_list = FALSE, store_Z_pip = TRUE
)
set.seed(seed_sim2[2])
fit_mh_sp_gr2 <- bvs_mh(
  X2, y2,
  adj_type = "ggm", sparse = TRUE, S_ggm = S_ggm_gr2,
  niter = niter_sim2, burnin = burnin_sim2,
  mu = hpars_sim2$mu, n_mh_gamma = sim2_n_mh_gamma, thin = thin_sim2,
  store_beta = TRUE, store_gamma = TRUE, store_Z_list = FALSE, store_Z_pip = TRUE
)

set.seed(seed_sim2[1])
fit_pg_sp_gr1 <- bvs_pg(
  X2, y2,
  adj_type = "ggm", sparse = TRUE, S_ggm = S_ggm_gr2,
  niter = niter_sim2, burnin = burnin_sim2,
  mu = hpars_sim2$mu, n_mh_gamma = sim2_n_mh_gamma, thin = thin_sim2,
  store_beta = TRUE, store_gamma = TRUE, store_Z_list = FALSE, store_Z_pip = TRUE
)
set.seed(seed_sim2[2])
fit_pg_sp_gr2 <- bvs_pg(
  X2, y2,
  adj_type = "ggm", sparse = TRUE, S_ggm = S_ggm_gr2,
  niter = niter_sim2, burnin = burnin_sim2,
  mu = hpars_sim2$mu, n_mh_gamma = sim2_n_mh_gamma, thin = thin_sim2,
  store_beta = TRUE, store_gamma = TRUE, store_Z_list = FALSE, store_Z_pip = TRUE
)

set.seed(seed_sim2[1])
fit_mh_fx_gr1 <- bvs_mh(
  X2, y2,
  adj_type = "fixed", adj_fixed = R_known2,
  niter = niter_sim2, burnin = burnin_sim2,
  mu = hpars_sim2$mu, thin = thin_sim2
)
set.seed(seed_sim2[2])
fit_mh_fx_gr2 <- bvs_mh(
  X2, y2,
  adj_type = "fixed", adj_fixed = R_known2,
  niter = niter_sim2, burnin = burnin_sim2,
  mu = hpars_sim2$mu, thin = thin_sim2
)

gelman_sim2 <- list(
  "MH (sparse GGM)" = safe_gelman_df(list(fit_mh_sp_gr1, fit_mh_sp_gr2), vars = vars_gr2, autoburnin = FALSE),
  "PG (sparse GGM)" = safe_gelman_df(list(fit_pg_sp_gr1, fit_pg_sp_gr2), vars = vars_gr2, autoburnin = FALSE),
  "MH (fixed adj)" = safe_gelman_df(list(fit_mh_fx_gr1, fit_mh_fx_gr2), vars = vars_gr2, autoburnin = FALSE)
)

gelman_sim2_summary <- do.call(rbind, lapply(names(gelman_sim2), function(mn) {
  gelman_method_summary(gelman_sim2[[mn]], mn)
}))

gelman_sim2_detail <- do.call(rbind, lapply(names(gelman_sim2), function(mn) {
  df <- gelman_sim2[[mn]]
  data.frame(
    Method = mn,
    Parameter = df$Parameter,
    Point_Rhat = round(df$Point_est, 3),
    Upper_CI_Rhat = round(df$Upper_CI, 3),
    stringsAsFactors = FALSE
  )
}))

knitr::kable(
  gelman_sim2_summary,
  caption = "Simulation 2 Gelman-Rubin diagnostics using two chains (two seeds) per method."
)
knitr::kable(
  gelman_sim2_detail,
  caption = "Simulation 2 PSRF details (Alpha and beta_1 to beta_4)."
)
```

```{r sim2-gelman-trace-2chain, fig.height=10, fig.width=12, fig.cap="Simulation 2: two-chain trace plots for representative coefficients across Gelman-Rubin diagnostic runs."}
beta_idx_gr2_trace <- unique(c(1L, 3L, 10L, min(p2, p_star2 + 35L)))
par(
  mfrow = c(3, length(beta_idx_gr2_trace)),
  mar = c(3, 3, 2, 1),
  mgp = c(1.8, 0.5, 0)
)
plot_two_chain_trace_panels(fit_mh_sp_gr1, fit_mh_sp_gr2, beta_idx_gr2_trace, "MH (sparse GGM)")
plot_two_chain_trace_panels(fit_pg_sp_gr1, fit_pg_sp_gr2, beta_idx_gr2_trace, "PG (sparse GGM)")
plot_two_chain_trace_panels(fit_mh_fx_gr1, fit_mh_fx_gr2, beta_idx_gr2_trace, "MH (fixed)")
```

## Variable selection performance

### Posterior inclusion probabilities

```{r sim2-pip, fig.height=5, fig.width=9, fig.cap="Posterior inclusion probabilities for the high-dimensional setting ($p = 1000$). Red bars: truly active variables (1--15). Most noise variables (16--1000) are correctly excluded."}
pip_mh_sp <- compute_pip(fit_mh_sp)
pip_pg_sp <- compute_pip(fit_pg_sp)
pip_fixed2 <- compute_pip(fit_mh_fixed2)

# Show first 50 variables + any with PIP > 0.1 beyond
show_vars <- unique(c(1:50, which(pip_mh_sp > 0.1), which(pip_pg_sp > 0.1), which(pip_fixed2 > 0.1)))
show_vars <- sort(show_vars)

par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
cols2 <- ifelse(show_vars <= p_star2, "firebrick", "grey60")

barplot(pip_mh_sp[show_vars],
  col = cols2, border = NA,
  main = "MH (sparse GGM)",
  xlab = "Variable", ylab = "PIP", ylim = c(0, 1),
  names.arg = show_vars, las = 2, cex.names = 0.5
)
abline(h = pip_threshold_sim2, lty = 2)

barplot(pip_pg_sp[show_vars],
  col = cols2, border = NA,
  main = "PG (sparse GGM)",
  xlab = "Variable", ylab = "PIP", ylim = c(0, 1),
  names.arg = show_vars, las = 2, cex.names = 0.5
)
abline(h = 0.5, lty = 2)

barplot(pip_fixed2[show_vars],
  col = cols2, border = NA,
  main = "MH (fixed adj)",
  xlab = "Variable", ylab = "PIP", ylim = c(0, 1),
  names.arg = show_vars, las = 2, cex.names = 0.5
)
abline(h = 0.5, lty = 2)
```

### Selection metrics

```{r sim2-metrics}
metrics2 <- rbind(
  cbind(Method = "MH (sparse GGM)", compute_metrics(pip_mh_sp, threshold = pip_threshold_sim2, p_star = p_star2)),
  cbind(Method = "PG (sparse GGM)", compute_metrics(pip_pg_sp, threshold = pip_threshold_sim2, p_star = p_star2)),
  cbind(Method = "MH (fixed adj)", compute_metrics(pip_fixed2, threshold = pip_threshold_sim2, p_star = p_star2))
)
knitr::kable(metrics2,
  caption = sprintf("Variable selection performance ($n = 200, p = 1000$, PIP threshold %.2f).", pip_threshold_sim2),
  row.names = FALSE
)
```

## Additional adjacency-type comparison (5 methods) using 50% true external edges

To compare adjacency types in the high-dimensional setup, we run five additional
MH models using the same `X2`, `y2`, and MCMC settings:

1. `adj_type = "fixed"` with external `R_ext50_2`,
2. `adj_type = "ggm"` (sparse Bayesian GGM),
3. `adj_type = "glasso"` (estimated from data),
4. `adj_type = "glasso_fixed"` (external `R_ext50_2` + glasso),
5. `adj_type = "ggm_fixed"` (external `R_ext50_2` + sparse Bayesian GGM).

```{r sim2-adjtype-run, cache=TRUE}
sim2_adj_dir <- "sim2_adjtype_saved_runs"
if (!dir.exists(sim2_adj_dir)) dir.create(sim2_adj_dir, recursive = TRUE)

niter_adj2 <- niter_sim2
burnin_adj2 <- burnin_sim2
thin_adj2 <- thin_sim2

sim2_f_fixed <- file.path(sim2_adj_dir, sprintf("sim2_fixed50_n%d_b%d_t%d.rds", niter_adj2, burnin_adj2, thin_adj2))
sim2_f_ggm <- file.path(sim2_adj_dir, sprintf("sim2_ggm_sparse_n%d_b%d_t%d.rds", niter_adj2, burnin_adj2, thin_adj2))
sim2_f_gls <- file.path(sim2_adj_dir, sprintf("sim2_glasso_n%d_b%d_t%d.rds", niter_adj2, burnin_adj2, thin_adj2))
sim2_f_gf <- file.path(sim2_adj_dir, sprintf("sim2_glasso_fixed_n%d_b%d_t%d.rds", niter_adj2, burnin_adj2, thin_adj2))
sim2_f_ggmf <- file.path(sim2_adj_dir, sprintf("sim2_ggm_fixed_sparse_n%d_b%d_t%d.rds", niter_adj2, burnin_adj2, thin_adj2))
sim2_f_sggm <- file.path(sim2_adj_dir, sprintf("sim2_Sggm_q90_n%d_p%d.rds", nrow(X2), ncol(X2)))
sim2_f_gladj <- file.path(sim2_adj_dir, "sim2_glasso_adj_ebic_cache1.rds")

cat("[sim2-adjtype] precompute S_ggm ...\n")
if (file.exists(sim2_f_sggm)) {
  t2_pre_s <- c(elapsed = 0)
  S_ggm_sim2 <- readRDS(sim2_f_sggm)
} else {
  t2_pre_s <- system.time({
    S_ggm_sim2 <- Matrix::crossprod(X2)
    if (isTRUE(quick_check)) {
      S_dense <- as.matrix(S_ggm_sim2)
      diag_s <- diag(S_dense)
      off_abs <- abs(S_dense[upper.tri(S_dense)])
      s_thr <- as.numeric(stats::quantile(off_abs, probs = 0.90, na.rm = TRUE))
      S_dense[abs(S_dense) < s_thr] <- 0
      diag(S_dense) <- diag_s
      S_ggm_sim2 <- Matrix::Matrix(S_dense, sparse = TRUE)
      cat(sprintf("[sim2-adjtype] quick-check S_ggm threshold = %.4f (keep top ~10%% off-diagonals)\n", s_thr))
    }
  })
  saveRDS(S_ggm_sim2, sim2_f_sggm)
}
cat(sprintf("[sim2-adjtype] precompute S_ggm done (elapsed %.2f sec)\n", as.numeric(t2_pre_s["elapsed"])))

cat("[sim2-adjtype] precompute glasso adjacency ...\n")
if (file.exists(sim2_f_gladj)) {
  t2_pre_gls <- c(elapsed = 0)
  R_glasso2 <- readRDS(sim2_f_gladj)
} else {
  t2_pre_gls <- system.time({
    R_glasso2 <- estimate_glasso_adj(X2, criterion = "ebic")
  })
  saveRDS(R_glasso2, sim2_f_gladj)
}
cat(sprintf("[sim2-adjtype] precompute glasso adjacency done (elapsed %.2f sec)\n", as.numeric(t2_pre_gls["elapsed"])))

cat("[sim2-adjtype] fixed ...\n")
if (file.exists(sim2_f_fixed)) {
  o <- readRDS(sim2_f_fixed)
  fit2_fixed50 <- o$fit
  t2_fixed50 <- o$time
} else {
  t2_fixed50 <- system.time({
    fit2_fixed50 <- bvs_mh(
      X2, y2,
      adj_type = "fixed", adj_fixed = R_ext50_2,
      niter = niter_adj2, burnin = burnin_adj2, thin = thin_adj2,
      mu = hpars_sim2$mu, n_mh_gamma = sim2_n_mh_gamma
    )
  })
  saveRDS(list(fit = fit2_fixed50, time = t2_fixed50), sim2_f_fixed)
}
cat(sprintf("[sim2-adjtype] fixed done (elapsed %.2f sec)\n", as.numeric(t2_fixed50["elapsed"])))

cat("[sim2-adjtype] ggm (sparse) ...\n")
if (file.exists(sim2_f_ggm)) {
  o <- readRDS(sim2_f_ggm)
  fit2_ggm <- o$fit
  t2_ggm <- o$time
} else {
  t2_ggm <- system.time({
    fit2_ggm <- bvs_mh(
      X2, y2,
      adj_type = "ggm",
      sparse = hd_sparse,
      S_ggm = S_ggm_sim2,
      niter = niter_adj2, burnin = burnin_adj2, thin = thin_adj2,
      mu = hpars_sim2$mu, n_mh_gamma = sim2_n_mh_gamma,
      store_beta = TRUE, store_gamma = TRUE, store_Z_list = FALSE, store_Z_pip = TRUE
    )
  })
  saveRDS(list(fit = fit2_ggm, time = t2_ggm), sim2_f_ggm)
}
cat(sprintf("[sim2-adjtype] ggm (sparse) done (elapsed %.2f sec)\n", as.numeric(t2_ggm["elapsed"])))

cat("[sim2-adjtype] glasso ...\n")
if (file.exists(sim2_f_gls)) {
  o <- readRDS(sim2_f_gls)
  fit2_glasso <- o$fit
  t2_glasso <- o$time
} else {
  t2_glasso <- system.time({
    fit2_glasso <- bvs_mh(
      X2, y2,
      adj_type = "fixed", adj_fixed = R_glasso2,
      niter = niter_adj2, burnin = burnin_adj2, thin = thin_adj2,
      mu = hpars_sim2$mu, n_mh_gamma = sim2_n_mh_gamma
    )
  })
  fit2_glasso$adj_type <- "glasso"
  saveRDS(list(fit = fit2_glasso, time = t2_glasso), sim2_f_gls)
}
cat(sprintf("[sim2-adjtype] glasso done (elapsed %.2f sec)\n", as.numeric(t2_glasso["elapsed"])))

cat("[sim2-adjtype] glasso_fixed ...\n")
if (file.exists(sim2_f_gf)) {
  o <- readRDS(sim2_f_gf)
  fit2_glasso_fixed <- o$fit
  t2_glasso_fixed <- o$time
} else {
  t2_glasso_fixed <- system.time({
    fit2_glasso_fixed <- bvs_mh(
      X2, y2,
      adj_type = "dual_fixed", adj_fixed = R_glasso2, adj_fixed2 = R_ext50_2,
      niter = niter_adj2, burnin = burnin_adj2, thin = thin_adj2,
      mu = hpars_sim2$mu, n_mh_gamma = sim2_n_mh_gamma,
      eta1_sd = hpars_sim2$eta1_sd, eta2_sd = hpars_sim2$eta2_sd,
      e_eta = hpars_sim2$e_eta, f_eta = hpars_sim2$f_eta,
      mu_tilde = hpars_sim2$mu_tilde,
      eta1_tilde = hpars_sim2$eta1_tilde, eta2_tilde = hpars_sim2$eta2_tilde
    )
  })
  fit2_glasso_fixed$adj_type <- "glasso_fixed"
  saveRDS(list(fit = fit2_glasso_fixed, time = t2_glasso_fixed), sim2_f_gf)
}
cat(sprintf("[sim2-adjtype] glasso_fixed done (elapsed %.2f sec)\n", as.numeric(t2_glasso_fixed["elapsed"])))

cat("[sim2-adjtype] ggm_fixed (sparse) ...\n")
if (file.exists(sim2_f_ggmf)) {
  o <- readRDS(sim2_f_ggmf)
  fit2_ggm_fixed <- o$fit
  t2_ggm_fixed <- o$time
} else {
  t2_ggm_fixed <- system.time({
    fit2_ggm_fixed <- bvs_mh(
      X2, y2,
      adj_type = "ggm_fixed", adj_fixed = R_ext50_2,
      sparse = hd_sparse,
      S_ggm = S_ggm_sim2,
      niter = niter_adj2, burnin = burnin_adj2, thin = thin_adj2,
      mu = hpars_sim2$mu, n_mh_gamma = sim2_n_mh_gamma,
      eta1_sd = hpars_sim2$eta1_sd, eta2_sd = hpars_sim2$eta2_sd,
      e_eta = hpars_sim2$e_eta, f_eta = hpars_sim2$f_eta,
      mu_tilde = hpars_sim2$mu_tilde,
      eta1_tilde = hpars_sim2$eta1_tilde, eta2_tilde = hpars_sim2$eta2_tilde,
      store_beta = TRUE, store_gamma = TRUE, store_Z_list = FALSE, store_Z_pip = TRUE
    )
  })
  saveRDS(list(fit = fit2_ggm_fixed, time = t2_ggm_fixed), sim2_f_ggmf)
}
cat(sprintf("[sim2-adjtype] ggm_fixed (sparse) done (elapsed %.2f sec)\n", as.numeric(t2_ggm_fixed["elapsed"])))
```

```{r sim2-adjtype-timing}
timing_adj2 <- data.frame(
  Method = c("Fixed (50% true edges)", "Sparse GGM", "Glasso", "Glasso + Fixed", "Sparse GGM + Fixed"),
  Elapsed_sec = round(c(
    t2_fixed50["elapsed"], t2_ggm["elapsed"], t2_glasso["elapsed"],
    t2_glasso_fixed["elapsed"], t2_ggm_fixed["elapsed"]
  ), 2),
  Iter_per_sec = round(niter_adj2 / c(
    t2_fixed50["elapsed"], t2_ggm["elapsed"], t2_glasso["elapsed"],
    t2_glasso_fixed["elapsed"], t2_ggm_fixed["elapsed"]
  ), 1)
)
knitr::kable(
  timing_adj2,
  caption = "Simulation 2 adjacency-type comparison timing (fixed input uses random 50% of true adjacency edges).",
  row.names = FALSE
)
```

```{r sim2-adjtype-pip, fig.height=8, fig.width=10, fig.cap="Simulation 2: PIP comparison across adjacency types (fixed, ggm, glasso, glasso_fixed, ggm_fixed)."}
pip2_fixed50 <- compute_pip(fit2_fixed50)
pip2_ggm <- compute_pip(fit2_ggm)
pip2_glasso <- compute_pip(fit2_glasso)
pip2_gf <- compute_pip(fit2_glasso_fixed)
pip2_ggmf <- compute_pip(fit2_ggm_fixed)

show_vars2_adj <- unique(c(
  1:50,
  which(pip2_fixed50 > 0.1),
  which(pip2_ggm > 0.1),
  which(pip2_glasso > 0.1),
  which(pip2_gf > 0.1),
  which(pip2_ggmf > 0.1)
))
show_vars2_adj <- sort(show_vars2_adj)
cols2_adj <- ifelse(show_vars2_adj <= p_star2, "firebrick", "grey65")

pip_list2 <- list(
  "Fixed (50% true)" = pip2_fixed50,
  "Sparse GGM" = pip2_ggm,
  "Glasso" = pip2_glasso,
  "Glasso + Fixed" = pip2_gf,
  "Sparse GGM + Fixed" = pip2_ggmf
)

par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))
for (mn in names(pip_list2)) {
  barplot(pip_list2[[mn]][show_vars2_adj],
    col = cols2_adj, border = NA, ylim = c(0, 1),
    main = mn, xlab = "Variable", ylab = "PIP",
    names.arg = show_vars2_adj, las = 2, cex.names = 0.45
  )
  abline(h = pip_threshold_sim2, lty = 2)
}
plot.new()
```

```{r sim2-adjtype-metrics}
metrics_adj2 <- rbind(
  cbind(Method = "Fixed (50% true)", compute_metrics(pip2_fixed50, threshold = pip_threshold_sim2, p_star = p_star2)),
  cbind(Method = "Sparse GGM", compute_metrics(pip2_ggm, threshold = pip_threshold_sim2, p_star = p_star2)),
  cbind(Method = "Glasso", compute_metrics(pip2_glasso, threshold = pip_threshold_sim2, p_star = p_star2)),
  cbind(Method = "Glasso + Fixed", compute_metrics(pip2_gf, threshold = pip_threshold_sim2, p_star = p_star2)),
  cbind(Method = "Sparse GGM + Fixed", compute_metrics(pip2_ggmf, threshold = pip_threshold_sim2, p_star = p_star2))
)
knitr::kable(
  metrics_adj2,
  caption = sprintf("Simulation 2 variable-selection performance for five adjacency-type methods (PIP threshold %.2f).", pip_threshold_sim2),
  row.names = FALSE
)
```

```{r sim2-adjtype-diagnostics}
if (!exists("safe_ess_sim", mode = "function")) {
  safe_ess_sim <- function(x, lag.max = 200L) {
    x <- as.numeric(x)
    x <- x[is.finite(x)]
    n <- length(x)
    if (n < 3L) {
      return(NA_real_)
    }
    v <- stats::var(x)
    if (!is.finite(v) || v <= .Machine$double.eps) {
      return(n)
    }
    ac <- as.numeric(stats::acf(x, lag.max = min(lag.max, n - 1L), plot = FALSE)$acf)[-1L]
    if (!length(ac)) {
      return(n)
    }
    first_nonpos <- which(ac <= 0)
    m <- if (length(first_nonpos)) first_nonpos[1L] - 1L else length(ac)
    if (m < 1L) {
      return(n)
    }
    tau <- 1 + 2 * sum(ac[seq_len(m)])
    if (!is.finite(tau) || tau <= 0) {
      return(n)
    }
    min(n, n / tau)
  }
}
if (!exists("lag1_acf_sim", mode = "function")) {
  lag1_acf_sim <- function(x) {
    x <- as.numeric(x)
    x <- x[is.finite(x)]
    if (length(x) < 3L) {
      return(NA_real_)
    }
    v <- stats::var(x)
    if (!is.finite(v) || v <= .Machine$double.eps) {
      return(NA_real_)
    }
    as.numeric(stats::acf(x, lag.max = 1L, plot = FALSE)$acf[2L])
  }
}

fits_adj2 <- list(
  "Fixed (50% true)" = fit2_fixed50,
  "Sparse GGM" = fit2_ggm,
  "Glasso" = fit2_glasso,
  "Glasso + Fixed" = fit2_glasso_fixed,
  "Sparse GGM + Fixed" = fit2_ggm_fixed
)
times_adj2 <- c(
  "Fixed (50% true)" = as.numeric(t2_fixed50["elapsed"]),
  "Sparse GGM" = as.numeric(t2_ggm["elapsed"]),
  "Glasso" = as.numeric(t2_glasso["elapsed"]),
  "Glasso + Fixed" = as.numeric(t2_glasso_fixed["elapsed"]),
  "Sparse GGM + Fixed" = as.numeric(t2_ggm_fixed["elapsed"])
)

pip_mat_adj2 <- sapply(fits_adj2, compute_pip)
top_idx_adj2 <- order(rowMeans(pip_mat_adj2), decreasing = TRUE)[1L]
top_name_adj2 <- if (!is.null(colnames(X2))) colnames(X2)[top_idx_adj2] else paste0("V", top_idx_adj2)

diag_adj2 <- do.call(rbind, lapply(names(fits_adj2), function(mn) {
  fit <- fits_adj2[[mn]]
  pip <- compute_pip(fit)
  beta_chain <- get_beta_matrix(fit)[, top_idx_adj2]
  ms <- model_size_chain(fit)
  ess <- safe_ess_sim(beta_chain)
  n_saved <- length(ms)
  iter_sec <- n_saved / max(as.numeric(times_adj2[mn]), 1e-8)
  data.frame(
    Method = mn,
    Runtime_sec = round(times_adj2[mn], 2),
    N_saved = n_saved,
    Iter_per_sec = round(iter_sec, 2),
    Mean_model_size = round(mean(ms), 2),
    SD_model_size = round(stats::sd(ms), 2),
    Median_model_size = round(stats::median(ms), 2),
    Q05_model_size = round(as.numeric(stats::quantile(ms, 0.05)), 2),
    Q95_model_size = round(as.numeric(stats::quantile(ms, 0.95)), 2),
    Top_gene_for_diag = top_name_adj2,
    PIP_top_gene = round(pip[top_idx_adj2], 3),
    Lag1_ACF_top_beta = round(lag1_acf_sim(beta_chain), 3),
    ESS_top_beta = round(ess, 1),
    ESS_per_sec = round(ess / max(as.numeric(times_adj2[mn]), 1e-8), 3),
    MCSE_top_beta = round(stats::sd(beta_chain) / sqrt(max(1, ess)), 4),
    stringsAsFactors = FALSE
  )
}))

knitr::kable(
  diag_adj2,
  caption = "Simulation 2 MCMC convergence diagnostics and sampling performance for five adjacency-type methods.",
  row.names = FALSE
)
```

```{r sim2-adjtype-trace, fig.height=8, fig.width=10, fig.cap="Simulation 2 adjacency-type comparison: trace plots of the representative top-PIP coefficient."}
par(mfrow = c(2, 3), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (mn in names(fits_adj2)) {
  beta_chain <- get_beta_matrix(fits_adj2[[mn]])[, top_idx_adj2]
  plot(beta_chain,
    type = "l", col = "steelblue",
    main = paste0(mn, ": ", top_name_adj2),
    xlab = "Iteration", ylab = "beta", cex.main = 0.85
  )
}
plot.new()
```

```{r sim2-adjtype-acf, fig.height=8, fig.width=10, fig.cap="Simulation 2 adjacency-type comparison: ACF of the representative top-PIP coefficient."}
par(mfrow = c(2, 3), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (mn in names(fits_adj2)) {
  beta_chain <- get_beta_matrix(fits_adj2[[mn]])[, top_idx_adj2]
  safe_acf_plot(beta_chain,
    lag.max = 50,
    main = paste0(mn, ": ACF(", top_name_adj2, ")"),
    col = "steelblue"
  )
}
plot.new()
```

# Summary of running time comparisons {#sec-runtime}

```{r runtime-summary-table}
standardize_timing <- function(df) {
  out <- data.frame(
    Method = as.character(df$Method),
    Elapsed_sec = NA_real_,
    Iter_per_sec = NA_real_,
    stringsAsFactors = FALSE
  )
  if ("Elapsed_sec" %in% names(df)) out$Elapsed_sec <- as.numeric(df$Elapsed_sec)
  if ("Elapsed" %in% names(df)) out$Elapsed_sec <- as.numeric(df$Elapsed)
  if ("Iter_per_sec" %in% names(df)) out$Iter_per_sec <- as.numeric(df$Iter_per_sec)
  if ("Iter_sec" %in% names(df)) out$Iter_per_sec <- as.numeric(df$Iter_sec)
  out
}

t1 <- standardize_timing(timing_df)
t2 <- standardize_timing(timing_df2)

combined_timing <- rbind(
  cbind(Setting = "Moderate (n=500, p=100)", t1),
  cbind(Setting = "High-dim (n=200, p=1000)", t2)
)
knitr::kable(combined_timing,
  caption = "Complete running time comparison across both simulation settings.",
  row.names = FALSE
)
```

```{r runtime-barplot, fig.height=5, fig.width=8, fig.cap="Running time (seconds) across all simulation methods and settings."}
par(mar = c(8, 4, 2, 1))
all_times <- c(
  t_mh["elapsed"], t_pg["elapsed"], t_pg_ggm["elapsed"],
  t_mh_sparse["elapsed"], t_pg_sparse["elapsed"], t_mh_fixed2["elapsed"]
)
all_names <- c(
  "MH fixed\n(n=500,p=100)", "PG fixed\n(n=500,p=100)",
  "PG GGM\n(n=500,p=100)",
  "MH sparse\n(n=200,p=1000)", "PG sparse\n(n=200,p=1000)",
  "MH fixed\n(n=200,p=1000)"
)
all_cols <- c(
  "steelblue", "coral", "seagreen",
  "steelblue", "coral", "gold3"
)

bp <- barplot(all_times,
  col = all_cols, border = NA,
  main = "Running Time Comparison",
  ylab = "Elapsed time (seconds)",
  names.arg = all_names, las = 2, cex.names = 0.75
)
text(bp, all_times + max(all_times) * 0.03,
  labels = paste0(round(all_times, 1), "s"), cex = 0.8
)
```

# Real-data example: Leukemia microarray with KEGG-informed dual adjacency {#sec-real}

To illustrate usage on real data, we use the Golub leukemia microarray dataset
[@Golub1999], a standard benchmark in genomic binary classification.

### Dataset profile (Golub leukemia)

- **Study year**: 1999 [@Golub1999].
- **Original study context**: oligonucleotide microarray profiling for acute
  leukemia subtype classification (ALL vs AML).
- **Reference scale in the original publication**: 72 patient samples and
  thousands of probe sets (widely cited as 7,129) [@Golub1999].
- **Data object used here**: the processed `multtest` benchmark objects
  (`golub`, `golub.cl`, `golub.gnames`). The exact dimensions used in your
  local installation are printed in the setup/inspection chunk below.

In this example:

- **Outcome variable (`y`)**: leukemia subtype indicator (`1 = ALL`, `0 = AML`).
- **Predictor variables (`X`)**: gene-expression intensities.
- **Variable-selection target**: identify which genes are associated with ALL vs AML in a logistic BVS model.

For vignette-scale runtime we retain the top 300 most variable genes and compare
five network-prior configurations:

1. Single-source `glasso` adjacency.
2. Single-source `ggm` adjacency.
3. Single-source `fixed` adjacency using KEGG only.
4. Dual-source `glasso_fixed` (external KEGG + glasso-estimated).
5. Dual-source `ggm_fixed` (external KEGG + Bayesian GGM-estimated).

The external biological adjacency is built from KEGG pathway co-membership
[@Kanehisa2023]: two genes are connected if they are annotated to at least one
common KEGG pathway.

### Variable description (Golub objects)

- `golub`: raw expression matrix provided by `multtest` (genes/probes x subjects).
- `golub.cl`: class label vector for each subject (ALL/AML coding).
- `golub.gnames`: gene/probe annotation table (probe IDs, gene symbols and related metadata).
- `X_raw`: transposed analysis matrix (`subjects x predictors`) built from `golub`.
- `y_real`: binary outcome for logistic BVS (`1 = ALL`, `0 = AML`).
- `X_real`: standardized, variance-filtered analysis matrix retaining top variable predictors.
- `gene_symbols`: predictor labels used as column names in `X_real`.

## Data setup and preprocessing

```{r realdata-setup, eval=TRUE}
# Required packages:
# multtest, AnnotationDbi, org.Hs.eg.db, KEGGREST
# Optional annotation package: hu6800.db (falls back to probe IDs when missing)
req_pkgs <- c("multtest")
missing_pkgs <- req_pkgs[!vapply(req_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs)) {
  stop("Install required packages first: ", paste(missing_pkgs, collapse = ", "))
}
if (!requireNamespace("hu6800.db", quietly = TRUE)) {
  message("Optional package 'hu6800.db' is not installed; using probe IDs as gene labels in Golub example.")
}
kegg_pkgs_real <- c("AnnotationDbi", "org.Hs.eg.db", "KEGGREST")
kegg_pkg_ok_real <- all(vapply(kegg_pkgs_real, requireNamespace, logical(1), quietly = TRUE))
if (!kegg_pkg_ok_real) {
  message(
    "KEGG annotation packages not fully available (",
    paste(kegg_pkgs_real[!vapply(kegg_pkgs_real, requireNamespace, logical(1), quietly = TRUE)], collapse = ", "),
    "). Using correlation-based external adjacency fallback for this test run."
  )
}

# Golub leukemia data (ALL vs AML)
data(golub, package = "multtest")
data(golub.cl, package = "multtest")
data(golub.gnames, package = "multtest")

X_raw <- t(golub) # n x p: samples x genes
y_real <- as.integer(golub.cl == 1) # ALL=1, AML=0

# Keep top variable genes for vignette-level run time
p_keep <- if (isTRUE(quick_check)) 120L else 300L
gene_var <- apply(X_raw, 2L, var)
keep_idx <- order(gene_var, decreasing = TRUE)[seq_len(p_keep)]
X_real <- scale(X_raw[, keep_idx, drop = FALSE])

# Map Affymetrix Hu6800 probe IDs to HGNC gene symbols via hu6800.db
probe_col <- if (ncol(golub.gnames) >= 3L) 3L else 2L
probe_ids_golub <- as.character(golub.gnames[keep_idx, probe_col])

if (requireNamespace("hu6800.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE)) {
  gene_symbols <- tryCatch(
    {
      sym <- AnnotationDbi::mapIds(
        hu6800.db::hu6800.db,
        keys = probe_ids_golub,
        keytype = "PROBEID",
        column = "SYMBOL",
        multiVals = "first"
      )
      as.character(sym)
    },
    error = function(e) {
      message("hu6800.db mapping failed: ", conditionMessage(e))
      rep(NA_character_, length(probe_ids_golub))
    }
  )
} else {
  message("hu6800.db not installed; using raw probe IDs as feature labels.")
  gene_symbols <- probe_ids_golub
}
# Fall back to probe IDs where symbol mapping is missing
gene_symbols[is.na(gene_symbols) | !nzchar(gene_symbols)] <-
  probe_ids_golub[is.na(gene_symbols) | !nzchar(gene_symbols)]
colnames(X_real) <- make.unique(gene_symbols)

cat(sprintf("Raw Golub data dimensions: n=%d, p=%d\n", nrow(X_raw), ncol(X_raw)))
cat(sprintf("Raw outcome counts: ALL=%d, AML=%d\n", sum(y_real == 1L), sum(y_real == 0L)))
cat(sprintf("Analysis subset dimensions after variance filtering: n=%d, p=%d\n", nrow(X_real), ncol(X_real)))
```

### Quick printout and summary (Golub)

```{r realdata-inspect, eval=TRUE}
cat("=== Golub: first 10 observations and summaries ===\n")

cat("\nFirst 10 rows of subject-level data (outcome + first 8 predictors from X_raw):\n")
print(head(
  data.frame(y = y_real, X_raw[, seq_len(min(8L, ncol(X_raw))), drop = FALSE]),
  10
))

cat("\nFirst 10 rows of annotation table (golub.gnames):\n")
print(head(as.data.frame(golub.gnames), 10))

cat("\nSummary of outcome variable (y_real):\n")
print(summary(as.factor(y_real)))

cat("\nSummary of full raw predictor matrix (overall values):\n")
print(summary(as.vector(X_raw)))

cat("\nSummary of first 10 predictors in X_raw:\n")
print(summary(as.data.frame(X_raw[, seq_len(min(10L, ncol(X_raw))), drop = FALSE])))
```

## KEGG-based external adjacency matrix

```{r realdata-kegg-adj, eval=TRUE}
fallback_real_adj <- function() {
  C_real <- abs(stats::cor(X_real, use = "pairwise.complete.obs"))
  diag(C_real) <- 0
  thr_real <- as.numeric(stats::quantile(C_real[upper.tri(C_real)], probs = 0.95, na.rm = TRUE))
  Rk <- matrix(as.integer(C_real >= thr_real), nrow(C_real), ncol(C_real))
  diag(Rk) <- 0L
  cat(sprintf(
    "Fallback external adjacency (|cor| >= %.3f) edges among retained genes: %d\n",
    thr_real, sum(Rk) / 2
  ))
  Rk
}

if (isTRUE(kegg_pkg_ok_real)) {
  build_kegg_adjacency <- function(symbols) {
    p <- length(symbols)

    # Map symbols -> Entrez IDs
    entrez <- tryCatch(
      AnnotationDbi::mapIds(
        org.Hs.eg.db::org.Hs.eg.db,
        keys = symbols,
        keytype = "SYMBOL",
        column = "ENTREZID",
        multiVals = "first"
      ),
      error = function(e) {
        return(rep(NA_character_, p))
      }
    )
    map_df <- data.frame(
      idx = seq_len(p),
      entrez = as.character(entrez),
      stringsAsFactors = FALSE
    )
    map_df <- map_df[!is.na(map_df$entrez) & nzchar(map_df$entrez), , drop = FALSE]

    if (nrow(map_df) == 0) {
      stop("No valid Entrez/SYMBOL mappings found for the retained features.")
    }

    # KEGG gene-pathway links
    kmap <- KEGGREST::keggLink("pathway", "hsa")
    if (!length(kmap)) stop("KEGG mapping retrieval failed.")

    if (startsWith(names(kmap)[1], "hsa:")) {
      gp <- data.frame(
        entrez = sub("^hsa:", "", names(kmap)),
        pathway = sub("^path:", "", unname(kmap)),
        stringsAsFactors = FALSE
      )
    } else {
      gp <- data.frame(
        entrez = sub("^hsa:", "", unname(kmap)),
        pathway = sub("^path:", "", names(kmap)),
        stringsAsFactors = FALSE
      )
    }

    hit <- merge(map_df, gp, by = "entrez")
    groups <- split(hit$idx, hit$pathway)

    R_kegg <- matrix(0L, p, p)
    for (g in groups) {
      g <- sort(unique(g))
      if (length(g) < 2L) next
      cmb <- utils::combn(g, 2L)
      R_kegg[cbind(cmb[1L, ], cmb[2L, ])] <- 1L
      R_kegg[cbind(cmb[2L, ], cmb[1L, ])] <- 1L
    }
    diag(R_kegg) <- 0L
    R_kegg
  }

  R_kegg <- tryCatch(
    build_kegg_adjacency(colnames(X_real)),
    error = function(e) {
      message("KEGG adjacency construction failed: ", conditionMessage(e))
      NULL
    }
  )
  if (is.null(R_kegg)) {
    R_kegg <- fallback_real_adj()
  } else {
    cat(sprintf("KEGG adjacency edges among retained genes: %d\n", sum(R_kegg) / 2))
  }
} else {
  R_kegg <- fallback_real_adj()
}
```

## Fit and compare five BVS models

```{r realdata-fit, eval=TRUE}
cfg_real <- mcmc_scale(
  full_niter = 500L, full_burnin = 200L, full_thin = 2L,
  quick_niter = 5L, quick_burnin = 2L, quick_thin = 1L
)
niter_real <- cfg_real$niter
burnin_real <- cfg_real$burnin
thin_real <- cfg_real$thin

real_n_mh_gamma <- if (isTRUE(quick_check)) 1L else 1L
real_tmax <- if (isTRUE(quick_check)) 8L else 16L
real_proposal_type <- 0L

real_cache_dir <- "realdata_saved_runs"
if (!dir.exists(real_cache_dir)) dir.create(real_cache_dir, recursive = TRUE)

real_ctrl_tag <- sprintf(
  "n%d_b%d_t%d_mh%d_T%d_p%d",
  niter_real, burnin_real, thin_real, real_n_mh_gamma, real_tmax, real_proposal_type
)

real_sggm_q <- if (isTRUE(quick_check)) 0.95 else 0.98
real_sggm_tag <- sprintf("q%02d", as.integer(round(100 * real_sggm_q)))
real_sggm_file <- file.path(
  real_cache_dir,
  sprintf("real_Sggm_%s_n%d_p%d.rds", real_sggm_tag, nrow(X_real), ncol(X_real))
)
if (file.exists(real_sggm_file)) {
  S_real_ggm <- readRDS(real_sggm_file)
} else {
  S_real_dense <- as.matrix(Matrix::crossprod(X_real))
  diag_real <- diag(S_real_dense)
  off_real <- abs(S_real_dense[upper.tri(S_real_dense)])
  s_real_thr <- as.numeric(stats::quantile(off_real, probs = real_sggm_q, na.rm = TRUE))
  S_real_dense[abs(S_real_dense) < s_real_thr] <- 0
  diag(S_real_dense) <- diag_real
  S_real_ggm <- Matrix::Matrix(S_real_dense, sparse = TRUE)
  saveRDS(S_real_ggm, real_sggm_file)
}
cat(sprintf(
  "[realdata-fit] sparse S_ggm ready (%s): nnz=%d, density=%.4f\n",
  real_sggm_tag, length(S_real_ggm@x), length(S_real_ggm@x) / (ncol(S_real_ggm)^2)
))

real_glasso_keep_ratio <- if (isTRUE(quick_check)) 0.20 else 0.10
real_glasso_max_degree <- if (isTRUE(quick_check)) 80L else 30L
real_glasso_tag <- sprintf(
  "spr%02d_deg%d",
  as.integer(round(100 * real_glasso_keep_ratio)),
  real_glasso_max_degree
)
real_glasso_raw_file <- file.path(
  real_cache_dir,
  sprintf("real_glasso_adj_ebic_n%d_p%d.rds", nrow(X_real), ncol(X_real))
)
real_glasso_sparse_file <- file.path(
  real_cache_dir,
  sprintf(
    "real_glasso_adj_ebic_%s_n%d_p%d.rds",
    real_glasso_tag, nrow(X_real), ncol(X_real)
  )
)
if (file.exists(real_glasso_raw_file)) {
  R_glasso_real_raw <- readRDS(real_glasso_raw_file)
} else {
  R_glasso_real_raw <- estimate_glasso_adj(X_real, criterion = "ebic")
  saveRDS(R_glasso_real_raw, real_glasso_raw_file)
}

if (!exists("sparsify_glasso_adj", mode = "function")) {
  sparsify_glasso_adj <- function(adj, score_mat, keep_ratio = 0.05, max_degree = 40L) {
    A <- (adj != 0) * 1L
    A <- ((A + t(A)) > 0) * 1L
    diag(A) <- 0L
    p <- nrow(A)

    e_idx <- which(upper.tri(A) & A == 1L, arr.ind = TRUE)
    n_edges <- nrow(e_idx)
    if (n_edges == 0L) {
      return(list(
        adj = A,
        stats = list(
          edges_before = 0L, edges_after = 0L,
          density_before = 0, density_after = 0,
          max_degree_before = 0L, max_degree_after = 0L
        )
      ))
    }

    keep_ratio <- min(max(as.numeric(keep_ratio), 0.001), 1)
    max_degree <- max(1L, as.integer(max_degree))
    target_edges <- max(1L, as.integer(ceiling(keep_ratio * n_edges)))

    sc <- score_mat[cbind(e_idx[, 1], e_idx[, 2])]
    sc[!is.finite(sc)] <- 0
    ord <- order(sc, decreasing = TRUE)
    e_ord <- e_idx[ord, , drop = FALSE]

    deg <- integer(p)
    keep <- logical(nrow(e_ord))
    kept <- 0L
    for (k in seq_len(nrow(e_ord))) {
      if (kept >= target_edges) break
      i <- e_ord[k, 1]
      j <- e_ord[k, 2]
      if (deg[i] < max_degree && deg[j] < max_degree) {
        keep[k] <- TRUE
        deg[i] <- deg[i] + 1L
        deg[j] <- deg[j] + 1L
        kept <- kept + 1L
      }
    }
    if (kept < target_edges) {
      for (k in seq_len(nrow(e_ord))) {
        if (kept >= target_edges) break
        if (!keep[k]) {
          keep[k] <- TRUE
          i <- e_ord[k, 1]
          j <- e_ord[k, 2]
          deg[i] <- deg[i] + 1L
          deg[j] <- deg[j] + 1L
          kept <- kept + 1L
        }
      }
    }

    e_keep <- e_ord[keep, , drop = FALSE]
    A_sp <- matrix(0L, p, p)
    if (nrow(e_keep) > 0L) {
      A_sp[cbind(e_keep[, 1], e_keep[, 2])] <- 1L
      A_sp[cbind(e_keep[, 2], e_keep[, 1])] <- 1L
    }
    diag(A_sp) <- 0L

    list(
      adj = A_sp,
      stats = list(
        edges_before = n_edges,
        edges_after = nrow(e_keep),
        density_before = n_edges / (p * (p - 1) / 2),
        density_after = nrow(e_keep) / (p * (p - 1) / 2),
        max_degree_before = max(rowSums(A)),
        max_degree_after = max(rowSums(A_sp))
      )
    )
  }
}

if (file.exists(real_glasso_sparse_file)) {
  R_glasso_real_sparse <- readRDS(real_glasso_sparse_file)
  gls_real_stats <- attr(R_glasso_real_sparse, "sparsify_stats")
} else {
  corr_abs_real <- abs(stats::cor(X_real))
  diag(corr_abs_real) <- 0
  gls_real <- sparsify_glasso_adj(
    R_glasso_real_raw,
    score_mat = corr_abs_real,
    keep_ratio = real_glasso_keep_ratio,
    max_degree = real_glasso_max_degree
  )
  R_glasso_real_sparse <- gls_real$adj
  gls_real_stats <- gls_real$stats
  attr(R_glasso_real_sparse, "sparsify_stats") <- gls_real_stats
  saveRDS(R_glasso_real_sparse, real_glasso_sparse_file)
}

if (is.null(gls_real_stats)) {
  gls_real_stats <- list(
    edges_before = sum(R_glasso_real_raw[upper.tri(R_glasso_real_raw)] != 0),
    edges_after = sum(R_glasso_real_sparse[upper.tri(R_glasso_real_sparse)] != 0),
    density_before = sum(R_glasso_real_raw[upper.tri(R_glasso_real_raw)] != 0) / (ncol(R_glasso_real_raw) * (ncol(R_glasso_real_raw) - 1) / 2),
    density_after = sum(R_glasso_real_sparse[upper.tri(R_glasso_real_sparse)] != 0) / (ncol(R_glasso_real_sparse) * (ncol(R_glasso_real_sparse) - 1) / 2),
    max_degree_before = max(rowSums(R_glasso_real_raw != 0)),
    max_degree_after = max(rowSums(R_glasso_real_sparse != 0))
  )
}

cat(sprintf(
  paste0(
    "[realdata-fit] sparsified glasso adjacency (%s): edges %d -> %d, ",
    "density %.4f -> %.4f, max degree %d -> %d\n"
  ),
  real_glasso_tag,
  gls_real_stats$edges_before, gls_real_stats$edges_after,
  gls_real_stats$density_before, gls_real_stats$density_after,
  gls_real_stats$max_degree_before, gls_real_stats$max_degree_after
))

real_f_glasso <- file.path(real_cache_dir, sprintf("real_glasso_%s_%s.rds", real_glasso_tag, real_ctrl_tag))
real_f_fixed <- file.path(real_cache_dir, sprintf("real_fixed_%s.rds", real_ctrl_tag))
real_f_ggm <- file.path(real_cache_dir, sprintf("real_ggm_sparse_%s_%s.rds", real_sggm_tag, real_ctrl_tag))
real_f_dual_glasso <- file.path(real_cache_dir, sprintf("real_glasso_fixed_%s_%s.rds", real_glasso_tag, real_ctrl_tag))
real_f_dual_ggm <- file.path(real_cache_dir, sprintf("real_ggm_fixed_sparse_%s_%s.rds", real_sggm_tag, real_ctrl_tag))

if (file.exists(real_f_glasso)) {
  o <- readRDS(real_f_glasso)
  fit_glasso <- o$fit
  t_real_glasso <- o$time
  cat(sprintf("[realdata-fit] loaded cache: %s\n", basename(real_f_glasso)))
} else {
  cat("[realdata-fit] running: Single glasso (sparsified fixed) ...\n")
  t_real_glasso <- system.time({
    fit_glasso <- bvs_pg(
      X_real, y_real,
      adj_type = "fixed",
      adj_fixed = R_glasso_real_sparse,
      niter = niter_real, burnin = burnin_real, thin = thin_real,
      n_mh_gamma = real_n_mh_gamma, Tmax = real_tmax, proposal_type = real_proposal_type
    )
    fit_glasso$adj_type <- "glasso"
  })
  saveRDS(list(fit = fit_glasso, time = t_real_glasso), real_f_glasso)
}

if (file.exists(real_f_fixed)) {
  o <- readRDS(real_f_fixed)
  fit_fixed <- o$fit
  t_real_fixed <- o$time
  cat(sprintf("[realdata-fit] loaded cache: %s\n", basename(real_f_fixed)))
} else {
  cat("[realdata-fit] running: Single KEGG fixed ...\n")
  t_real_fixed <- system.time({
    fit_fixed <- bvs_pg(
      X_real, y_real,
      adj_type = "fixed",
      adj_fixed = R_kegg,
      niter = niter_real, burnin = burnin_real, thin = thin_real,
      n_mh_gamma = real_n_mh_gamma, Tmax = real_tmax, proposal_type = real_proposal_type
    )
  })
  saveRDS(list(fit = fit_fixed, time = t_real_fixed), real_f_fixed)
}

if (file.exists(real_f_ggm)) {
  o <- readRDS(real_f_ggm)
  fit_ggm <- o$fit
  t_real_ggm <- o$time
  cat(sprintf("[realdata-fit] loaded cache: %s\n", basename(real_f_ggm)))
} else {
  cat("[realdata-fit] running: Single sparse Bayesian GGM ...\n")
  t_real_ggm <- system.time({
    fit_ggm <- bvs_pg(
      X_real, y_real,
      adj_type = "ggm",
      sparse = TRUE,
      S_ggm = S_real_ggm,
      niter = niter_real, burnin = burnin_real, thin = thin_real,
      n_mh_gamma = real_n_mh_gamma, Tmax = real_tmax, proposal_type = real_proposal_type,
      store_beta = TRUE, store_gamma = TRUE, store_Z_list = FALSE, store_Z_pip = TRUE
    )
  })
  saveRDS(list(fit = fit_ggm, time = t_real_ggm), real_f_ggm)
}

if (file.exists(real_f_dual_glasso)) {
  o <- readRDS(real_f_dual_glasso)
  fit_dual_glasso <- o$fit
  t_real_dual_glasso <- o$time
  cat(sprintf("[realdata-fit] loaded cache: %s\n", basename(real_f_dual_glasso)))
} else {
  cat("[realdata-fit] running: Dual KEGG + glasso (sparsified fixed) ...\n")
  t_real_dual_glasso <- system.time({
    fit_dual_glasso <- bvs_pg(
      X_real, y_real,
      adj_type = "dual_fixed",
      adj_fixed = R_glasso_real_sparse,
      adj_fixed2 = R_kegg,
      niter = niter_real, burnin = burnin_real, thin = thin_real,
      n_mh_gamma = real_n_mh_gamma, Tmax = real_tmax, proposal_type = real_proposal_type,
      mu = hpars_real$mu,
      eta1_sd = hpars_real$eta1_sd, eta2_sd = hpars_real$eta2_sd,
      e_eta = hpars_real$e_eta, f_eta = hpars_real$f_eta,
      mu_tilde = hpars_real$mu_tilde,
      eta1_tilde = hpars_real$eta1_tilde, eta2_tilde = hpars_real$eta2_tilde
    )
    fit_dual_glasso$adj_type <- "glasso_fixed"
  })
  saveRDS(list(fit = fit_dual_glasso, time = t_real_dual_glasso), real_f_dual_glasso)
}

if (file.exists(real_f_dual_ggm)) {
  o <- readRDS(real_f_dual_ggm)
  fit_dual_ggm <- o$fit
  t_real_dual_ggm <- o$time
  cat(sprintf("[realdata-fit] loaded cache: %s\n", basename(real_f_dual_ggm)))
} else {
  cat("[realdata-fit] running: Dual KEGG + sparse Bayesian GGM ...\n")
  t_real_dual_ggm <- system.time({
    fit_dual_ggm <- bvs_pg(
      X_real, y_real,
      adj_type = "ggm_fixed",
      adj_fixed = R_kegg,
      sparse = TRUE,
      S_ggm = S_real_ggm,
      niter = niter_real, burnin = burnin_real, thin = thin_real,
      n_mh_gamma = real_n_mh_gamma, Tmax = real_tmax, proposal_type = real_proposal_type,
      mu = hpars_real$mu,
      eta1_sd = hpars_real$eta1_sd, eta2_sd = hpars_real$eta2_sd,
      e_eta = hpars_real$e_eta, f_eta = hpars_real$f_eta,
      mu_tilde = hpars_real$mu_tilde,
      eta1_tilde = hpars_real$eta1_tilde, eta2_tilde = hpars_real$eta2_tilde,
      store_beta = TRUE, store_gamma = TRUE, store_Z_list = FALSE, store_Z_pip = TRUE
    )
  })
  saveRDS(list(fit = fit_dual_ggm, time = t_real_dual_ggm), real_f_dual_ggm)
}

fits_real <- list(
  "Single: glasso" = fit_glasso,
  "Single: KEGG fixed" = fit_fixed,
  "Single: Bayesian GGM" = fit_ggm,
  "Dual: KEGG + glasso" = fit_dual_glasso,
  "Dual: KEGG + Bayesian GGM" = fit_dual_ggm
)

times_real <- c(
  "Single: glasso" = as.numeric(t_real_glasso["elapsed"]),
  "Single: KEGG fixed" = as.numeric(t_real_fixed["elapsed"]),
  "Single: Bayesian GGM" = as.numeric(t_real_ggm["elapsed"]),
  "Dual: KEGG + glasso" = as.numeric(t_real_dual_glasso["elapsed"]),
  "Dual: KEGG + Bayesian GGM" = as.numeric(t_real_dual_ggm["elapsed"])
)
```

## Final selected variables by method

```{r realdata-selected, eval=TRUE}
pip_threshold_real <- 0.5

format_gene_list <- function(gene_vec, max_show = 40L) {
  if (!length(gene_vec)) {
    return("None")
  }
  if (length(gene_vec) <= max_show) {
    return(paste(gene_vec, collapse = ", "))
  }
  paste0(
    paste(gene_vec[seq_len(max_show)], collapse = ", "),
    sprintf(", ... (+%d more)", length(gene_vec) - max_show)
  )
}

selected_table <- do.call(rbind, lapply(names(fits_real), function(mn) {
  fit <- fits_real[[mn]]
  pip <- compute_pip(fit)
  sel_idx <- which(pip >= pip_threshold_real)
  ord <- order(pip, decreasing = TRUE)
  top20 <- colnames(X_real)[ord[seq_len(min(20L, length(ord)))]]

  data.frame(
    Method = mn,
    N_selected_PIP_ge_0_5 = length(sel_idx),
    Selected_genes_PIP_ge_0_5 = format_gene_list(colnames(X_real)[sel_idx], max_show = 40L),
    Top20_genes_by_PIP = paste(top20, collapse = ", "),
    stringsAsFactors = FALSE
  )
}))

knitr::kable(
  selected_table,
  caption = "Final selected variables for each real-data model (selection threshold PIP >= 0.5)."
)
```

## MCMC convergence diagnostics and sampling performance

```{r realdata-diagnostics, eval=TRUE}
safe_ess <- function(x, lag.max = 200L) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  n <- length(x)
  if (n < 3L) {
    return(NA_real_)
  }
  v <- stats::var(x)
  if (!is.finite(v) || v <= .Machine$double.eps) {
    return(n)
  }

  ac <- as.numeric(stats::acf(x, lag.max = min(lag.max, n - 1L), plot = FALSE)$acf)[-1L]
  if (!length(ac)) {
    return(n)
  }
  first_nonpos <- which(ac <= 0)
  m <- if (length(first_nonpos)) first_nonpos[1L] - 1L else length(ac)
  if (m < 1L) {
    return(n)
  }
  tau <- 1 + 2 * sum(ac[seq_len(m)])
  if (!is.finite(tau) || tau <= 0) {
    return(n)
  }
  min(n, n / tau)
}

lag1_acf <- function(x) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  if (length(x) < 3L) {
    return(NA_real_)
  }
  v <- stats::var(x)
  if (!is.finite(v) || v <= .Machine$double.eps) {
    return(NA_real_)
  }
  as.numeric(stats::acf(x, lag.max = 1L, plot = FALSE)$acf[2L])
}

# Representative gene for diagnostics: highest average PIP across methods
pip_mat_real <- sapply(fits_real, compute_pip)
pip_mean_real <- rowMeans(pip_mat_real)
top_gene_idx_real <- order(pip_mean_real, decreasing = TRUE)[1L]
top_gene_name_real <- colnames(X_real)[top_gene_idx_real]

diag_df_real <- do.call(rbind, lapply(names(fits_real), function(mn) {
  fit <- fits_real[[mn]]
  pip <- compute_pip(fit)
  beta_chain <- get_beta_matrix(fit)[, top_gene_idx_real]
  ms <- model_size_chain(fit)
  ess_beta <- safe_ess(beta_chain)
  n_saved <- length(ms)
  iter_sec <- n_saved / max(as.numeric(times_real[mn]), 1e-8)

  data.frame(
    Method = mn,
    Runtime_sec = round(times_real[mn], 2),
    N_saved = n_saved,
    Iter_per_sec = round(iter_sec, 2),
    Mean_model_size = round(mean(ms), 2),
    SD_model_size = round(stats::sd(ms), 2),
    Median_model_size = round(stats::median(ms), 2),
    Q05_model_size = round(as.numeric(stats::quantile(ms, 0.05)), 2),
    Q95_model_size = round(as.numeric(stats::quantile(ms, 0.95)), 2),
    Top_gene_for_diag = top_gene_name_real,
    PIP_top_gene = round(pip[top_gene_idx_real], 3),
    Lag1_ACF_top_beta = round(lag1_acf(beta_chain), 3),
    ESS_top_beta = round(ess_beta, 1),
    ESS_per_sec = round(ess_beta / max(as.numeric(times_real[mn]), 1e-8), 3),
    MCSE_top_beta = round(stats::sd(beta_chain) / sqrt(max(1, ess_beta)), 4),
    stringsAsFactors = FALSE
  )
}))

knitr::kable(
  diag_df_real,
  caption = "MCMC diagnostic summary and sampling performance for the five real-data models."
)
```

### Gelman-Rubin diagnostics (2 chains, 2 seeds)

```{r realdata-gelman, eval=TRUE}
vars_gr_real <- seq_len(min(5L, ncol(X_real)))
seed_real <- c(20261301L, 20261302L)

# Representative real-data model for two-chain Gelman-Rubin check:
# PG with single glasso-estimated adjacency.
set.seed(seed_real[1])
fit_real_gr1 <- bvs_pg(
  X_real, y_real,
  adj_type = "glasso",
  glasso_criterion = "ebic",
  niter = niter_real, burnin = burnin_real, thin = thin_real
)
set.seed(seed_real[2])
fit_real_gr2 <- bvs_pg(
  X_real, y_real,
  adj_type = "glasso",
  glasso_criterion = "ebic",
  niter = niter_real, burnin = burnin_real, thin = thin_real
)

gelman_real <- safe_gelman_df(
  list(fit_real_gr1, fit_real_gr2),
  vars = vars_gr_real,
  autoburnin = FALSE
)
gelman_real_summary <- gelman_method_summary(gelman_real, "Single: glasso (PG)")

knitr::kable(
  gelman_real_summary,
  caption = "Real-data example 1 Gelman-Rubin summary using two chains (two seeds)."
)
knitr::kable(
  data.frame(
    Method = "Single: glasso (PG)",
    Parameter = gelman_real$Parameter,
    Point_Rhat = round(gelman_real$Point_est, 3),
    Upper_CI_Rhat = round(gelman_real$Upper_CI, 3),
    stringsAsFactors = FALSE
  ),
  caption = "Real-data example 1 PSRF details (Alpha and first five coefficients)."
)
```

```{r realdata-gelman-trace-2chain, eval=TRUE, fig.height=3.6, fig.width=12, fig.cap="Real-data example 1: two-chain trace plots for representative regression coefficients from Gelman-Rubin diagnostic runs."}
beta_mat_real_gr1 <- get_beta_matrix(fit_real_gr1)
if (ncol(beta_mat_real_gr1) > 0) {
  beta_idx_real_gr <- head(order(colMeans(abs(beta_mat_real_gr1)), decreasing = TRUE), 4L)
} else {
  beta_idx_real_gr <- seq_len(min(4L, ncol(X_real)))
}
beta_idx_real_gr <- as.integer(beta_idx_real_gr)
par(
  mfrow = c(1, length(beta_idx_real_gr)),
  mar = c(3, 3, 2, 1),
  mgp = c(1.8, 0.5, 0)
)
plot_two_chain_trace_panels(fit_real_gr1, fit_real_gr2, beta_idx_real_gr, "PG (glasso)")
```

```{r realdata-modelsize-trace, eval=TRUE, fig.height=7, fig.width=10, fig.cap="Trace plots of model size across MCMC iterations for the five real-data models."}
n_panels_real <- length(fits_real)
nrow_real <- 2
ncol_real <- ceiling(n_panels_real / nrow_real)
par(mfrow = c(nrow_real, ncol_real), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (mn in names(fits_real)) {
  plot(model_size_chain(fits_real[[mn]]),
    type = "l", col = "darkblue",
    main = mn, xlab = "Iteration", ylab = "Model size", cex.main = 0.9
  )
}
```

```{r realdata-topgene-trace, eval=TRUE, fig.height=7, fig.width=10, fig.cap="Trace plots of the representative top-PIP gene coefficient across all five methods."}
par(mfrow = c(nrow_real, ncol_real), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (mn in names(fits_real)) {
  beta_chain <- get_beta_matrix(fits_real[[mn]])[, top_gene_idx_real]
  plot(beta_chain,
    type = "l", col = "steelblue",
    main = paste0(mn, ": ", top_gene_name_real),
    xlab = "Iteration", ylab = "beta", cex.main = 0.9
  )
}
```

```{r realdata-topgene-runmean, eval=TRUE, fig.height=7, fig.width=10, fig.cap="Running-mean plots of the representative top-PIP gene coefficient across all five methods."}
par(mfrow = c(nrow_real, ncol_real), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (mn in names(fits_real)) {
  beta_chain <- get_beta_matrix(fits_real[[mn]])[, top_gene_idx_real]
  rm_vals <- running_mean(beta_chain)
  plot(rm_vals,
    type = "l", col = "darkgreen",
    main = paste0(mn, ": running mean(", top_gene_name_real, ")"),
    xlab = "Iteration", ylab = "Running mean(beta)", cex.main = 0.85
  )
}
```

```{r realdata-topgene-acf, eval=TRUE, fig.height=7, fig.width=10, fig.cap="ACF plots of the representative top-PIP gene coefficient across all five methods."}
par(mfrow = c(nrow_real, ncol_real), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (mn in names(fits_real)) {
  beta_chain <- get_beta_matrix(fits_real[[mn]])[, top_gene_idx_real]
  safe_acf_plot(beta_chain,
    lag.max = 50,
    main = paste0(mn, ": ACF(", top_gene_name_real, ")"),
    col = "steelblue"
  )
}
```

```{r realdata-runtime-barplot, eval=TRUE, fig.height=4, fig.width=9, fig.cap="Runtime comparison (seconds) across the five real-data models."}
ord_rt <- order(times_real, decreasing = TRUE)
cols_real <- c("steelblue", "coral", "seagreen3", "goldenrod3", "mediumpurple3")
barplot(times_real[ord_rt],
  col = cols_real[ord_rt],
  border = NA, las = 2, ylab = "Elapsed time (sec)",
  main = "Real-data runtime comparison"
)
```

# Real-data example (p >> n): Leukemia microarray with sparse high-dimensional BVS {#sec-real-hd}

As a second real-data analysis in a strict high-dimensional regime, we consider
the Acute Lymphoblastic Leukemia (ALL) microarray benchmark [@Chiaretti2004], where the number of
gene-expression predictors far exceeds the sample size.

### Dataset profile (ALL microarray)

- **Study year**: 2004 [@Chiaretti2004].
- **Original benchmark scale**: 128 tissue samples and 12,625 processed genes in the canonical benchmark representation.
- **Data-generation context**: high-density oligonucleotide microarray expression profiling
  on leukemia tumor tissues. Here we focus on classifying patients with the BCR/ABL fusion versus those negative for it.
- **Data object used here**: `ALL::ALL` ExpressionSet from Bioconductor.
  Exact dimensions used in your installation are printed in the setup/inspection
  chunk below.

In this example:

- **Outcome variable (`y`)**: binary tumor classification (BCR/ABL vs NEG).
- **Predictor variables (`X`)**: high-dimensional gene-expression measurements.
- **Bayesian variable-selection target**: identify genes associated with tumor
  status under logistic regression with network-informed Ising priors.

We compare five adjacency configurations under sparse high-dimensional
MCMC settings:

1. Single-source `glasso`.
2. Single-source `ggm` with sparse backend.
3. Single-source `fixed` adjacency using KEGG only.
4. Dual-source `glasso_fixed` (KEGG + glasso).
5. Dual-source `ggm_fixed` (KEGG + sparse Bayesian GGM).

### Variable description (ALL objects)

- `ALL`: Bioconductor `ExpressionSet` (12625 genes × 128 samples); `exprs(ALL)` gives the gene expression matrix and `ALL$mol.biol` gives tissue subtypes.
- `X_cc`: cleaned numeric matrix (samples × genes, i.e., transposed from the ExpressionSet) used as raw predictor matrix in this vignette.
- `probe_ids`: original probe identifiers from `rownames(ALL)`.
- `gene_symbols_hd`: mapped gene symbols via `hgu95av2.db` (fallback to probe IDs when the annotation package is unavailable).
- `y_hd`: binary outcome for logistic BVS (`1 = BCR/ABL`, `0 = NEG`).
- `X_hd`: standardized, variance-filtered high-dimensional predictor matrix.
- `gene_labels_hd`: final predictor labels used in the BVS analysis.

## Data setup for high-dimensional real example

```{r realhd-setup, eval=TRUE}
# Required packages for this section:
# ALL (Bioconductor), Biobase (Bioconductor), AnnotationDbi, org.Hs.eg.db, KEGGREST, hgu95av2.db
req_pkgs_hd <- c("ALL", "Biobase", "hgu95av2.db")
missing_hd <- req_pkgs_hd[!vapply(req_pkgs_hd, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_hd)) {
  stop(
    "Install required Bioconductor packages first via BiocManager::install(c(",
    paste0('"', missing_hd, '"', collapse = ", "), "))"
  )
}
kegg_pkgs_hd <- c("AnnotationDbi", "org.Hs.eg.db", "KEGGREST")
kegg_pkg_ok_hd <- all(vapply(kegg_pkgs_hd, requireNamespace, logical(1), quietly = TRUE))
if (!kegg_pkg_ok_hd) {
  message(
    "KEGG annotation packages not fully available (",
    paste(kegg_pkgs_hd[!vapply(kegg_pkgs_hd, requireNamespace, logical(1), quietly = TRUE)], collapse = ", "),
    "). Using correlation-based external adjacency fallback for this test run."
  )
}

if (!requireNamespace("hgu95av2.db", quietly = TRUE) ||
  !requireNamespace("AnnotationDbi", quietly = TRUE)) {
  message("Package 'hgu95av2.db' not installed; using probe IDs as feature labels.")
}

# Chiaretti et al. (2004) acute lymphoblastic leukemia (ALL) data
data("ALL", package = "ALL")

# Subset patients to BCR/ABL vs NEG
bcr_neg <- ALL$mol.biol %in% c("BCR/ABL", "NEG")
ALL_sub <- ALL[, bcr_neg]

# Extract expression matrix (genes x samples -> transpose to samples x genes)
X_cc <- t(Biobase::exprs(ALL_sub)) # n x p: samples x genes

# Binary outcome: tumor = 1, normal = 0
y_hd <- as.integer(ALL_sub$mol.biol == "BCR/ABL")

# Probe/gene labels (row names of the ExpressionSet = probe IDs)
probe_ids <- rownames(ALL_sub)
if (is.null(probe_ids)) probe_ids <- paste0("probe_", seq_len(ncol(X_cc)))

if (requireNamespace("hgu95av2.db", quietly = TRUE) &&
  requireNamespace("AnnotationDbi", quietly = TRUE)) {
  gene_symbols_hd <- tryCatch(
    AnnotationDbi::mapIds(
      hgu95av2.db::hgu95av2.db,
      keys = probe_ids,
      keytype = "PROBEID",
      column = "SYMBOL",
      multiVals = "first"
    ),
    error = function(e) {
      warning("Warning: PROBEID mapping failed (hgu95av2.db). Using basic probe tags instead.")
      rep(NA_character_, length(probe_ids))
    }
  )
  gene_symbols_hd <- as.character(gene_symbols_hd)
  gene_symbols_hd[is.na(gene_symbols_hd) | !nzchar(gene_symbols_hd)] <-
    probe_ids[is.na(gene_symbols_hd) | !nzchar(gene_symbols_hd)]
} else {
  gene_symbols_hd <- probe_ids
}

colnames(X_cc) <- make.unique(gene_symbols_hd)

# Keep top variable genes for stable run time while maintaining p >> n
p_keep_hd <- if (isTRUE(quick_check)) min(300L, ncol(X_cc)) else min(1000L, ncol(X_cc))
var_hd <- apply(X_cc, 2L, var)
keep_hd <- order(var_hd, decreasing = TRUE)[seq_len(p_keep_hd)]
X_hd <- scale(X_cc[, keep_hd, drop = FALSE])
gene_labels_hd <- colnames(X_hd)

cat(sprintf("Raw colonCA dimensions: n=%d, p=%d\n", nrow(X_cc), ncol(X_cc)))
cat(sprintf("Analysis subset dimensions after filtering: n=%d, p=%d\n", nrow(X_hd), ncol(X_hd)))
cat(sprintf(
  "Outcome: tumor (y=1) n=%d, normal (y=0) n=%d\n",
  sum(y_hd == 1L), sum(y_hd == 0L)
))
```

### Quick printout and summary (ColonCancer)

```{r realhd-inspect, eval=TRUE}
cat("=== colonCA (Alon 1999): first 10 observations and summaries ===\n")

cat("\nFirst 10 rows of subject-level data (outcome + first 8 predictors from X_cc):\n")
print(head(
  data.frame(y = y_hd, X_cc[, seq_len(min(8L, ncol(X_cc))), drop = FALSE]),
  10
))

cat("\nFirst 10 probe/gene labels used in analysis:\n")
print(head(data.frame(
  probe_id = probe_ids[seq_len(min(10L, length(probe_ids)))],
  gene_label = colnames(X_cc)[seq_len(min(10L, ncol(X_cc)))]
), 10))

cat("\nSummary of outcome variable (y_hd):\n")
print(summary(as.factor(y_hd)))

cat("\nSummary of full raw predictor matrix (overall values):\n")
print(summary(as.vector(X_cc)))

cat("\nSummary of first 10 predictors in X_cc:\n")
print(summary(as.data.frame(X_cc[, seq_len(min(10L, ncol(X_cc))), drop = FALSE])))
```

## KEGG external adjacency for high-dimensional example

```{r realhd-kegg-adj, eval=TRUE}
fallback_hd_adj <- function() {
  C_hd <- abs(stats::cor(X_hd, use = "pairwise.complete.obs"))
  diag(C_hd) <- 0
  thr_hd <- as.numeric(stats::quantile(C_hd[upper.tri(C_hd)], probs = 0.99, na.rm = TRUE))
  Rk <- matrix(as.integer(C_hd >= thr_hd), nrow(C_hd), ncol(C_hd))
  diag(Rk) <- 0L
  cat(sprintf(
    "Fallback external adjacency (|cor| >= %.3f) edges among retained high-dimensional genes: %d\n",
    thr_hd, sum(Rk) / 2
  ))
  Rk
}

if (isTRUE(kegg_pkg_ok_hd)) {
  build_kegg_adjacency_hd <- function(symbols) {
    p <- length(symbols)

    entrez <- tryCatch(
      AnnotationDbi::mapIds(
        org.Hs.eg.db::org.Hs.eg.db,
        keys = symbols,
        keytype = "SYMBOL",
        column = "ENTREZID",
        multiVals = "first"
      ),
      error = function(e) {
        return(rep(NA_character_, p))
      }
    )

    map_df <- data.frame(
      idx = seq_len(p),
      entrez = as.character(entrez),
      stringsAsFactors = FALSE
    )
    map_df <- map_df[!is.na(map_df$entrez) & nzchar(map_df$entrez), , drop = FALSE]

    if (nrow(map_df) == 0) {
      stop("No valid Entrez/SYMBOL mappings found for the retained features.")
    }

    kmap <- KEGGREST::keggLink("pathway", "hsa")
    if (!length(kmap)) stop("KEGG mapping retrieval failed.")

    if (startsWith(names(kmap)[1], "hsa:")) {
      gp <- data.frame(
        entrez = sub("^hsa:", "", names(kmap)),
        pathway = sub("^path:", "", unname(kmap)),
        stringsAsFactors = FALSE
      )
    } else {
      gp <- data.frame(
        entrez = sub("^hsa:", "", unname(kmap)),
        pathway = sub("^path:", "", names(kmap)),
        stringsAsFactors = FALSE
      )
    }

    hit <- merge(map_df, gp, by = "entrez")
    groups <- split(hit$idx, hit$pathway)

    R_kegg_hd <- matrix(0L, p, p)
    for (g in groups) {
      g <- sort(unique(g))
      if (length(g) < 2L) next
      cmb <- utils::combn(g, 2L)
      R_kegg_hd[cbind(cmb[1L, ], cmb[2L, ])] <- 1L
      R_kegg_hd[cbind(cmb[2L, ], cmb[1L, ])] <- 1L
    }
    diag(R_kegg_hd) <- 0L
    R_kegg_hd
  }

  R_kegg_hd <- tryCatch(
    build_kegg_adjacency_hd(gene_labels_hd),
    error = function(e) {
      message("KEGG adjacency construction failed (high-dimensional): ", conditionMessage(e))
      NULL
    }
  )
  if (is.null(R_kegg_hd)) {
    R_kegg_hd <- fallback_hd_adj()
  } else {
    cat(sprintf("KEGG edges among retained high-dimensional genes: %d\n", sum(R_kegg_hd) / 2))
  }
} else {
  R_kegg_hd <- fallback_hd_adj()
}
```

## Fit five high-dimensional BVS models

```{r realhd-fit, eval=TRUE}
cfg_realhd <- mcmc_scale(
  full_niter = 500L, full_burnin = 200L, full_thin = 2L,
  quick_niter = 5L, quick_burnin = 2L, quick_thin = 1L
)
niter_hd <- cfg_realhd$niter
burnin_hd <- cfg_realhd$burnin
thin_hd <- cfg_realhd$thin
hd_sparse <- TRUE

realhd_cache_dir <- "realhd_saved_runs"
if (!dir.exists(realhd_cache_dir)) dir.create(realhd_cache_dir, recursive = TRUE)
realhd_glasso_keep_ratio <- if (isTRUE(quick_check)) 0.20 else 0.05
realhd_glasso_max_degree <- if (isTRUE(quick_check)) 120L else 40L
realhd_glasso_tag <- sprintf(
  "spr%02d_deg%d",
  as.integer(round(100 * realhd_glasso_keep_ratio)),
  realhd_glasso_max_degree
)
realhd_glasso_raw_file <- file.path(
  realhd_cache_dir,
  sprintf("realhd_glasso_adj_ebic_n%d_p%d.rds", nrow(X_hd), ncol(X_hd))
)
realhd_glasso_sparse_file <- file.path(
  realhd_cache_dir,
  sprintf(
    "realhd_glasso_adj_ebic_%s_n%d_p%d.rds",
    realhd_glasso_tag, nrow(X_hd), ncol(X_hd)
  )
)

if (file.exists(realhd_glasso_raw_file)) {
  R_glasso_hd_raw <- readRDS(realhd_glasso_raw_file)
} else {
  R_glasso_hd_raw <- estimate_glasso_adj(X_hd, criterion = "ebic")
  saveRDS(R_glasso_hd_raw, realhd_glasso_raw_file)
}

if (!exists("sparsify_glasso_adj", mode = "function")) {
  sparsify_glasso_adj <- function(adj, score_mat, keep_ratio = 0.05, max_degree = 40L) {
    A <- (adj != 0) * 1L
    A <- ((A + t(A)) > 0) * 1L
    diag(A) <- 0L
    p <- nrow(A)

    e_idx <- which(upper.tri(A) & A == 1L, arr.ind = TRUE)
    n_edges <- nrow(e_idx)
    if (n_edges == 0L) {
      return(list(
        adj = A,
        stats = list(
          edges_before = 0L, edges_after = 0L,
          density_before = 0, density_after = 0,
          max_degree_before = 0L, max_degree_after = 0L
        )
      ))
    }

    keep_ratio <- min(max(as.numeric(keep_ratio), 0.001), 1)
    max_degree <- max(1L, as.integer(max_degree))
    target_edges <- max(1L, as.integer(ceiling(keep_ratio * n_edges)))

    sc <- score_mat[cbind(e_idx[, 1], e_idx[, 2])]
    sc[!is.finite(sc)] <- 0
    ord <- order(sc, decreasing = TRUE)
    e_ord <- e_idx[ord, , drop = FALSE]

    deg <- integer(p)
    keep <- logical(nrow(e_ord))
    kept <- 0L

    for (k in seq_len(nrow(e_ord))) {
      if (kept >= target_edges) break
      i <- e_ord[k, 1]
      j <- e_ord[k, 2]
      if (deg[i] < max_degree && deg[j] < max_degree) {
        keep[k] <- TRUE
        deg[i] <- deg[i] + 1L
        deg[j] <- deg[j] + 1L
        kept <- kept + 1L
      }
    }

    if (kept < target_edges) {
      for (k in seq_len(nrow(e_ord))) {
        if (kept >= target_edges) break
        if (!keep[k]) {
          keep[k] <- TRUE
          i <- e_ord[k, 1]
          j <- e_ord[k, 2]
          deg[i] <- deg[i] + 1L
          deg[j] <- deg[j] + 1L
          kept <- kept + 1L
        }
      }
    }

    e_keep <- e_ord[keep, , drop = FALSE]
    A_sp <- matrix(0L, p, p)
    if (nrow(e_keep) > 0L) {
      A_sp[cbind(e_keep[, 1], e_keep[, 2])] <- 1L
      A_sp[cbind(e_keep[, 2], e_keep[, 1])] <- 1L
    }
    diag(A_sp) <- 0L

    list(
      adj = A_sp,
      stats = list(
        edges_before = n_edges,
        edges_after = nrow(e_keep),
        density_before = n_edges / (p * (p - 1) / 2),
        density_after = nrow(e_keep) / (p * (p - 1) / 2),
        max_degree_before = max(rowSums(A)),
        max_degree_after = max(rowSums(A_sp))
      )
    )
  }
}

if (file.exists(realhd_glasso_sparse_file)) {
  R_glasso_hd_sparse <- readRDS(realhd_glasso_sparse_file)
  gls_hd_stats <- attr(R_glasso_hd_sparse, "sparsify_stats")
} else {
  corr_abs_hd <- abs(stats::cor(X_hd))
  diag(corr_abs_hd) <- 0
  gls_hd <- sparsify_glasso_adj(
    R_glasso_hd_raw,
    score_mat = corr_abs_hd,
    keep_ratio = realhd_glasso_keep_ratio,
    max_degree = realhd_glasso_max_degree
  )
  R_glasso_hd_sparse <- gls_hd$adj
  gls_hd_stats <- gls_hd$stats
  attr(R_glasso_hd_sparse, "sparsify_stats") <- gls_hd_stats
  saveRDS(R_glasso_hd_sparse, realhd_glasso_sparse_file)
}

if (is.null(gls_hd_stats)) {
  gls_hd_stats <- list(
    edges_before = sum(R_glasso_hd_raw[upper.tri(R_glasso_hd_raw)] != 0),
    edges_after = sum(R_glasso_hd_sparse[upper.tri(R_glasso_hd_sparse)] != 0),
    density_before = sum(R_glasso_hd_raw[upper.tri(R_glasso_hd_raw)] != 0) / (ncol(R_glasso_hd_raw) * (ncol(R_glasso_hd_raw) - 1) / 2),
    density_after = sum(R_glasso_hd_sparse[upper.tri(R_glasso_hd_sparse)] != 0) / (ncol(R_glasso_hd_sparse) * (ncol(R_glasso_hd_sparse) - 1) / 2),
    max_degree_before = max(rowSums(R_glasso_hd_raw != 0)),
    max_degree_after = max(rowSums(R_glasso_hd_sparse != 0))
  )
}

cat(sprintf(
  paste0(
    "[realhd-fit] sparsified glasso adjacency (%s): edges %d -> %d, ",
    "density %.4f -> %.4f, max degree %d -> %d\n"
  ),
  realhd_glasso_tag,
  gls_hd_stats$edges_before, gls_hd_stats$edges_after,
  gls_hd_stats$density_before, gls_hd_stats$density_after,
  gls_hd_stats$max_degree_before, gls_hd_stats$max_degree_after
))

t_hd_glasso <- system.time({
  fit_hd_glasso <- bvs_pg(
    X_hd, y_hd,
    adj_type = "fixed",
    adj_fixed = R_glasso_hd_sparse,
    niter = niter_hd, burnin = burnin_hd, thin = thin_hd
  )
  fit_hd_glasso$adj_type <- "glasso"
})

t_hd_fixed <- system.time({
  fit_hd_fixed <- bvs_pg(
    X_hd, y_hd,
    adj_type = "fixed",
    adj_fixed = R_kegg_hd,
    niter = niter_hd, burnin = burnin_hd, thin = thin_hd
  )
})

t_hd_ggm <- system.time({
  fit_hd_ggm <- bvs_pg(
    X_hd, y_hd,
    adj_type = "ggm",
    sparse = hd_sparse,
    niter = niter_hd, burnin = burnin_hd, thin = thin_hd,
    store_beta = TRUE, store_gamma = TRUE, store_Z_list = FALSE, store_Z_pip = TRUE
  )
})

t_hd_dual_glasso <- system.time({
  fit_hd_dual_glasso <- bvs_pg(
    X_hd, y_hd,
    adj_type = "dual_fixed",
    adj_fixed = R_glasso_hd_sparse,
    adj_fixed2 = R_kegg_hd,
    niter = niter_hd, burnin = burnin_hd, thin = thin_hd,
    mu = hpars_realhd$mu,
    eta1_sd = hpars_realhd$eta1_sd, eta2_sd = hpars_realhd$eta2_sd,
    e_eta = hpars_realhd$e_eta, f_eta = hpars_realhd$f_eta,
    mu_tilde = hpars_realhd$mu_tilde,
    eta1_tilde = hpars_realhd$eta1_tilde, eta2_tilde = hpars_realhd$eta2_tilde
  )
  fit_hd_dual_glasso$adj_type <- "glasso_fixed"
})

t_hd_dual_ggm <- system.time({
  fit_hd_dual_ggm <- bvs_pg(
    X_hd, y_hd,
    adj_type = "ggm_fixed",
    adj_fixed = R_kegg_hd,
    sparse = hd_sparse,
    niter = niter_hd, burnin = burnin_hd, thin = thin_hd,
    mu = hpars_realhd$mu,
    eta1_sd = hpars_realhd$eta1_sd, eta2_sd = hpars_realhd$eta2_sd,
    e_eta = hpars_realhd$e_eta, f_eta = hpars_realhd$f_eta,
    mu_tilde = hpars_realhd$mu_tilde,
    eta1_tilde = hpars_realhd$eta1_tilde, eta2_tilde = hpars_realhd$eta2_tilde,
    store_beta = TRUE, store_gamma = TRUE, store_Z_list = FALSE, store_Z_pip = TRUE
  )
})

fits_hd <- list(
  "Single: glasso" = fit_hd_glasso,
  "Single: KEGG fixed" = fit_hd_fixed,
  "Single: sparse Bayesian GGM" = fit_hd_ggm,
  "Dual: KEGG + glasso" = fit_hd_dual_glasso,
  "Dual: KEGG + sparse Bayesian GGM" = fit_hd_dual_ggm
)

times_hd <- c(
  "Single: glasso" = as.numeric(t_hd_glasso["elapsed"]),
  "Single: KEGG fixed" = as.numeric(t_hd_fixed["elapsed"]),
  "Single: sparse Bayesian GGM" = as.numeric(t_hd_ggm["elapsed"]),
  "Dual: KEGG + glasso" = as.numeric(t_hd_dual_glasso["elapsed"]),
  "Dual: KEGG + sparse Bayesian GGM" = as.numeric(t_hd_dual_ggm["elapsed"])
)
```

## Final selected variables for the high-dimensional real-data models

```{r realhd-selected, eval=TRUE}
pip_threshold_hd <- 0.5

format_gene_list_hd <- function(gene_vec, max_show = 40L) {
  if (!length(gene_vec)) {
    return("None")
  }
  if (length(gene_vec) <= max_show) {
    return(paste(gene_vec, collapse = ", "))
  }
  paste0(
    paste(gene_vec[seq_len(max_show)], collapse = ", "),
    sprintf(", ... (+%d more)", length(gene_vec) - max_show)
  )
}

selected_hd <- do.call(rbind, lapply(names(fits_hd), function(mn) {
  fit <- fits_hd[[mn]]
  pip <- compute_pip(fit)
  sel_idx <- which(pip >= pip_threshold_hd)
  ord <- order(pip, decreasing = TRUE)
  top20 <- gene_labels_hd[ord[seq_len(min(20L, length(ord)))]]

  data.frame(
    Method = mn,
    N_selected_PIP_ge_0_5 = length(sel_idx),
    Selected_genes_PIP_ge_0_5 = format_gene_list_hd(gene_labels_hd[sel_idx], max_show = 40L),
    Top20_genes_by_PIP = paste(top20, collapse = ", "),
    stringsAsFactors = FALSE
  )
}))

knitr::kable(
  selected_hd,
  caption = "Final selected genes in the high-dimensional real-data example (PIP threshold >= 0.5)."
)
```

## MCMC diagnostics and sampling performance (high-dimensional real data)

```{r realhd-diagnostics, eval=TRUE}
safe_ess_hd <- function(x, lag.max = 200L) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  n <- length(x)
  if (n < 3L) {
    return(NA_real_)
  }
  v <- stats::var(x)
  if (!is.finite(v) || v <= .Machine$double.eps) {
    return(n)
  }

  ac <- as.numeric(stats::acf(x, lag.max = min(lag.max, n - 1L), plot = FALSE)$acf)[-1L]
  if (!length(ac)) {
    return(n)
  }
  first_nonpos <- which(ac <= 0)
  m <- if (length(first_nonpos)) first_nonpos[1L] - 1L else length(ac)
  if (m < 1L) {
    return(n)
  }
  tau <- 1 + 2 * sum(ac[seq_len(m)])
  if (!is.finite(tau) || tau <= 0) {
    return(n)
  }
  min(n, n / tau)
}

lag1_acf_hd <- function(x) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  if (length(x) < 3L) {
    return(NA_real_)
  }
  v <- stats::var(x)
  if (!is.finite(v) || v <= .Machine$double.eps) {
    return(NA_real_)
  }
  as.numeric(stats::acf(x, lag.max = 1L, plot = FALSE)$acf[2L])
}

# Representative marker for diagnostics: highest mean PIP across all methods
pip_mat_hd <- sapply(fits_hd, compute_pip)
pip_mean_hd <- rowMeans(pip_mat_hd)
top_gene_idx_hd <- order(pip_mean_hd, decreasing = TRUE)[1L]
top_gene_name_hd <- gene_labels_hd[top_gene_idx_hd]

diag_hd <- do.call(rbind, lapply(names(fits_hd), function(mn) {
  fit <- fits_hd[[mn]]
  pip <- compute_pip(fit)
  beta_chain <- get_beta_matrix(fit)[, top_gene_idx_hd]
  ms <- model_size_chain(fit)
  ess_beta <- safe_ess_hd(beta_chain)
  n_saved <- length(ms)
  iter_sec <- n_saved / max(as.numeric(times_hd[mn]), 1e-8)

  data.frame(
    Method = mn,
    Runtime_sec = round(times_hd[mn], 2),
    N_saved = n_saved,
    Iter_per_sec = round(iter_sec, 2),
    Mean_model_size = round(mean(ms), 2),
    SD_model_size = round(stats::sd(ms), 2),
    Median_model_size = round(stats::median(ms), 2),
    Q05_model_size = round(as.numeric(stats::quantile(ms, 0.05)), 2),
    Q95_model_size = round(as.numeric(stats::quantile(ms, 0.95)), 2),
    Top_gene_for_diag = top_gene_name_hd,
    PIP_top_gene = round(pip[top_gene_idx_hd], 3),
    Lag1_ACF_top_beta = round(lag1_acf_hd(beta_chain), 3),
    ESS_top_beta = round(ess_beta, 1),
    ESS_per_sec = round(ess_beta / max(as.numeric(times_hd[mn]), 1e-8), 3),
    MCSE_top_beta = round(stats::sd(beta_chain) / sqrt(max(1, ess_beta)), 4),
    stringsAsFactors = FALSE
  )
}))

knitr::kable(
  diag_hd,
  caption = "MCMC diagnostics and sampling performance for the five high-dimensional real-data models."
)
```

### Gelman-Rubin diagnostics (2 chains, 2 seeds)

```{r realhd-gelman, eval=TRUE}
vars_gr_hd <- seq_len(min(5L, ncol(X_hd)))
seed_hd <- c(20261401L, 20261402L)

# Representative high-dimensional real-data model for two-chain Gelman-Rubin check:
# PG with sparse Bayesian GGM adjacency.
S_hd_gr <- Matrix::crossprod(X_hd)
if (isTRUE(quick_check)) {
  S_hd_dense <- as.matrix(S_hd_gr)
  d_hd <- diag(S_hd_dense)
  thr_hd <- as.numeric(stats::quantile(abs(S_hd_dense[upper.tri(S_hd_dense)]), probs = 0.90, na.rm = TRUE))
  S_hd_dense[abs(S_hd_dense) < thr_hd] <- 0
  diag(S_hd_dense) <- d_hd
  S_hd_gr <- Matrix::Matrix(S_hd_dense, sparse = TRUE)
}

set.seed(seed_hd[1])
fit_hd_gr1 <- bvs_pg(
  X_hd, y_hd,
  adj_type = "ggm",
  sparse = TRUE,
  S_ggm = S_hd_gr,
  niter = niter_hd, burnin = burnin_hd, thin = thin_hd,
  store_beta = TRUE, store_gamma = TRUE, store_Z_list = FALSE, store_Z_pip = TRUE
)
set.seed(seed_hd[2])
fit_hd_gr2 <- bvs_pg(
  X_hd, y_hd,
  adj_type = "ggm",
  sparse = TRUE,
  S_ggm = S_hd_gr,
  niter = niter_hd, burnin = burnin_hd, thin = thin_hd,
  store_beta = TRUE, store_gamma = TRUE, store_Z_list = FALSE, store_Z_pip = TRUE
)

gelman_hd <- safe_gelman_df(
  list(fit_hd_gr1, fit_hd_gr2),
  vars = vars_gr_hd,
  autoburnin = FALSE
)
gelman_hd_summary <- gelman_method_summary(gelman_hd, "Single: sparse Bayesian GGM (PG)")

knitr::kable(
  gelman_hd_summary,
  caption = "Real-data example 2 (high-dimensional) Gelman-Rubin summary using two chains (two seeds)."
)
knitr::kable(
  data.frame(
    Method = "Single: sparse Bayesian GGM (PG)",
    Parameter = gelman_hd$Parameter,
    Point_Rhat = round(gelman_hd$Point_est, 3),
    Upper_CI_Rhat = round(gelman_hd$Upper_CI, 3),
    stringsAsFactors = FALSE
  ),
  caption = "Real-data example 2 PSRF details (Alpha and first five coefficients)."
)
```

```{r realhd-gelman-trace-2chain, eval=TRUE, fig.height=3.6, fig.width=12, fig.cap="Real-data example 2 ($p \\gg n$): two-chain trace plots for representative regression coefficients from Gelman-Rubin diagnostic runs."}
beta_mat_hd_gr1 <- get_beta_matrix(fit_hd_gr1)
if (ncol(beta_mat_hd_gr1) > 0) {
  beta_idx_hd_gr <- head(order(colMeans(abs(beta_mat_hd_gr1)), decreasing = TRUE), 4L)
} else {
  beta_idx_hd_gr <- seq_len(min(4L, ncol(X_hd)))
}
beta_idx_hd_gr <- as.integer(beta_idx_hd_gr)
par(
  mfrow = c(1, length(beta_idx_hd_gr)),
  mar = c(3, 3, 2, 1),
  mgp = c(1.8, 0.5, 0)
)
plot_two_chain_trace_panels(fit_hd_gr1, fit_hd_gr2, beta_idx_hd_gr, "PG (sparse GGM)")
```

```{r realhd-modelsize-trace, eval=TRUE, fig.height=7, fig.width=10, fig.cap="Trace plots of model size for the five high-dimensional real-data models."}
n_panels_hd <- length(fits_hd)
nrow_hd <- 2
ncol_hd <- ceiling(n_panels_hd / nrow_hd)
par(mfrow = c(nrow_hd, ncol_hd), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (mn in names(fits_hd)) {
  plot(model_size_chain(fits_hd[[mn]]),
    type = "l", col = "darkblue",
    main = mn, xlab = "Iteration", ylab = "Model size", cex.main = 0.85
  )
}
```

```{r realhd-topgene-trace, eval=TRUE, fig.height=7, fig.width=10, fig.cap="Trace plots of the top shared marker-gene coefficient in high-dimensional real data across all five methods."}
par(mfrow = c(nrow_hd, ncol_hd), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (mn in names(fits_hd)) {
  beta_chain <- get_beta_matrix(fits_hd[[mn]])[, top_gene_idx_hd]
  plot(beta_chain,
    type = "l", col = "steelblue",
    main = paste0(mn, ": ", top_gene_name_hd),
    xlab = "Iteration", ylab = "beta", cex.main = 0.85
  )
}
```

```{r realhd-topgene-runmean, eval=TRUE, fig.height=7, fig.width=10, fig.cap="Running-mean plots of the top shared marker-gene coefficient in high-dimensional real data across all five methods."}
par(mfrow = c(nrow_hd, ncol_hd), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (mn in names(fits_hd)) {
  beta_chain <- get_beta_matrix(fits_hd[[mn]])[, top_gene_idx_hd]
  rm_vals <- running_mean(beta_chain)
  plot(rm_vals,
    type = "l", col = "darkgreen",
    main = paste0(mn, ": running mean(", top_gene_name_hd, ")"),
    xlab = "Iteration", ylab = "Running mean(beta)", cex.main = 0.82
  )
}
```

```{r realhd-topgene-acf, eval=TRUE, fig.height=7, fig.width=10, fig.cap="ACF plots of the top shared marker-gene coefficient in high-dimensional real data across all five methods."}
par(mfrow = c(nrow_hd, ncol_hd), mar = c(3, 3, 2, 1), mgp = c(1.8, 0.5, 0))
for (mn in names(fits_hd)) {
  beta_chain <- get_beta_matrix(fits_hd[[mn]])[, top_gene_idx_hd]
  safe_acf_plot(beta_chain,
    lag.max = 50,
    main = paste0(mn, ": ACF(", top_gene_name_hd, ")"),
    col = "steelblue"
  )
}
```

```{r realhd-runtime-barplot, eval=TRUE, fig.height=4, fig.width=9, fig.cap="Runtime comparison (seconds) for the five high-dimensional real-data models."}
ord_rt_hd <- order(times_hd, decreasing = TRUE)
cols_hd <- c("steelblue", "coral", "seagreen3", "goldenrod3", "mediumpurple3")
barplot(times_hd[ord_rt_hd],
  col = cols_hd[ord_rt_hd],
  border = NA, las = 2, ylab = "Elapsed time (sec)",
  main = "High-dimensional real-data runtime comparison"
)
```

# Summary and discussion {#sec-summary}

The **BVS.DAdj** package provides a comprehensive toolkit for Bayesian variable
selection in logistic regression, distinguished by its ability to integrate
dual adjacency structures into the Ising/MRF prior on selection indicators.
The key features demonstrated in this article include:

1. **Two MCMC samplers**: The MH sampler provides a general-purpose approach,
   while the PG-augmented Gibbs sampler offers improved mixing through
   data augmentation that renders the logistic likelihood conditionally Gaussian.

2. **Flexible adjacency handling**: Users can supply known adjacency matrices,
   estimate them via graphical lasso, or learn them jointly through Bayesian GGM
   with SSVS priors.  The dual-network option combines both approaches.

3. **Scalability**: Sparse high-dimensional backends using CSC-format covariance
   storage and targeted MH gamma updates enable analysis of datasets with
   $p \gg n$, as demonstrated in the $p = 1000$ simulation.

4. **Computational efficiency**: All sampling backends are implemented in C++
   via **Rcpp** and **RcppArmadillo**, with the Propp-Wilson perfect simulation
   for the Möller auxiliary variable update also fully compiled.

The simulation studies show that the dual-adjacency MRF prior improves
sensitivity for detecting weakly associated variables that are neighbors in
the graph, consistent with the theoretical motivation.  The PG sampler typically
shows lower autocorrelation than the MH sampler in moderate dimensions, while
the sparse backends scale well to $p = 1000$.

The two real-data examples (Golub leukemia and Colon tumor microarray) further
illustrate practical use in both moderate and strict high-dimensional
($p \gg n$) regimes, including method-specific selected-variable summaries and
MCMC diagnostic reporting.

## Computational details

All simulations were performed on an Apple M-series ARM64 processor running
macOS with R version `r R.version.string`.  The package uses C++17 with
`-O2 -DARMA_NO_DEBUG` compiler flags.

## Software availability

The **BVS.DAdj** package (version 0.1.0) is available at
[https://github.com/yuy113/BVS.DAdj](https://github.com/yuy113/BVS.DAdj).
It requires R $\geq$ 3.5.0 with **Rcpp**, **RcppArmadillo**, and **Matrix**.
The **huge** package is optional for graphical lasso adjacency estimation.

# References {-}
